---
title: "RRHO2"
author: "Brynleigh Payne"
date: "2026-07-10"
output: html_document
---

#### Modified from:
#### https://github.com/imillercrews/TRES_heat_biomarkers_bl/blob/main/linger_tres_gene_expression_script.R#L2264  
### RRHO2 Correlation analysis of gene expression
# R 4.3.1 
 
# Clear Memory and Check working directory
```{r}

rm(list = ls(all = TRUE))
getwd()
```

#### load libraries ####
```{r}
## install packages
# BiocManager::install("DESeq2")
# BiocManager::install("vsn")
# BiocManager::install("ComplexHeatmap")
# devtools::install_github("RRHO2/RRHO2", build_opts = c("--no-resave-data", "--no-manual"))
#remotes::install_github("RRHO2/RRHO2")

## load libraries
library(tidyverse)
library("DESeq2")

# GO annotation
library(AnnotationForge)

# DEG
library(pvclust)
library(vsn)

# variance
library(variancePartition)

# interaction
library(ComplexHeatmap)
library(RRHO2)

# WGCNA
library(WGCNA)

# GO term analysis
library(clusterProfiler)
#library(org.Hrustica.eg.db)
library(enrichplot)
library(AnnotationDbi)
library(GO.db)

```


### RRHO2 - Correlation Analysis

## Blood - Run RRHO2 Correlation for males and females heat vs control response
# DESeq2 ran for each sex separately: = ~ 0 + treatment
```{r}
# Read in Blood DESeq2 DGE data
blood.res.females.heat.df <- read_csv("BloodRNAseq/mRNA_LibraryPreps/DifferentialGeneExpression/DEG_female_deseq.csv")

blood.res.males.heat.df <- read_csv("BloodRNAseq/mRNA_LibraryPreps/DifferentialGeneExpression/DEG_male_deseq.csv")

# Check intersection of gene lists
male_genes <- blood.res.males.heat.df$Gene_ID
female_genes <- blood.res.females.heat.df$Gene_ID

length(male_genes)
length(female_genes)

length(intersect(male_genes, female_genes))
length(setdiff(male_genes, female_genes))
length(setdiff(female_genes, male_genes))

setequal(blood.res.males.heat.df$Gene_ID, blood.res.females.heat.df$Gene_ID)

# Identify common genes to use for RRHO2
common <- intersect(blood.res.males.heat.df$Gene_ID, blood.res.females.heat.df$Gene_ID)

blood.res.males.heat.df <- blood.res.males.heat.df %>%
    dplyr::filter(Gene_ID %in% common)

blood.res.females.heat.df <- blood.res.females.heat.df %>%
    dplyr::filter(Gene_ID %in% common)

setequal(blood.res.males.heat.df$Gene_ID, blood.res.females.heat.df$Gene_ID)

# Run RRHO2 Correlation for males and females heat vs control response
male_rrho <- blood.res.males.heat.df |>
  dplyr::mutate(metric = sign(log2FoldChange) * -log10(pvalue)) |>
  dplyr::select(Gene_ID, metric) |>
  tidyr::drop_na() |>
  as.data.frame()

female_rrho <- blood.res.females.heat.df |>
  dplyr::mutate(metric = sign(log2FoldChange) * -log10(pvalue)) |>
  dplyr::select(Gene_ID, metric) |>
  tidyr::drop_na() |>
  as.data.frame()

str(male_rrho)

rrho_output <- RRHO2_initialize(
    male_rrho,
    female_rrho,
    labels = c("Male Heat vs Control - Blood", "Female Heat vs Control - Blood"),
    boundary = 0.1
)

# Get data frame of quadrant genes 
rrho_output.df = data.frame(Gene = rrho_output$genelist_uu$gene_list_overlap_uu,
                            Direction = 'up.up') %>% 
  full_join(data.frame(Gene = rrho_output$genelist_dd$gene_list_overlap_dd,
                       Direction = 'down.down'))%>% 
  full_join(data.frame(Gene = c(rrho_output$genelist_ud$gene_list_overlap_ud,
                                NA),
                       Direction = 'up.down'))%>% 
  full_join(data.frame(Gene = c(rrho_output$genelist_du$gene_list_overlap_du,
                                NA),
                       Direction = 'down.up')) %>% 
  na.omit()

# Plot
png('RRHO2_heat_by_sex_blood.png')
RRHO2_heatmap(rrho_output)
dev.off()

```


## Liver - Run RRHO2 Correlation for males and females heat vs control response
# DESeq2 ran for each sex separately: = ~ 0 + treatment
```{r}
# Read in Blood DESeq2 DGE data
liver.res.females.heat.df <- read_csv("LiverRNASeq/DGE/DEG_female_deseq_allgenes.csv")

liver.res.males.heat.df <- read_csv("LiverRNASeq/DGE/DEG_male_deseq_allgenes.csv")

# Check intersection of gene lists
male_genes <- liver.res.males.heat.df$Gene_ID
female_genes <- liver.res.females.heat.df$Gene_ID

length(male_genes)
length(female_genes)

length(intersect(male_genes, female_genes))
length(setdiff(male_genes, female_genes))
length(setdiff(female_genes, male_genes))

setequal(liver.res.males.heat.df$Gene_ID, liver.res.females.heat.df$Gene_ID)

# Identify common genes to use for RRHO2
common <- intersect(liver.res.males.heat.df$Gene_ID, liver.res.females.heat.df$Gene_ID)

liver.res.males.heat.df <- liver.res.males.heat.df %>%
    dplyr::filter(Gene_ID %in% common)

liver.res.females.heat.df <- liver.res.females.heat.df %>%
    dplyr::filter(Gene_ID %in% common)

setequal(liver.res.males.heat.df$Gene_ID, liver.res.females.heat.df$Gene_ID)

# Run RRHO2 Correlation for males and females heat vs control response
male_rrho <- liver.res.males.heat.df |>
  dplyr::mutate(metric = sign(log2FoldChange) * -log10(pvalue)) |>
  dplyr::select(Gene_ID, metric) |>
  tidyr::drop_na() |>
  as.data.frame()

female_rrho <- liver.res.females.heat.df |>
  dplyr::mutate(metric = sign(log2FoldChange) * -log10(pvalue)) |>
  dplyr::select(Gene_ID, metric) |>
  tidyr::drop_na() |>
  as.data.frame()

str(male_rrho)

rrho_output <- RRHO2_initialize(
    male_rrho,
    female_rrho,
    labels = c("Male Heat vs Control - Liver", "Female Heat vs Control - Liver"),
    boundary = 0.1
)

# Get data frame of quadrant genes 
rrho_output.df = data.frame(Gene = rrho_output$genelist_uu$gene_list_overlap_uu,
                            Direction = 'up.up') %>% 
  full_join(data.frame(Gene = rrho_output$genelist_dd$gene_list_overlap_dd,
                       Direction = 'down.down'))%>% 
  full_join(data.frame(Gene = c(rrho_output$genelist_ud$gene_list_overlap_ud,
                                NA),
                       Direction = 'up.down'))%>% 
  full_join(data.frame(Gene = c(rrho_output$genelist_du$gene_list_overlap_du,
                                NA),
                       Direction = 'down.up')) %>% 
  na.omit()

# Plot
png('RRHO2_heat_by_sex_liver.png')
RRHO2_heatmap(rrho_output)
dev.off()

```


## Blood vs Liver - Run RRHO2 Correlation for males heat vs control response
# DESeq2 ran for each sex separately: = ~ 0 + treatment
```{r}
# Read in male DESeq2 DGE data
blood.res.males.heat.df <- read_csv("BloodRNAseq/mRNA_LibraryPreps/DifferentialGeneExpression/DEG_male_deseq.csv")
liver.res.males.heat.df <- read_csv("LiverRNASeq/DGE/DEG_male_deseq_allgenes.csv")

# Check intersection of gene lists
blood_male_genes <- blood.res.males.heat.df$Gene_ID
liver_male_genes <- liver.res.males.heat.df$Gene_ID

length(blood_male_genes)
length(liver_male_genes)

length(intersect(blood_male_genes, liver_male_genes))
length(setdiff(blood_male_genes, liver_male_genes))
length(setdiff(liver_male_genes, blood_male_genes))

setequal(liver.res.males.heat.df$Gene_ID, blood.res.males.heat.df$Gene_ID)

# Identify common genes to use for RRHO2
common <- intersect(liver.res.males.heat.df$Gene_ID, blood.res.males.heat.df$Gene_ID)

liver.res.males.heat.df <- liver.res.males.heat.df %>%
    dplyr::filter(Gene_ID %in% common)

blood.res.males.heat.df <- blood.res.males.heat.df %>%
    dplyr::filter(Gene_ID %in% common)

setequal(liver.res.males.heat.df$Gene_ID, blood.res.males.heat.df$Gene_ID)

# Run RRHO2 Correlation for males heat vs control response
liver_male_rrho <- liver.res.males.heat.df |>
  dplyr::mutate(metric = sign(log2FoldChange) * -log10(pvalue)) |>
  dplyr::select(Gene_ID, metric) |>
  tidyr::drop_na() |>
  as.data.frame()

blood_male_rrho <- blood.res.males.heat.df |>
  dplyr::mutate(metric = sign(log2FoldChange) * -log10(pvalue)) |>
  dplyr::select(Gene_ID, metric) |>
  tidyr::drop_na() |>
  as.data.frame()

rrho_output <- RRHO2_initialize(
    liver_male_rrho,
    blood_male_rrho,
    labels = c("Male Heat vs Control - Liver", "Male Heat vs Control - Blood"),
    boundary = 0.1
)

# Get data frame of quadrant genes 
rrho_output.df = data.frame(Gene = rrho_output$genelist_uu$gene_list_overlap_uu,
                            Direction = 'up.up') %>% 
  full_join(data.frame(Gene = rrho_output$genelist_dd$gene_list_overlap_dd,
                       Direction = 'down.down'))%>% 
  full_join(data.frame(Gene = c(rrho_output$genelist_ud$gene_list_overlap_ud,
                                NA),
                       Direction = 'up.down'))%>% 
  full_join(data.frame(Gene = c(rrho_output$genelist_du$gene_list_overlap_du,
                                NA),
                       Direction = 'down.up')) %>% 
  na.omit()

# Plot
png('RRHO2_male_heat_by_tissue.png')
RRHO2_heatmap(rrho_output)
dev.off()

```


## Blood vs Liver - Run RRHO2 Correlation for females heat vs control response
# DESeq2 ran for each sex separately: = ~ 0 + treatment
```{r}
# Read in male DESeq2 DGE data
blood.res.females.heat.df <- read_csv("BloodRNAseq/mRNA_LibraryPreps/DifferentialGeneExpression/DEG_female_deseq.csv")
liver.res.females.heat.df <- read_csv("LiverRNASeq/DGE/DEG_female_deseq_allgenes.csv")

# Check intersection of gene lists
blood_female_genes <- blood.res.females.heat.df$Gene_ID
liver_female_genes <- liver.res.females.heat.df$Gene_ID

length(blood_female_genes)
length(liver_female_genes)

length(intersect(blood_female_genes, liver_female_genes))
length(setdiff(blood_female_genes, liver_female_genes))
length(setdiff(liver_female_genes, blood_female_genes))

setequal(liver.res.females.heat.df$Gene_ID, blood.res.females.heat.df$Gene_ID)

# Identify common genes to use for RRHO2
common <- intersect(liver.res.females.heat.df$Gene_ID, blood.res.females.heat.df$Gene_ID)

liver.res.females.heat.df <- liver.res.females.heat.df %>%
    dplyr::filter(Gene_ID %in% common)

blood.res.females.heat.df <- blood.res.females.heat.df %>%
    dplyr::filter(Gene_ID %in% common)

setequal(liver.res.females.heat.df$Gene_ID, blood.res.females.heat.df$Gene_ID)

# Run RRHO2 Correlation for females heat vs control response
liver_female_rrho <- liver.res.females.heat.df |>
  dplyr::mutate(metric = sign(log2FoldChange) * -log10(pvalue)) |>
  dplyr::select(Gene_ID, metric) |>
  tidyr::drop_na() |>
  as.data.frame()

blood_female_rrho <- blood.res.females.heat.df |>
  dplyr::mutate(metric = sign(log2FoldChange) * -log10(pvalue)) |>
  dplyr::select(Gene_ID, metric) |>
  tidyr::drop_na() |>
  as.data.frame()

rrho_output <- RRHO2_initialize(
    liver_female_rrho,
    blood_female_rrho,
    labels = c("Female Heat vs Control - Liver", "Female Heat vs Control - Blood"),
    boundary = 0.1
)

# Get data frame of quadrant genes 
rrho_output.df = data.frame(Gene = rrho_output$genelist_uu$gene_list_overlap_uu,
                            Direction = 'up.up') %>% 
  full_join(data.frame(Gene = rrho_output$genelist_dd$gene_list_overlap_dd,
                       Direction = 'down.down'))%>% 
  full_join(data.frame(Gene = c(rrho_output$genelist_ud$gene_list_overlap_ud,
                                NA),
                       Direction = 'up.down'))%>% 
  full_join(data.frame(Gene = c(rrho_output$genelist_du$gene_list_overlap_du,
                                NA),
                       Direction = 'down.up')) %>% 
  na.omit()

# Plot
png('RRHO2_female_heat_by_tissue.png')
RRHO2_heatmap(rrho_output)
dev.off()

```



# paper
pdf('figures/paper/DEG_linger_heat_sex_3.pdf',
    height = 6.5,
    width = 6.5)
RRHO2_heatmap(rrho_output)
dev.off()


# gene categories
rrho_output.df %>% 
  count(Direction) %>% 
  ggplot(aes(x = Direction,
             y = n,
             label = n)) +
  geom_label() +
  theme_classic() +
  ylab('Number of genes') +
  xlab('RRHO category')
ggsave('/Users/brynleighpayne/Downloads/ZebraFinch_DamageFitnessProject/BloodRNAseq/mRNA_LibraryPreps/DifferentialGeneExpression/RRHO_genes_per_category.png') 

### graph logfc across sexes
# get max and min
max = max(abs(res.males.heat.df$log2FoldChange),
          abs(res.females.heat.df$log2FoldChange))

## graph DEG for each sex
# all DEG
res.males.heat.df %>%
  dplyr::select(Gene,
                log2FoldChange) %>%
  dplyr::rename(male_fc = log2FoldChange) %>%
  inner_join(res.females.heat.df %>%
               dplyr::select(Gene, log2FoldChange) %>%
               dplyr::rename(female_fc = log2FoldChange),
             by = "Gene") %>%
  mutate(Concordance = case_when(sign(male_fc) != sign(female_fc) ~ 'Dis',
                                 TRUE ~ 'Con')) %>% 
  left_join(res.treat.df %>%
              mutate(Sig = ifelse(padj < 0.16,
                                   'Sig',
                                   'Not sig')) %>% 
              dplyr::select(Gene,
                            Sig)) %>%
  mutate(Sig = ifelse(is.na(Sig),
                      'Not sig',
                      Sig),
         label = ifelse(Sig == 'Sig',
                        Gene,
                        NA)) %>% 
  arrange(Sig) %>% 
  ggplot(aes(x = male_fc,
             y = female_fc,
             color = Sig)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = 'dashed') +
  geom_point() +
  ggrepel::geom_label_repel(aes(label = label),
                            color = 'black',
                            force = 15,
                            force_pull = 0,
                            point.padding = 0.5,
                            box.padding = 1) +
  theme_classic() +
  xlim(-max,
       max) +
  ylim(-max,
       max) +
  scale_color_manual(values = c('black',
                                'red'))+
    ggtitle('All DEG')
ggsave('figures/DEG/Interaction scatterplot DEG.png')

# each sex
```{r}
res.males.heat.df %>%
  mutate(Male.sig = ifelse(padj < 0.05,
                      'Sig',
                      'Not sig')) %>% 
  dplyr::select(Gene_ID,
                log2FoldChange,
                Male.sig) %>%
  dplyr::rename(male_fc = log2FoldChange) %>%
  inner_join(res.females.heat.df %>%
               mutate(Female.sig = ifelse(padj < 0.05,
                                   'Sig',
                                   'Not sig')) %>%
               dplyr::select(Gene_ID,
                             log2FoldChange,
                             Female.sig) %>%
               dplyr::rename(female_fc = log2FoldChange),
             by = "Gene_ID") %>%
  mutate(Concordance = case_when(sign(male_fc) != sign(female_fc) ~ 'Dis',
                                 TRUE ~ 'Con'))  %>%
  mutate(Sig = case_when(Male.sig == 'Sig' & Female.sig != 'Sig' ~ 'Male',
                         Male.sig != 'Sig' & Female.sig == 'Sig' ~ 'Female',
                         Male.sig == 'Sig' & Female.sig == 'Sig' ~ 'Both',
                         TRUE ~ 'Not sig'),
         Sig.order = case_when(Sig == 'Both' ~ 1,
                               Sig == 'Male' ~ 2,
                               Sig == 'Female' ~ 3,
                               Sig == 'Not sig' ~ 0)) %>% 
  arrange(Sig.order) %>% 
  mutate(Label = ifelse(Gene_ID %in% c('LOC120765528',
                                    'RYBP'),
                        Gene_ID,
                        NA)) %>% 
  ggplot(aes(x = male_fc,
             y = female_fc,
             color = Sig)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = 'dashed') +
  geom_point() +
  ggrepel::geom_label_repel(aes(label = Label),
             color = 'black') +
  theme_classic() +
  xlim(-max,
       max) +
  ylim(-max,
       max) + 
  scale_color_manual(values = c('purple',
                                'darkred',
                                'darkblue',
                                'grey'))+
  labs(color = 'Sig (padj<0.05)') +
  ggtitle('Sex specific DEG')
ggsave('/Users/brynleighpayne/Downloads/ZebraFinch_DamageFitnessProject/BloodRNAseq/mRNA_LibraryPreps/DifferentialGeneExpression/Interaction scatterplot sex specific DEG.png')

# paper
res.males.heat.df =read.csv('/Users/brynleighpayne/Downloads/ZebraFinch_DamageFitnessProject/BloodRNAseq/mRNA_LibraryPreps/DifferentialGeneExpression/res.males.heat.df.csv')

res.females.heat.df =read.csv('/Users/brynleighpayne/Downloads/ZebraFinch_DamageFitnessProject/BloodRNAseq/mRNA_LibraryPreps/DifferentialGeneExpression/res.females.heat.df.csv')

res.males.heat.df %>%
  mutate(Male.sig = ifelse(padj < 0.05,
                           'Sig',
                           'Not sig')) %>% 
  dplyr::select(Gene,
                log2FoldChange,
                Male.sig) %>%
  dplyr::rename(male_fc = log2FoldChange) %>%
  inner_join(res.females.heat.df %>%
               mutate(Female.sig = ifelse(padj < 0.05,
                                          'Sig',
                                          'Not sig')) %>%
               dplyr::select(Gene,
                             log2FoldChange,
                             Female.sig) %>%
               dplyr::rename(female_fc = log2FoldChange),
             by = "Gene") %>%
  mutate(Concordance = case_when(sign(male_fc) != sign(female_fc) ~ 'Dis',
                                 TRUE ~ 'Con'))  %>%
  mutate(Sig = case_when(Male.sig == 'Sig' & Female.sig != 'Sig' ~ 'Male',
                         Male.sig != 'Sig' & Female.sig == 'Sig' ~ 'Female',
                         Male.sig == 'Sig' & Female.sig == 'Sig' ~ 'Both',
                         TRUE ~ 'Not sig'),
         Sig.order = case_when(Sig == 'Both' ~ 1,
                               Sig == 'Male' ~ 2,
                               Sig == 'Female' ~ 3,
                               Sig == 'Not sig' ~ 0),
         Sig.color = case_when(female_fc > 0 & Sig == 'Both' ~ 'Heat',
                               female_fc > 0 & Sig == 'Male' ~ 'Heat',
                               female_fc > 0 & Sig == 'Female' ~ 'Heat',
                               female_fc < 0 & Sig == 'Both' ~ 'Con',
                               female_fc < 0 & Sig == 'Male' ~ 'Con',
                               female_fc < 0 & Sig == 'Femal' ~ 'Con',
                               TRUE ~ 'none'),
         Sig.shape = case_when(Sig == 'Both' ~ 'Both',
                               Sig == 'Male' | Sig == 'Female' ~ 'none',
                               TRUE ~ 'none')) %>% 
  arrange(desc(Sig.color)) %>% 
  ggplot(aes(x = male_fc,
             y = female_fc,
             color = Sig.color,
             shape = Sig.shape)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  # geom_abline(slope = 1,
  #             intercept = 0,
  #             linetype = 'dashed') +
  geom_point(size = 3) +
  # ggrepel::geom_label_repel(aes(label = Label),
  #                           color = 'black') +
  theme_classic() +
  # xlim(-max,
  #      max) +
  # ylim(-max,
  #      max) + 
  scale_color_manual(values = c('#464b9f',
                                '#f26622',
                                'grey'))+
  labs(color = 'Sig (padj<0.05)') +
  ggtitle('Sex specific DEG') +
  coord_fixed(ratio = 2) +
  theme(legend.position = 'none') +
  # theme(legend.position = 'inside',
  #       legend.position.inside = c(0.9,
  #                                  0.2)) +
  xlab('Male log2FC')+
  ylab('Female log2FC')
ggsave('/Users/brynleighpayne/Downloads/ZebraFinch_DamageFitnessProject/BloodRNAseq/mRNA_LibraryPreps/DifferentialGeneExpression/Deg_linger_heat_sex_2a.pdf',
       height = 3.25,
       width = 3.25)

# check overlap 
res.males.heat.df %>%
  mutate(Male.sig = ifelse(padj < 0.05,
                           'Sig',
                           'Not sig')) %>% 
  dplyr::select(Gene,
                log2FoldChange,
                Male.sig) %>%
  dplyr::rename(male_fc = log2FoldChange) %>%
  inner_join(res.females.heat.df %>%
               mutate(Female.sig = ifelse(padj < 0.05,
                                          'Sig',
                                          'Not sig')) %>%
               dplyr::select(Gene,
                             log2FoldChange,
                             Female.sig) %>%
               dplyr::rename(female_fc = log2FoldChange),
             by = "Gene") %>%
  mutate(Concordance = case_when(sign(male_fc) != sign(female_fc) ~ 'Dis',
                                 TRUE ~ 'Con')) %>% 
  mutate(Sig = case_when(Male.sig == 'Sig' & Female.sig != 'Sig' & male_fc > 0 ~ 'Male.up',
                         Male.sig == 'Sig' & Female.sig != 'Sig' & male_fc < 0 ~ 'Male.down',
                         Male.sig != 'Sig' & Female.sig == 'Sig' & female_fc > 0  ~ 'Female.up',
                         Male.sig != 'Sig' & Female.sig == 'Sig' & female_fc < 0  ~ 'Female.down',
                         Male.sig == 'Sig' & Female.sig == 'Sig' & male_fc > 0 ~ 'Both.up',
                         Male.sig == 'Sig' & Female.sig == 'Sig'& male_fc < 0 ~ 'Both.down',
                         TRUE ~ 'Not sig'),
         Sig.order = case_when(Sig == 'Both' ~ 1,
                               Sig == 'Male' ~ 2,
                               Sig == 'Female' ~ 3,
                               Sig == 'Not sig' ~ 0)) %>% 
  full_join(res.treat.df %>% 
              filter(padj < 0.155) %>% 
              mutate(DEG.heat = 'DEG.heat') %>% 
              dplyr::select(Gene,
                            DEG.heat)) %>% 
  count(Sig,
        DEG.heat,
        Concordance)

```


# Sig DEG.heat Concordance    n
# 1    Both.down DEG.heat         Con    1
# 2    Both.down     <NA>         Con   36
# 3      Both.up     <NA>         Con   14
# 4  Female.down     <NA>         Con    5
# 5    Female.up     <NA>         Con    2
# 6    Male.down     <NA>         Con   30
# 7      Male.up     <NA>         Con    8
# 8      Not sig DEG.heat         Con    4
# 9      Not sig DEG.heat         Dis    1
# 10     Not sig     <NA>         Con 8473
# 11     Not sig     <NA>         Dis  730


## heat boxplots both sexes
# get TOP DEG
# get overlapping DEG
res.sex.both.heat.df.deg.top =
  res.males.heat.df %>% 
  mutate(Sig.male = ifelse(padj <= 0.05,
                           'Sig',
                           'Not sig')) %>% 
  dplyr::select(Sig.male,
                Gene,
                padj) %>% 
  left_join(res.females.heat.df %>% 
              mutate(Sig.female = ifelse(padj <= 0.05,
                                         'Sig',
                                         'Not sig')) %>% 
              dplyr::select(Sig.female,
                            Gene,
                            padj),
            by = 'Gene',
            suffix = c('_male',
                       '_female')) %>% 
  mutate(Sig.both = ifelse(Sig.male == 'Sig' & Sig.female == 'Sig',
                           'Both',
                           NA)) %>% 
  filter(Sig.both == 'Both') %>% 
    dplyr::select(Gene,
                  padj_male,
                  padj_female) %>% 
    pivot_longer(cols = -c('Gene'),
                 names_to = 'Sex',
                 values_to = 'padj') %>% 
    group_by(Sex) %>% 
    slice_min(order_by = padj,
              n = 1,
              with_ties = FALSE) %>% 
    ungroup() %>% 
  pull(Gene) %>% 
  unique()
    
    
    
# normalize and scale counts
deg.sex.heat.mat = assay(vst(dds,
                    blind = F))[res.sex.both.heat.df.deg.top,] %>% 
  t() %>% 
  scale() %>% 
  as.data.frame() %>% 
  rownames_to_column('Sample') %>% 
  pivot_longer(-c(Sample),
               names_to = 'Gene',
               values_to = 'z.score') %>% 
  left_join(data.linger.sample.meta) %>% 
  left_join(data.linger.sample.sex)


## graph each DEG boxplots
# treatment temp
# paper
deg.sex.heat.mat %>% 
  mutate(ID = paste0(Treatment,
                     '.',
                     Sex)) %>% 
  ggplot(aes(x = ID,
             y = z.score,
             color = Treatment)) +
  geom_boxplot(aes(group = ID),
               outlier.shape = NA) +
  geom_point(size = 2.5,
             aes(shape = Sex)) +
  theme_classic() +
  theme(axis.text.x = element_blank()) +
  ylab('Gene expression (z-score)') +
  xlab('') +
  scale_color_manual(values = c('#464b9f',
                                '#f26622')) +
  facet_wrap(~Gene,
             scale = 'free',
             ncol = 2)  +
  theme(legend.position = 'none')
ggsave('figures/paper/DEG_linger_heat_sex_1b.pdf',
       height = 3.25,
       width = 3.25,
       dpi = 720)

