---
title: "Blood Time-Series DGE — Female A–E Timecourse (repeated measures)"
author: "Ali Amer"
date: "2026-08-07"
output: html_document
---

<!-- 
SCRIPT Info:

Differential expression across a five-point time-series (A-E).

  Where along the time-series does expression differ, and from what?
     -> Four vs-Control contrasts (B-A, C-A, D-A, E-A): does expression differ from control at any particular timepoint?
     -> Three adjacent-timepoint contrasts (C-B, D-C, E-D): what changed since
        the previous timepoint?

REPEATED MEASURES:
9 birds: assigned to the 2-week recovery group (E) - contributed two draws (same bird sampled at Heat and then recovery) The other 40 birds contributed a single draw each (49 birds, 58 samples total). Models are fit with block = ID_Band, which treats those 9 paired birds as correlated and every other bird as independent.  

Notes: 
  Occasion (from the sample name ZF_<band>_<occasion>_BL):
      B = baseline draw (2 wk pre-heat)   <- EXCLUDED here -> keep_occasions
      D = draw at the bird's own assigned endpoint (day of dissection)
      H = draw taken during heat exposure
      
  Group / Treatment (the A-E scheme used in time_map and all contrasts):
      A = Control, B = Heat, C = +2 h, D = +24 h, E = +2 wk

-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Libraries

```{r libraries}
rm(list = ls(all = TRUE))

library(edgeR)
library(limma)
library(ggplot2)
```

# CONFIG

```{r config}

CONFIG <- list(
  counts_file     = "gene_count_matrix (3).csv", # Raw Count Data
  master_file     = "AHDB_Exp1_MasterSpreadSheet.csv", # Kept to double check the pheno data file and make sure samples match
  samplelist_file = "PHENO_DATA_AllBlood.txt", # Metadata File
  out_dir         = "Results",

  drop_ribofree   = TRUE,           # exclude Ribo-free library re-preps

  # Occasion is the timepoint blood was drawn, not the treatment group:
  #   D = terminal draw at the bird's own assigned timepoint
  #   H = draw taken during heat exposure.
  keep_occasions  = c("D", "H"),

  # If a listed sample is absent from the count matrix, FALSE drops it with a warning whereas TRUE stops the code from running further.
  strict_alignment = FALSE,

  ## Duration of heat exposure
  heat_duration_h = 5,
  time_map        = NULL,      # see below

  fdr_cut         = 0.05,

  focus_contrast  = "Heat_vs_Control",

  ## Degrees of freedom for the natural spline over log1p(hours)
  spline_df       = 3,

  ## ID_Bands to label on the PCA plot. 
  flag_bands      = numeric(0),

  ## Clustering. PART returns the number of clusters; minSize is the floor on
  ## cluster size, part_recursion the number of bootstraps at each split.
  do_clustering   = TRUE,
  part_minclus    = 25,
  part_recursion  = 100,

  ## Contrasts whose significant genes seed the clustering
   cluster_seed_contrasts = c("Heat_vs_Control", "Rec2h_vs_Control",
                             "Rec24h_vs_Control", "Rec2wk_vs_Control",
                             "Rec2h_vs_Heat", "Rec24h_vs_Rec2h",
                             "Rec2wk_vs_Rec24h"),

  seed            = 1234 # Specified so seed doesn't change when running GSEA multiple times
)

## Hours since the START of heat exposure. Control set at 0.
CONFIG$time_map <- c(
  A = 0,                                # Control (never heated)
  B = CONFIG$heat_duration_h,           # end of heat exposure
  C = CONFIG$heat_duration_h + 2,       # +2 h recovery
  D = CONFIG$heat_duration_h + 24,      # +24 h recovery
  E = CONFIG$heat_duration_h + 24 * 14  # +2 wk recovery
)

## Labels for Plots 
CONFIG$treat_labels <- c(
  A = "Control",            B = "Heat",
  C = "Heat_2-HR_Repair",   D = "Heat_24-HR_Repair",
  E = "Heat_2-WK_Repair"
)

dir.create(CONFIG$out_dir, recursive = TRUE, showWarnings = FALSE)
set.seed(CONFIG$seed)
```

# Load data and build the metadata table

Three datafiles: 
1. Count matrix (expression), 
2. Pheno file (Treatment Information)
3. Master Spreadsheet (Sanity Check for myself that samples are assigned to correct treatments by cross-referencing against the Pheno File)

```{r load-metadata}

Raw_Counts <- read.csv(CONFIG$counts_file, row.names = 1,
                              check.names = FALSE)
pheno      <- read.delim(CONFIG$samplelist_file, check.names = TRUE,
                                stringsAsFactors = FALSE)
MasterSpreadsheet <- read.csv(CONFIG$master_file, check.names = TRUE)

stopifnot(all(c("sample", "treatment") %in% colnames(pheno)),
          !any(duplicated(pheno$sample)))

## Ribo-free preps are samples run through a different library kit. 
if (CONFIG$drop_ribofree) {
  is_ribofree <- grepl("_RiboFree$", pheno$sample)
  cat("RiboFree re-preps dropped:", sum(is_ribofree), "\n")
  pheno <- pheno[!is_ribofree, ]
}

## Parse band ID and draw occasion out of the sample name. (ZF_4719_D_L; Band_ID = 4719, D: Dissection Timepoint)

id_pat <- "^ZF_(\\d+)_([A-Z])_BL$"
bad <- pheno$sample[!grepl(id_pat, pheno$sample)]
if (length(bad)) {
  stop("Sample names not matching 'ZF_<band>_<occasion>_BL': ",
       paste(utils::head(bad, 10), collapse = ", "))
}

meta <- data.frame(
  RNAseq_Sample_ID = pheno$sample,
  ID_Band          = as.numeric(sub(id_pat, "\\1", pheno$sample)),
  Occasion         = sub(id_pat, "\\2", pheno$sample),
  Pheno_Treatment  = pheno$treatment
)

bird_info <- unique(MasterSpreadsheet[, c("ID_Band", "Treatment", "Sex")])
stopifnot(!any(duplicated(bird_info$ID_Band)))

missing_birds <- setdiff(meta$ID_Band, bird_info$ID_Band)
if (length(missing_birds)) {
  stop("Bands in the pheno file but not in the master: ",
       paste(missing_birds, collapse = ", "))
}

meta <- merge(meta, bird_info, by = "ID_Band", all.x = TRUE)
stopifnot(!anyNA(meta$Treatment), !anyNA(meta$Sex))

## Females only, D and H draws only. Drops the B (baseline) draws and all males.
n_before <- nrow(meta)
meta <- meta[meta$Sex == "F" & meta$Occasion %in% CONFIG$keep_occasions, ]
cat("Dropped", n_before - nrow(meta), "rows (males + B draws);",
    nrow(meta), "retained\n")

cat("\nDraw occasion x assigned treatment group (all listed samples):\n")
print(table(meta$Occasion, meta$Treatment))

```

# Cross-check the pheno treatment labels against the master

```{r treatment-crosscheck}

lab <- sub("^Baseline \\(2 weeks prior to ", "", meta$Pheno_Treatment)
lab <- sub("\\)$", "", lab)
meta$Pheno_Letter <- names(CONFIG$treat_labels)[match(lab, CONFIG$treat_labels)]
stopifnot(!anyNA(meta$Pheno_Letter))   # an unmapped label string


expected <- ifelse(meta$Occasion == "H", "B", as.character(meta$Treatment))
mismatch <- meta[meta$Pheno_Letter != expected,
                 c("RNAseq_Sample_ID", "Occasion", "Pheno_Treatment", "Treatment")]

if (nrow(mismatch)) {
  cat("Pheno/master mismatches — master used:\n"); print(mismatch)
} else {
  cat("Pheno treatment labels consistent with the master.\n")
}

```

# Merge against the count matrix

```{r check-alignment}

missing_cols <- setdiff(meta$RNAseq_Sample_ID, colnames(Raw_Counts))
extra_cols   <- setdiff(colnames(Raw_Counts), meta$RNAseq_Sample_ID)

cat("Count matrix columns:", ncol(Raw_Counts),
    "| Metadata rows:", nrow(meta), "\n")
cat("In counts but not in metadata (", length(extra_cols),
    " — includes dropped RiboFree):\n", sep = "")
print(extra_cols)

# Sanity Check:
if (length(missing_cols)) {
  cat("\nListed but NOT in the count matrix (", length(missing_cols), "):\n",
      sep = "")
  print(missing_cols)
  if (CONFIG$strict_alignment) {
    stop("Reconcile the naming convention in ", CONFIG$samplelist_file,
         " against colnames(Raw_Counts).")
  }
  ## ZF_4717 was never sent for sequencing.
  meta <- meta[!meta$RNAseq_Sample_ID %in% missing_cols, ]
  cat("Dropped (strict_alignment = FALSE).\n")
}
```

# Filter sample draws into timepoint groups

An `H` draw was taken *during heat* from a bird assigned to recovery group E, so it
belongs in the Heat group at 5 h; not in the group the bird was eventually killed
in. Group B therefore contains two kinds of sample: the `D` draws of the 10 birds
assigned to Heat, and the `H` draws of 9 E birds. Those same 9 E birds also
appear in group E via their own `D` draw.

`block = ID_Band` in the model below is what handles that. It does not assume
every bird was sampled repeatedly: a bird with a single sample is a block of one,
has no within-block pair, and is treated as independent. The consensus
correlation enters the fit globally, so it affects the standard errors of every
contrast, not only the ones spanning the B <-> E pair, I think.......

```{r pool-groups}

meta$Group <- factor(ifelse(meta$Occasion == "H", "B",
                            as.character(meta$Treatment)),
                     levels = names(CONFIG$time_map))

meta$Hours   <- CONFIG$time_map[as.character(meta$Group)]
meta$TimeVar <- log1p(meta$Hours)

rownames(meta) <- meta$RNAseq_Sample_ID
counts <- Raw_Counts[, rownames(meta), drop = FALSE]

cat("Samples per group after pooling:\n"); print(table(meta$Group))
cat("\nBirds contributing 1 / 2 / ... draws:\n"); print(table(table(meta$ID_Band)))
cat("\nSamples:", nrow(meta), "| Birds:", length(unique(meta$ID_Band)), "\n")

pat <- data.frame(
  Groups    = unname(tapply(as.character(meta$Group), meta$ID_Band,
                            function(g) paste(sort(g), collapse = ","))),
  Treatment = unname(tapply(as.character(meta$Treatment), meta$ID_Band,
                            function(t) t[1])),
  n         = unname(tapply(meta$Group, meta$ID_Band, length))
)

#Sanity Check: 
  
cat("\nAssigned treatment x observed group pattern (one row per bird):\n")
print(table(pat$Treatment, pat$Groups))

```

# Filter genes and TMM-normalise

```{r normalize}

dge <- edgeR::DGEList(counts = counts, samples = meta, group = meta$Group)
dge <- dge[edgeR::filterByExpr(dge, group = dge$samples$group), ,
           keep.lib.sizes = FALSE]
dge <- edgeR::calcNormFactors(dge, method = "TMM")
cat("Genes retained:", nrow(dge), "| Samples:", ncol(dge), "\n")

logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 1)

```

# QC: PCA: Visual Check 

```{r pca, fig.width = 7, fig.height = 5}

pca  <- stats::prcomp(t(logcpm), center = TRUE, scale. = FALSE)
n_pc <- min(5, ncol(pca$x))
pct  <- round(100 * summary(pca)$importance[2, 1:2], 1)

pca_df <- data.frame(
  PC1 = pca$x[, 1], PC2 = pca$x[, 2],
  Treatment = dge$samples$group,
  Band      = dge$samples$ID_Band,
  Occasion  = dge$samples$Occasion
)
pca_df$Flag <- ifelse(pca_df$Band %in% CONFIG$flag_bands,
                      as.character(pca_df$Band), NA_character_)

p_pca <- ggplot2::ggplot(pca_df, ggplot2::aes(PC1, PC2, colour = Treatment)) +
  ggplot2::geom_point(size = 3) +
  ggplot2::stat_ellipse(type = "norm", linewidth = 0.4, na.rm = TRUE) +
  ggplot2::geom_text(ggplot2::aes(label = Flag), colour = "black",
                     vjust = -1.1, size = 3, na.rm = TRUE) +
  ggplot2::labs(x = paste0("PC1 (", pct[1], "%)"),
                y = paste0("PC2 (", pct[2], "%)"),
              ) +
  ggplot2::theme_bw()

p_pca

ggplot2::ggsave(file.path(CONFIG$out_dir, "PCA_timecourse_with_outliers.png"),
                p_pca, width = 7, height = 5, dpi = 300)


## Shapes by occasion shows where those two draws (repeated measures) lie.
p_pca_B <- ggplot2::ggplot(subset(pca_df, Treatment == "B"),
                           ggplot2::aes(PC1, PC2, shape = Occasion)) +
  ggplot2::geom_point(size = 3) +
  ggplot2::labs(title = "Group B only, by draw occasion") +
  ggplot2::theme_bw()
p_pca_B

ggplot2::ggsave(file.path(CONFIG$out_dir, "PCA_groupB_by_occasion.png"),
                p_pca_B, width = 6, height = 5, dpi = 300)

## Euclidean distance from each sample to its own group centroid (mean) 
scores    <- pca$x[, seq_len(n_pc), drop = FALSE]
grp       <- as.character(dge$samples$group)
centroids <- t(vapply(split(seq_len(nrow(scores)), grp),
                      function(i) colMeans(scores[i, , drop = FALSE]),
                      numeric(n_pc)))

dist_tab <- data.frame(
  Sample    = rownames(scores),
  Band      = dge$samples$ID_Band,
  Occasion  = dge$samples$Occasion,
  Treatment = dge$samples$group,
  DistToCentroid = round(sqrt(rowSums((scores - centroids[grp, , drop = FALSE])^2)), 2)
)
dist_tab <- dist_tab[order(-dist_tab$DistToCentroid), ]
print(utils::head(dist_tab, 10))
utils::write.csv(dist_tab, file.path(CONFIG$out_dir, "PCA_centroid_distances.csv"),
                 row.names = FALSE)
```

# Model: group means and contrasts

`Group` is the biological condition the sample
represents, not the bird's assigned endpoint — so an E bird's `H` draw sits in
group B and its own `D` draw sits in group E.

```{r group-model}

grp_factor <- droplevels(dge$samples$group)
design_0   <- stats::model.matrix(~ 0 + grp_factor) # the addition of zero removes the intercept so we get 5 columns representing the group mean of each treatment group
colnames(design_0) <- levels(grp_factor)

# For Repeated Measures: Asks when two samples across timepoints come from the same bird, where do they fall in relation to the group mean

fit_0 <- edgeR::voomLmFit(dge, design_0, block = dge$samples$ID_Band)
cat("within-bird correlation (group design):",
    round(fit_0$correlation, 3), "\n")

## Contrasts use the A–E letters
# A (Control), B (Heat), C (2-Hr Recovery), D (24-hr recovery), E (2-Week Recovery)

con <- limma::makeContrasts(
  ## vs Control 
  Heat_vs_Control   = B - A,
  Rec2h_vs_Control  = C - A,
  Rec24h_vs_Control = D - A,
  Rec2wk_vs_Control = E - A,
  
  Rec2wk_vs_Heat    = E - B,
  
  ## Sequential Comparison
  Rec2h_vs_Heat     = C - B,
  Rec24h_vs_Rec2h   = D - C,
  Rec2wk_vs_Rec24h  = E - D,
  levels = design_0
)
fit_con <- limma::eBayes(limma::contrasts.fit(fit_0, con), robust = TRUE)

step_set <- c("Heat_vs_Control", "Rec2h_vs_Heat",
              "Rec24h_vs_Rec2h", "Rec2wk_vs_Rec24h", "Rec2wk_vs_Control")

per_con <- lapply(colnames(con), function(cn)
  limma::topTable(fit_con, coef = cn, number = Inf, sort.by = "none"))
names(per_con) <- colnames(con)

cat("\nDEGs per contrast (FDR <", CONFIG$fdr_cut, "):\n")
print(vapply(per_con, function(x) sum(x$adj.P.Val < CONFIG$fdr_cut), numeric(1)))

focus <- per_con[[CONFIG$focus_contrast]]
sig   <- focus$adj.P.Val < CONFIG$fdr_cut
cat("\nFocus contrast:", CONFIG$focus_contrast, "-", sum(sig), "genes\n")

```

# Spline model: the same seven contrasts

Fit a spline on log1p(hours) rather than as a five-level
factor design. the contrasts are built by evaluating the fitted curve at each timepoint and
differencing those fitted values — giving the same seven comparisons as above,
estimated under a smooth model instead of independent group means as the factor model


```{r spline-model}

tv       <- dge$samples$TimeVar
ns_basis <- splines::ns(tv, df = CONFIG$spline_df)

des_sp <- cbind(Intercept = 1, ns_basis)
colnames(des_sp) <- c("Intercept", paste0("ns", seq_len(ncol(ns_basis))))

fit_sp <- edgeR::voomLmFit(dge, des_sp, block = dge$samples$ID_Band)
cat("within-bird correlation (spline design):",
    round(fit_sp$correlation, 3), "\n")

## Fitted-value design row for each of the five timepoints.
tp_tv <- log1p(CONFIG$time_map)
X_tp  <- cbind(1, stats::predict(ns_basis, tp_tv))
rownames(X_tp) <- names(CONFIG$time_map)
colnames(X_tp) <- colnames(des_sp)

pairs_tp <- list(
  Heat_vs_Control   = c("B", "A"),
  Rec2h_vs_Control  = c("C", "A"),
  Rec24h_vs_Control = c("D", "A"),
  Rec2wk_vs_Control = c("E", "A"),
  Rec2h_vs_Heat     = c("C", "B"),
  Rec24h_vs_Rec2h   = c("D", "C"),
  Rec2wk_vs_Rec24h  = c("E", "D"),
  Rec2wk_vs_Heat    = c("E", "B")
)

con_sp <- vapply(pairs_tp, function(p) X_tp[p[1], ] - X_tp[p[2], ],
                 numeric(ncol(X_tp)))
rownames(con_sp) <- colnames(des_sp)

fit_sp_con <- limma::eBayes(limma::contrasts.fit(fit_sp, con_sp), robust = TRUE)

per_con_sp <- lapply(colnames(con_sp), function(cn)
  limma::topTable(fit_sp_con, coef = cn, number = Inf, sort.by = "none"))
names(per_con_sp) <- colnames(con_sp)

cat("\nSpline-model DEGs per contrast (FDR <", CONFIG$fdr_cut, "):\n")
print(vapply(per_con_sp, function(x) sum(x$adj.P.Val < CONFIG$fdr_cut), numeric(1)))

## Side-by-side with the factor model.
compare_tab <- data.frame(
  Contrast = names(per_con),
  Factor   = vapply(per_con, function(x) sum(x$adj.P.Val < CONFIG$fdr_cut), numeric(1)),
  Spline   = vapply(per_con_sp[names(per_con)],
                    function(x) sum(x$adj.P.Val < CONFIG$fdr_cut), numeric(1)),
  Overlap  = vapply(names(per_con), function(cn)
               length(intersect(
                 rownames(per_con[[cn]])[per_con[[cn]]$adj.P.Val       < CONFIG$fdr_cut],
                 rownames(per_con_sp[[cn]])[per_con_sp[[cn]]$adj.P.Val < CONFIG$fdr_cut])),
               numeric(1)),
  row.names = NULL
)
print(compare_tab)
utils::write.csv(compare_tab,
                 file.path(CONFIG$out_dir, "Spline_vs_factor_contrasts.csv"),
                 row.names = FALSE)
```

# Scatter Plot - Test (Ignore)

Genes significant in the focus contrast (`Heat_vs_Control` by default). Each
panel plots that gene's heat response on the x-axis against its state at one
recovery timepoint on the y-axis, both relative to Control. A point on the
dashed y = x line has not recovered at all


```{r scatter, fig.width = 10, fig.height = 4}

heat_lfc   <- per_con$Heat_vs_Control$logFC
panel_labs <- c(Rec2h_vs_Control  = "+2 h",
                Rec24h_vs_Control = "+24 h",
                Rec2wk_vs_Control = "+2 wk")

rev_df <- do.call(rbind, lapply(names(panel_labs), function(cn) {
  data.frame(GeneID   = rownames(focus)[sig],
             Heat_LFC = heat_lfc[sig],
             Rec_LFC  = per_con[[cn]]$logFC[sig],
             Panel    = factor(panel_labs[cn], levels = panel_labs))
}))

slope_tab <- do.call(rbind, lapply(split(rev_df, rev_df$Panel), function(d) {
  f <- stats::lm(Rec_LFC ~ 0 + Heat_LFC, data = d)
  data.frame(Panel = d$Panel[1],
             Slope = round(stats::coef(f)[1], 3),
             SE    = round(summary(f)$coefficients[1, 2], 3),
             R2    = round(summary(f)$r.squared, 3),
             row.names = NULL)
}))
print(slope_tab)

slope_tab$Label <- sprintf("slope = %.2f \u00b1 %.2f\nR\u00b2 = %.2f",
                           slope_tab$Slope, slope_tab$SE, slope_tab$R2)

p_rev <- ggplot2::ggplot(rev_df, ggplot2::aes(Heat_LFC, Rec_LFC)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey60") +
  ## Dashed y = x: where a gene would sit if nothing had recovered.
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
  ggplot2::geom_point(alpha = 0.25, size = 1, colour = "steelblue") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ 0 + x,
                       colour = "firebrick", linewidth = 0.8, se = TRUE) +
  ggplot2::geom_text(data = slope_tab,
                     ggplot2::aes(x = -Inf, y = Inf, label = Label),
                     hjust = -0.1, vjust = 1.3, size = 3) +
  ggplot2::facet_wrap(~ Panel, nrow = 1) +
  ggplot2::labs(x = "logFC Heat vs Control", y = "logFC Recovery vs Control",
                ) +
  ggplot2::theme_bw()
p_rev

ggplot2::ggsave(file.path(CONFIG$out_dir, "scatter.png"),
                p_rev, width = 10, height = 4, dpi = 300)
utils::write.csv(slope_tab[, c("Panel", "Slope", "SE", "R2")],
                 file.path(CONFIG$out_dir, "slopes.csv"), row.names = FALSE)
```

#  trajectory clustering



```{r clustering, fig.width = 9, fig.height = 6}

source("https://raw.githubusercontent.com/Ylefol/TimeSeriesAnalysis/master/R/clusterGenomics.R")
args(part)

sig_genes <- unique(unlist(lapply(CONFIG$cluster_seed_contrasts, function(cn) {
  tt <- per_con[[cn]]
  rownames(tt)[tt$adj.P.Val < CONFIG$fdr_cut]
})))
cat("Seed contrasts:", paste(CONFIG$cluster_seed_contrasts, collapse = " U "),
    "\nGenes seeding the clustering:", length(sig_genes), "\n")
stopifnot(length(sig_genes) > CONFIG$part_minclus)

grp_means <- t(apply(logcpm[sig_genes, , drop = FALSE], 1,
                     function(y) tapply(y, dge$samples$group, mean)))
grp_z <- t(scale(t(grp_means)))              # z-scoring removes amplitude
grp_z <- grp_z[stats::complete.cases(grp_z), , drop = FALSE]
cat("Genes clustered:", nrow(grp_z), "of", length(sig_genes), "\n")

part_res <- part(grp_z,
                 minSize = CONFIG$part_minclus,
                 B       = CONFIG$part_recursion,
                 dist.method = "euclidean",
                 linkage     = "average")

cl <- part_res$lab.hatK
cat("PART clusters:", part_res$hatK,
    "| unplaced genes:", length(part_res$outliers), "\n")
print(table(cl))

keep  <- cl != 0
grp_z <- grp_z[keep, , drop = FALSE]
cl    <- cl[keep]

clust_df <- do.call(rbind, lapply(sort(unique(cl)), function(k) {
  m    <- grp_z[cl == k, , drop = FALSE]
  sd_k <- if (nrow(m) > 1) apply(m, 2, stats::sd) else rep(0, ncol(m))
  data.frame(Treatment = factor(CONFIG$treat_labels[colnames(grp_z)],
                                levels = CONFIG$treat_labels),
             Mean = colMeans(m), SD = sd_k,
             Cluster = paste0("Cluster ", k, " (n=", nrow(m), ")"))
}))

p_clust <- ggplot2::ggplot(clust_df, ggplot2::aes(Treatment, Mean, group = 1)) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = Mean - SD, ymax = Mean + SD),
                       fill = "steelblue", alpha = 0.25) +
  ggplot2::geom_line(colour = "steelblue", linewidth = 0.9) +
  ggplot2::geom_point(colour = "steelblue", size = 2) +
  ggplot2::facet_wrap(~ Cluster) +
  ggplot2::labs(x = NULL, y = "z-scored mean logCPM",
                ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
p_clust

ggplot2::ggsave(file.path(CONFIG$out_dir, "Trajectory_clusters.png"),
                p_clust, width = 9, height = 6, dpi = 300)
utils::write.csv(data.frame(GeneID = rownames(grp_z), Cluster = cl),
                 file.path(CONFIG$out_dir, "PART_clusters.csv"),
                 row.names = FALSE)
```

# Outputs

`Timeseries_contrast_results.csv` — one row per tested gene, with logFC, raw p
and FDR for all seven contrasts side by side. This is the table downstream
scripts should read.

`Sig_<contrast>.csv` — the significant genes for each contrast that has any,
already sorted by FDR.

```{r outputs}

## All seven contrasts, wide. per_con tables are all sort.by = "none", so rows
## are aligned across contrasts and can be cbind-ed directly.
gene_ids <- rownames(per_con[[1]])
stopifnot(all(vapply(per_con, function(x) identical(rownames(x), gene_ids),
                     logical(1))))

all_tab <- do.call(cbind, lapply(names(per_con), function(cn)
  stats::setNames(per_con[[cn]][, c("logFC", "P.Value", "adj.P.Val")],
                  paste0(c("logFC_", "PValue_", "FDR_"), cn))))

out <- data.frame(
  GeneID = gene_ids,
  Symbol = sub("^gene-([^|]+)\\|.*$", "\\1", gene_ids),
  logCPM = per_con[[1]]$AveExpr,
  all_tab,
  row.names = NULL, check.names = FALSE
)
out <- out[order(out[[paste0("FDR_", CONFIG$focus_contrast)]]), ]

w <- function(x, f) utils::write.csv(x, file.path(CONFIG$out_dir, f),
                                     row.names = FALSE)

w(out, "Timeseries_contrast_results.csv")

## Same table for the spline model.
sp_tab <- do.call(cbind, lapply(names(per_con_sp), function(cn)
  stats::setNames(per_con_sp[[cn]][, c("logFC", "P.Value", "adj.P.Val")],
                  paste0(c("logFC_", "PValue_", "FDR_"), cn))))

out_sp <- data.frame(
  GeneID = rownames(per_con_sp[[1]]),
  Symbol = sub("^gene-([^|]+)\\|.*$", "\\1", rownames(per_con_sp[[1]])),
  logCPM = per_con_sp[[1]]$AveExpr,
  sp_tab,
  row.names = NULL, check.names = FALSE
)
w(out_sp, "Timeseries_contrast_results_spline.csv")

## Per-contrast significant gene lists
for (cn in names(per_con)) {
  tt  <- per_con[[cn]]
  hit <- tt$adj.P.Val < CONFIG$fdr_cut
  if (!any(hit)) next
  sig_tab <- data.frame(
    GeneID = gene_ids[hit],
    Symbol = sub("^gene-([^|]+)\\|.*$", "\\1", gene_ids[hit]),
    logCPM = tt$AveExpr[hit],
    logFC  = tt$logFC[hit],
    PValue = tt$P.Value[hit],
    FDR    = tt$adj.P.Val[hit],
    row.names = NULL, check.names = FALSE
  )
  sig_tab <- sig_tab[order(sig_tab$FDR), ]
  w(sig_tab, paste0("Sig_", cn, ".csv"))
}

```

