---
title: "Blood Time-Series DGE — Female A–E Timecourse (repeated measures)"
author: "Ali Amer"
date: "2026-08-07"
output: html_document
---

<!-- 
SCRIPT Info:

Differential expression across a five-point time-series, where the same
individual can contribute more than one sample. Three questions, in order:

  1.  does this gene change at all across the time-series?
  
This was done in two different ways: factor ANODEV and a spline
                model 
  2. where along the time-series does expression change?
  
This was done through adjacent-timepoint contrasts (B-A, C-B, D-C, E-D)
                plus E-A as a "did it return to baseline?" test.
                
Repeated Measures: Blood was drawn more than once for recovey birds, so residuals are not independent. 
Each model was fit with "(..., block = ID_Band) - to account for the repeated measures.

-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Libraries

```{r libraries}
rm(list = ls(all = TRUE))

library(edgeR)
library(limma)
library(splines)
library(ggplot2)
```

# CONFIG

```{r config}
CONFIG <- list(
  counts_file     = "gene_count_matrix (3).csv", # Raw Count Data
  master_file     = "AHDB_Exp1_MasterSpreadSheet.csv", # Kept to double check the pheno data file and make sure samples match
  samplelist_file = "PHENO_DATA_AllBlood.txt", # Metadata File
  out_dir         = "Results",

  drop_ribofree   = TRUE,           # exclude Ribo-free library re-preps

  ## Sample names are ZF_<band>_<occasion>_BL. Occasion is the timepoint blood was DRAWN, not the
  ## treatment group:
  ##   D = terminal draw at the bird's own assigned timepoint
  ##   H = draw taken during heat exposure, from birds assigned to recover
  keep_occasions  = c("D", "H"),

  ## If a listed sample is absent from the count matrix, FALSE drops it with a warning whereas TRUE stops the code from running further.
  strict_alignment = FALSE,

  ## Duration of heat exposure
  heat_duration_h = 5,
  time_map        = NULL,      # see below

  ## Spline Model
  spline_df       = 4,         #  DF -> 5 timepoints - 1 = 4
  fdr_cut         = 0.05,      

  ## For Cluster Analysis - Ignore for now (Need to think more on how to choose the appropriate number of clusters. Each cluster represents genes with similar expression patterns)
  
  do_clustering   = FALSE,
  n_clusters      = 6,

  seed            = 1234 # Specified so seed doesn't change when running GSEA multiple times
)

## Hours since the START of heat exposure. Control set at 0.
CONFIG$time_map <- c(
  A = 0,                                # Control (never heated)
  B = CONFIG$heat_duration_h,           # end of heat exposure
  C = CONFIG$heat_duration_h + 2,       # +2 h recovery
  D = CONFIG$heat_duration_h + 24,      # +24 h recovery
  E = CONFIG$heat_duration_h + 24 * 14  # +2 wk recovery
)

## Labels for Plots 
CONFIG$treat_labels <- c(
  A = "Control",            B = "Heat",
  C = "Heat_2-HR_Repair",   D = "Heat_24-HR_Repair",
  E = "Heat_2-WK_Repair"
)

dir.create(CONFIG$out_dir, recursive = TRUE, showWarnings = FALSE)
set.seed(CONFIG$seed)
```

# Load data and build the metadata table

Three datafiles: 
1. Count matrix (expression), 
2. Pheno file (Treatment Information)
3. Master Spreadsheet (Sanity Check for myself that samples are assigned to correct treatments by cross-referencing against the Pheno File)

```{r load-metadata}

Raw_Counts <- utils::read.csv(CONFIG$counts_file, row.names = 1,
                              check.names = FALSE)
pheno      <- read.delim(CONFIG$samplelist_file, check.names = TRUE,
                                stringsAsFactors = FALSE)
MasterSpreadsheet <- read.csv(CONFIG$master_file, check.names = TRUE)

stopifnot(all(c("sample", "treatment") %in% colnames(pheno)),
          !any(duplicated(pheno$sample)))

## Ribo-free preps are samples run through a different library kit. 
if (CONFIG$drop_ribofree) {
  is_ribofree <- grepl("_RiboFree$", pheno$sample)
  cat("RiboFree re-preps dropped:", sum(is_ribofree), "\n")
  pheno <- pheno[!is_ribofree, ]
}

## Parse band ID and draw occasion out of the sample name. (ZF_4719_D_L; Band_ID = 4719, D: Dissection Timepoint)

id_pat <- "^ZF_(\\d+)_([A-Z])_BL$"
bad <- pheno$sample[!grepl(id_pat, pheno$sample)]
if (length(bad)) {
  stop("Sample names not matching 'ZF_<band>_<occasion>_BL': ",
       paste(utils::head(bad, 10), collapse = ", "))
}

meta <- data.frame(
  RNAseq_Sample_ID = pheno$sample,
  ID_Band          = as.numeric(sub(id_pat, "\\1", pheno$sample)),
  Occasion         = sub(id_pat, "\\2", pheno$sample),
  Pheno_Treatment  = pheno$treatment
)

bird_info <- unique(MasterSpreadsheet[, c("ID_Band", "Treatment", "Sex")])
stopifnot(!any(duplicated(bird_info$ID_Band)))

missing_birds <- setdiff(meta$ID_Band, bird_info$ID_Band)
if (length(missing_birds)) {
  stop("Bands in the pheno file but not in the master: ",
       paste(missing_birds, collapse = ", "))
}

meta <- merge(meta, bird_info, by = "ID_Band", all.x = TRUE)
stopifnot(!anyNA(meta$Treatment), !anyNA(meta$Sex))

## Females only, D and H draws only. Drops the B (baseline) draws and all males.
n_before <- nrow(meta)
meta <- meta[meta$Sex == "F" & meta$Occasion %in% CONFIG$keep_occasions, ]
cat("Dropped", n_before - nrow(meta), "rows (males + B draws);",
    nrow(meta), "retained\n")

cat("\nDraw occasion x assigned treatment group (all listed samples):\n")
print(table(meta$Occasion, meta$Treatment))

cat("\nDraw occasion x sex:\n")
print(table(meta$Occasion, meta$Sex, meta$Pheno_Treatment))
```

# Cross-check the pheno treatment labels against the master

```{r treatment-crosscheck}

lab <- sub("^Baseline \\(2 weeks prior to ", "", meta$Pheno_Treatment)
lab <- sub("\\)$", "", lab)
meta$Pheno_Letter <- names(CONFIG$treat_labels)[match(lab, CONFIG$treat_labels)]
stopifnot(!anyNA(meta$Pheno_Letter))   # an unmapped label string

expected <- ifelse(meta$Occasion == "H", "B", as.character(meta$Treatment))
mismatch <- meta[meta$Pheno_Letter != expected,
                 c("RNAseq_Sample_ID", "Occasion", "Pheno_Treatment", "Treatment")]

if (nrow(mismatch)) {
  cat("Pheno/master mismatches — master used:\n"); print(mismatch)
} else {
  cat("Pheno treatment labels consistent with the master.\n")
}

## limma and edgeR name the same quantities differently. This renames
## topTable output to edgeR's column names so the downstream chunks are
## identical to the liver script.

```

# Merge against the count matrix

```{r check-alignment}

missing_cols <- setdiff(meta$RNAseq_Sample_ID, colnames(Raw_Counts))
extra_cols   <- setdiff(colnames(Raw_Counts), meta$RNAseq_Sample_ID)

cat("Count matrix columns:", ncol(Raw_Counts),
    "| Metadata rows:", nrow(meta), "\n")
cat("In counts but not in metadata (", length(extra_cols),
    " — includes dropped RiboFree):\n", sep = "")
print(extra_cols)

# Sanity Check:
if (length(missing_cols)) {
  cat("\nListed but NOT in the count matrix (", length(missing_cols), "):\n",
      sep = "")
  print(missing_cols)
  if (CONFIG$strict_alignment) {
    stop("Reconcile the naming convention in ", CONFIG$samplelist_file,
         " against colnames(Raw_Counts).")
  }
  ## ZF_4717 was never sent for sequencing.
  meta <- meta[!meta$RNAseq_Sample_ID %in% missing_cols, ]
  cat("Dropped (strict_alignment = FALSE).\n")
}
```

# Filter sample draws into timepoint groups

An `H` draw was taken *during heat* from a bird assigned to
recover later, so it belongs in the Heat group at 5 h; not in the group the
bird was eventually killed in. Group B therefore contains two kinds of sample:
the `D` draws of birds assigned to Heat, and the `H` draws of C/D/E birds.

Note: A recovery bird contributes to group B
*and* to its own group, so contrasts like C−B are part within-bird and part
between-bird. `duplicateCorrelation` would handle this *I think..... Double check*

```{r filter-pool}

# Sanity Check: 
stopifnot(all(CONFIG$keep_occasions %in% meta$Occasion))
if (!is.null(CONFIG$outlier_bands)) {
  stopifnot(all(CONFIG$outlier_bands %in% meta$ID_Band))
}

keep <- meta$Treatment %in% names(CONFIG$time_map)
if (!is.null(CONFIG$outlier_bands)) keep <- keep & !meta$ID_Band %in% CONFIG$outlier_bands
meta <- meta[keep, ]

stopifnot(all(meta$Treatment[meta$Occasion == "H"] %in% c("C", "D", "E")))

meta$Group <- factor(ifelse(meta$Occasion == "H", "B",
                            as.character(meta$Treatment)),
                     levels = names(CONFIG$time_map))

meta$Hours   <- CONFIG$time_map[as.character(meta$Group)]
meta$TimeVar <- log1p(meta$Hours)

rownames(meta) <- meta$RNAseq_Sample_ID
counts <- Raw_Counts[, rownames(meta), drop = FALSE]

cat("Samples per group after pooling:\n"); print(table(meta$Group))
cat("\nBirds contributing 1 / 2 / ... draws:\n"); print(table(table(meta$ID_Band)))
cat("\nSamples:", nrow(meta), "| Birds:", length(unique(meta$ID_Band)), "\n")

```

# Filter genes and TMM-normalise

```{r normalize}

dge <- edgeR::DGEList(counts = counts, samples = meta, group = meta$Group)
dge <- dge[edgeR::filterByExpr(dge, group = dge$samples$group), ,
           keep.lib.sizes = FALSE]
dge <- edgeR::calcNormFactors(dge, method = "TMM")
cat("Genes retained:", nrow(dge), "| Samples:", ncol(dge), "\n")

logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 2)

```

# QC: PCA: Visual Check for any outlier samples

```{r pca, fig.width = 7, fig.height = 5}

pca  <- stats::prcomp(t(logcpm), center = TRUE, scale. = FALSE)
n_pc <- min(5, ncol(pca$x))
pct  <- round(100 * summary(pca)$importance[2, 1:2], 1)

pca_df <- data.frame(
  PC1 = pca$x[, 1], PC2 = pca$x[, 2],
  Treatment = dge$samples$group,
  Band      = dge$samples$ID_Band
)
pca_df$Flag <- ifelse(pca_df$Band %in% CONFIG$flag_bands,
                      as.character(pca_df$Band), NA_character_)

p_pca <- ggplot2::ggplot(pca_df, ggplot2::aes(PC1, PC2, colour = Treatment)) +
  ggplot2::geom_point(size = 3) +
  ggplot2::stat_ellipse(type = "norm", linewidth = 0.4, na.rm = TRUE) +
  ggplot2::geom_text(ggplot2::aes(label = Flag), colour = "black",
                     vjust = -1.1, size = 3, na.rm = TRUE) +
  ggplot2::labs(x = paste0("PC1 (", pct[1], "%)"),
                y = paste0("PC2 (", pct[2], "%)"),
                title = "Blood, female timecourse (candidate outliers labelled)") +
  ggplot2::theme_bw()

p_pca

ggplot2::ggsave(file.path(CONFIG$out_dir, "PCA_timecourse_with_outliers.png"),
                p_pca, width = 7, height = 5, dpi = 300)

## Euclidean distance from each sample to its own group centroid (mean) over PC1..n_pc
scores    <- pca$x[, seq_len(n_pc), drop = FALSE]
grp       <- as.character(dge$samples$group)
centroids <- t(vapply(split(seq_len(nrow(scores)), grp),
                      function(i) colMeans(scores[i, , drop = FALSE]),
                      numeric(n_pc)))

dist_tab <- data.frame(
  Sample    = rownames(scores),
  Band      = dge$samples$ID_Band,
  Occasion  = dge$samples$Occasion,
  Treatment = dge$samples$group,
  DistToCentroid = round(sqrt(rowSums((scores - centroids[grp, , drop = FALSE])^2)), 2)
)
dist_tab <- dist_tab[order(-dist_tab$DistToCentroid), ]
print(utils::head(dist_tab, 10))
utils::write.csv(dist_tab, file.path(CONFIG$out_dir, "PCA_centroid_distances.csv"),
                 row.names = FALSE)
```

# Model 1: Spline 

```{r spline-model}

spl <- splines::ns(dge$samples$TimeVar, df = CONFIG$spline_df)
design_spline <- stats::model.matrix(~ spl)
colnames(design_spline) <- c("Intercept",
                             paste0("spl", seq_len(CONFIG$spline_df)))

fit_spline <- edgeR::voomLmFit(dge, design_spline, block = dge$samples$ID_Band)
cat("Consensus within-bird correlation (spline design):",
    round(fit_spline$correlation, 3), "\n")
fit_spline <- limma::eBayes(fit_spline, robust = TRUE)

res_spline <- tidy_tt(limma::topTable(fit_spline,
                                      coef = 2:(CONFIG$spline_df + 1),
                                      number = Inf, sort.by = "none"))
cat("Spline-significant genes (FDR <", CONFIG$fdr_cut, "):",
    sum(res_spline$FDR < CONFIG$fdr_cut), "\n")

```

# Model 2: group means, ANODEV and all contrasts

```{r group-model}

grp_factor <- droplevels(dge$samples$group)
design_0   <- stats::model.matrix(~ 0 + grp_factor) # the addition of zero removes the intercept so we get 5 columns representing the group mean of each treatment group
colnames(design_0) <- levels(grp_factor)

# For Repeated Measures: Asks when two samples across timepoints come from the same bird, where do they fall in relation to the group mean

fit_0 <- edgeR::voomLmFit(dge, design_0, block = dge$samples$ID_Band)
cat("Consensus within-bird correlation (group design):",
    round(fit_0$correlation, 3), "\n")

## Contrasts use the A–E letters
# A (Control), B (Heat), C (2-Hr Recovery), D (24-hr recovery), E (2-Week Recovery)

con <- limma::makeContrasts(
  ## vs Control 
  Heat_vs_Control   = B - A,
  Rec2h_vs_Control  = C - A,
  Rec24h_vs_Control = D - A,
  Rec2wk_vs_Control = E - A,
  
  ## Sequential Comparison
  Rec2h_vs_Heat     = C - B,
  Rec24h_vs_Rec2h   = D - C,
  Rec2wk_vs_Rec24h  = E - D,
  levels = design_0
)
fit_con <- limma::eBayes(limma::contrasts.fit(fit_0, con), robust = TRUE)

anodev_set <- c("Heat_vs_Control", "Rec2h_vs_Control",
                "Rec24h_vs_Control", "Rec2wk_vs_Control")
step_set   <- c("Heat_vs_Control", "Rec2h_vs_Heat",
                "Rec24h_vs_Rec2h", "Rec2wk_vs_Rec24h", "Rec2wk_vs_Control")

res_grp <- tidy_tt(limma::topTable(fit_con, coef = anodev_set,
                                   number = Inf, sort.by = "none"))
cat("Factor ANODEV significant genes:", sum(res_grp$FDR < CONFIG$fdr_cut), "\n")

## sort.by = "none" keeps every table in the row order of dge.
per_con <- lapply(colnames(con), function(cn)
  tidy_tt(limma::topTable(fit_con, coef = cn, number = Inf, sort.by = "none")))
names(per_con) <- colnames(con)

cat("\nDEGs per contrast (FDR <", CONFIG$fdr_cut, "):\n")
print(vapply(per_con, function(x) sum(x$FDR < CONFIG$fdr_cut), numeric(1)))
```

## Spline vs Factor Model

```{r overlap}
print(table(spline = res_spline$FDR < CONFIG$fdr_cut,
            factor = res_grp$FDR   < CONFIG$fdr_cut))
```


```{r reversal-scatter, fig.width = 10, fig.height = 4}

sig        <- res_spline$FDR < CONFIG$fdr_cut   
heat_lfc   <- per_con$Heat_vs_Control$logFC
panel_labs <- c(Rec2h_vs_Control  = "+2 h",
                Rec24h_vs_Control = "+24 h",
                Rec2wk_vs_Control = "+2 wk")

rev_df <- do.call(rbind, lapply(names(panel_labs), function(cn) {
  data.frame(GeneID   = rownames(res_spline)[sig],
             Heat_LFC = heat_lfc[sig],
             Rec_LFC  = per_con[[cn]]$logFC[sig],
             Panel    = factor(panel_labs[cn], levels = panel_labs))
}))

slope_tab <- do.call(rbind, lapply(split(rev_df, rev_df$Panel), function(d) {
  f <- stats::lm(Rec_LFC ~ 0 + Heat_LFC, data = d)
  data.frame(Panel = d$Panel[1],
             Slope = round(stats::coef(f)[1], 3),
             SE    = round(summary(f)$coefficients[1, 2], 3),
             R2    = round(summary(f)$r.squared, 3),
             row.names = NULL)
}))
print(slope_tab)

slope_tab$Label <- sprintf("slope = %.2f \u00b1 %.2f\nR\u00b2 = %.2f",
                           slope_tab$Slope, slope_tab$SE, slope_tab$R2)

p_rev <- ggplot2::ggplot(rev_df, ggplot2::aes(Heat_LFC, Rec_LFC)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey60") +
  ## Dashed y = x: where a gene would sit if nothing had recovered..... come back to this
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
  ggplot2::geom_point(alpha = 0.25, size = 1, colour = "steelblue") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ 0 + x,
                       colour = "firebrick", linewidth = 0.8, se = TRUE) +
  ggplot2::geom_text(data = slope_tab,
                     ggplot2::aes(x = -Inf, y = Inf, label = Label),
                     hjust = -0.1, vjust = 1.3, size = 3) +
  ggplot2::facet_wrap(~ Panel, nrow = 1) +
  ggplot2::labs(x = "logFC Heat vs Control", y = "logFC Recovery vs Control") +
  ggplot2::theme_bw()
p_rev

ggplot2::ggsave(file.path(CONFIG$out_dir, "Reversal_scatter.png"),
                p_rev, width = 10, height = 4, dpi = 300)
utils::write.csv(slope_tab[, c("Panel", "Slope", "SE", "R2")],
                 file.path(CONFIG$out_dir, "Reversal_slopes.csv"), row.names = FALSE)
```

#  trajectory clustering


```{r clustering, eval = CONFIG$do_clustering, fig.width = 9, fig.height = 6}
sig_genes <- rownames(res_spline)[res_spline$FDR < CONFIG$fdr_cut]
stopifnot(length(sig_genes) > CONFIG$n_clusters)

grp_means <- t(apply(logcpm[sig_genes, , drop = FALSE], 1,
                     function(y) tapply(y, dge$samples$group, mean)))
grp_z <- t(scale(t(grp_means)))              # z-scoring removes amplitude
grp_z <- grp_z[stats::complete.cases(grp_z), , drop = FALSE]
cat("Genes clustered:", nrow(grp_z), "of", length(sig_genes), "significant\n")

km <- stats::kmeans(grp_z, centers = CONFIG$n_clusters, nstart = 50, iter.max = 100)

clust_df <- do.call(rbind, lapply(seq_len(CONFIG$n_clusters), function(k) {
  m  <- grp_z[km$cluster == k, , drop = FALSE]
  se <- if (nrow(m) > 1) apply(m, 2, function(v) stats::sd(v) / sqrt(length(v)))
        else rep(0, ncol(m))
  data.frame(Treatment = factor(CONFIG$treat_labels[colnames(grp_z)],
                                levels = CONFIG$treat_labels),
             Mean = colMeans(m), SE = se,
             Cluster = paste0("Cluster ", k, " (n=", nrow(m), ")"))
}))

p_clust <- ggplot2::ggplot(clust_df, ggplot2::aes(Treatment, Mean, group = 1)) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = Mean - SE, ymax = Mean + SE),
                       fill = "steelblue", alpha = 0.25) +
  ggplot2::geom_line(colour = "steelblue", linewidth = 0.9) +
  ggplot2::geom_point(colour = "steelblue", size = 2) +
  ggplot2::facet_wrap(~ Cluster, scales = "free_y") +
  ggplot2::labs(x = NULL, y = "z-scored mean logCPM") +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
p_clust

ggplot2::ggsave(file.path(CONFIG$out_dir, "Trajectory_clusters.png"),
                p_clust, width = 9, height = 6, dpi = 300)
```

# Outputs

One row per tested gene: both model FDRs, then logFC and FDR for each
adjacent contrast. 
```{r outputs}
step_tab <- do.call(cbind, lapply(step_set, function(cn)
  stats::setNames(per_con[[cn]][, c("logFC", "FDR")],
                  paste0(c("logFC_", "FDR_"), cn))))

out <- data.frame(
  GeneID        = rownames(res_spline),
  Symbol        = sub("^gene-([^|]+)\\|.*$", "\\1", rownames(res_spline)),
  logCPM        = res_spline$logCPM,
  PValue_spline = res_spline$PValue,
  FDR_spline    = res_spline$FDR,
  FDR_factor    = res_grp$FDR,
  step_tab,
  row.names = NULL, check.names = FALSE
)

if (CONFIG$do_clustering) {
  out$Cluster <- unname(km$cluster[out$GeneID])   # NA for non-significant genes
}
out <- out[order(out$FDR_factor), ]

utils::write.csv(out, file.path(CONFIG$out_dir, "Timeseries_spline_results.csv"),
                 row.names = FALSE)

if (CONFIG$do_clustering) {
  grp_means_out <- as.data.frame(grp_means)
  colnames(grp_means_out) <- CONFIG$treat_labels[colnames(grp_means_out)]
  utils::write.csv(grp_means_out,
                   file.path(CONFIG$out_dir, "Sig_gene_group_means_logCPM.csv"))

  for (k in seq_len(CONFIG$n_clusters)) {
    utils::write.csv(out[which(out$Cluster == k), c("GeneID", "Symbol")],
                     file.path(CONFIG$out_dir, paste0("Cluster_", k, "_genes.csv")),
                     row.names = FALSE)
  }
}
```

```{r session}
sessionInfo()
```
