---
title: "Blood Time-Series DGE — Female A–E Timecourse (repeated measures)"
author: "Ali Amer"
date: "2026-08-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


```{r libraries}
rm(list = ls(all = TRUE))
getwd()

library(edgeR)     
library(limma)
library(splines)
library(ggplot2)
```

# CONFIG

```{r config}
CONFIG <- list(
  counts_file     = "gene_count_matrix (3).csv",
  master_file     = "AHDB_Exp1_MasterSpreadSheet.csv",
  samplelist_file = "PHENO_DATA_AllBlood.txt",
  out_dir         = "Results",

  sex_levels      = "F",
  outlier_bands   = numeric(0),     # none removed yet
  drop_ribofree   = TRUE,           # exclude _RiboFree library re-preps

  ## Which draw occasions to keep. B = Baseline (2 wk pre-heat),
  ## D = terminal draw at the bird's timepoint, H = draw at heat exposure
  ## (recovery birds).
  keep_occasions  = c("B", "D", "H"),

  ## Hours since the START of heat exposure. Control anchored at 0.
  heat_duration_h = 5,
  time_map        = NULL,           # filled below

  spline_df       = 3,              # 3 df on 5 timepoints (4 df available)
  fdr_cut         = 0.05,
  n_clusters      = 6,
  seed            = 1234
)

CONFIG$time_map <- c(
  A = 0,
  B = CONFIG$heat_duration_h,
  C = CONFIG$heat_duration_h + 2,
  D = CONFIG$heat_duration_h + 24,
  E = CONFIG$heat_duration_h + 24 * 14
)

## Labels for plots and output only — NOT used in makeContrasts (hyphens would
## be parsed as subtraction). These match the pheno-file treatment strings.
CONFIG$treat_labels <- c(
  A = "Control", B = "Heat", C = "Heat_2-HR_Repair",
  D = "Heat_24-HR_Repair", E = "Heat_2-WK_Repair"
)

dir.create(CONFIG$out_dir, recursive = TRUE, showWarnings = FALSE)
set.seed(CONFIG$seed)
```

# Load data and build metadata

Occasion-to-group rules (group at the moment of the draw):

* **B (Baseline)** — pre-exposure, pooled into group **A** at time 0.
  Caveat: drawn ~2 wk before the Control terminal draws it is pooled with;
  assumes no handling/batch difference between draw occasions.
* **D** — the bird's own Treatment group (A–E) from the master spreadsheet.
* **H** — taken at the heat exposure of recovery birds (2-HR, 24-HR, or
  2-WK repair), pooled into group **B** at `heat_duration_h`.

```{r load-metadata}
Raw_Counts        <- utils::read.csv(CONFIG$counts_file, row.names = 1,
                                     check.names = FALSE)
MasterSpreadsheet <- utils::read.csv(CONFIG$master_file, check.names = TRUE)
pheno             <- utils::read.delim(CONFIG$samplelist_file,
                                       check.names = TRUE,
                                       stringsAsFactors = FALSE)
stopifnot(all(c("sample", "treatment") %in% colnames(pheno)),
          !any(duplicated(pheno$sample)))

## Drop the RiboFree library re-preps (different chemistry = batch effect)
is_ribofree <- grepl("_RiboFree$", pheno$sample)
cat("RiboFree re-preps found:", sum(is_ribofree), "\n")
if (CONFIG$drop_ribofree) pheno <- pheno[!is_ribofree, ]

## Parse band and draw occasion from the sample name (ZF_4531_B_BL)
id_pat <- "^ZF_(\\d+)_([A-Z])_BL$"
bad <- pheno$sample[!grepl(id_pat, pheno$sample)]
if (length(bad)) {
  stop("Sample names not matching 'ZF_<band>_<occasion>_BL': ",
       paste(utils::head(bad, 10), collapse = ", "))
}
pheno$ID_Band  <- as.numeric(sub(id_pat, "\\1", pheno$sample))
pheno$Occasion <- sub(id_pat, "\\2", pheno$sample)
stopifnot(!anyNA(pheno$ID_Band),
          all(pheno$Occasion %in% c("B", "D", "H")),
          !any(duplicated(pheno[, c("ID_Band", "Occasion")])))

meta <- data.frame(
  RNAseq_Sample_ID = pheno$sample,
  ID_Band          = pheno$ID_Band,
  Occasion         = pheno$Occasion,
  Pheno_Treatment  = pheno$treatment
)

## One bird-level Treatment/Sex row per band from the master (authoritative)
bird_info <- unique(MasterSpreadsheet[, c("ID_Band", "Treatment", "Sex")])
stopifnot(!any(duplicated(bird_info$ID_Band)))

missing_birds <- setdiff(meta$ID_Band, bird_info$ID_Band)
if (length(missing_birds)) {
  stop("Bands in the pheno file but not in the master spreadsheet: ",
       paste(missing_birds, collapse = ", "))
}

meta <- merge(meta, bird_info, by = "ID_Band", all.x = TRUE)
stopifnot(!anyNA(meta$Treatment), !anyNA(meta$Sex))

cat("Occasion x Treatment (all listed blood samples):\n")
print(table(meta$Occasion, meta$Treatment, useNA = "ifany"))
cat("\nOccasion x Sex:\n")
print(table(meta$Occasion, meta$Sex, useNA = "ifany"))
```

# Cross-check pheno treatment labels against the master

The pheno `treatment` column encodes the state at the draw. Convert it to a
letter and compare with what the occasion rules + master imply. Mismatches
are printed and the master wins.

```{r treatment-crosscheck}
lab <- sub("^Baseline \\(2 weeks prior to ", "", meta$Pheno_Treatment)
lab <- sub("\\)$", "", lab)
meta$Pheno_Letter <- names(CONFIG$treat_labels)[match(lab, CONFIG$treat_labels)]
stopifnot(!anyNA(meta$Pheno_Letter))   # unmapped label string

## Expected letter per row: B/D rows should name the bird's Treatment
## (for B it is the parenthetical "prior to <group>"); H rows should say Heat.
expected <- ifelse(meta$Occasion == "H", "B", as.character(meta$Treatment))

mismatch <- meta[meta$Pheno_Letter != expected,
                 c("RNAseq_Sample_ID", "Occasion", "Pheno_Treatment",
                   "Treatment")]
if (nrow(mismatch)) {
  cat("Pheno/master treatment mismatches (master used):\n")
  print(mismatch)
} else {
  cat("Pheno treatment labels consistent with the master.\n")
}
```

# Check count-matrix alignment

```{r check-alignment}
missing_cols <- setdiff(meta$RNAseq_Sample_ID, colnames(Raw_Counts))
extra_cols   <- setdiff(colnames(Raw_Counts), meta$RNAseq_Sample_ID)

cat("Count matrix columns:", ncol(Raw_Counts),
    "| Metadata rows:", nrow(meta), "\n")
cat("\nIn metadata but NOT in counts (", length(missing_cols), "):\n", sep = "")
print(missing_cols)
cat("In counts but not in metadata (", length(extra_cols),
    ", includes dropped RiboFree):\n", sep = "")
print(extra_cols)

if (length(missing_cols)) {
  stop("RNAseq_Sample_IDs absent from the count matrix (", length(missing_cols),
       "). Reconcile the naming convention in ", CONFIG$samplelist_file,
       " against colnames(Raw_Counts).")
}
```

# Filter samples and assign groups

```{r filter-samples}
stopifnot(all(CONFIG$sex_levels %in% unique(meta$Sex)),
          all(CONFIG$keep_occasions %in% unique(meta$Occasion)))
if (length(CONFIG$outlier_bands)) {
  stopifnot(all(CONFIG$outlier_bands %in% meta$ID_Band))
}

meta <- meta[meta$Sex %in% CONFIG$sex_levels &
             meta$Treatment %in% names(CONFIG$time_map) &
             meta$Occasion %in% CONFIG$keep_occasions &
             !(meta$ID_Band %in% CONFIG$outlier_bands), ]

## H draws come from recovery birds (2-HR, 24-HR, or 2-WK repair)
stopifnot(all(meta$Treatment[meta$Occasion == "H"] %in% c("C", "D", "E")))

meta$Group <- ifelse(meta$Occasion == "B", "A",
              ifelse(meta$Occasion == "H", "B",
                     as.character(meta$Treatment)))
meta$Group <- factor(meta$Group, levels = names(CONFIG$time_map))

rownames(meta) <- meta$RNAseq_Sample_ID
counts <- Raw_Counts[, rownames(meta), drop = FALSE]
stopifnot(identical(colnames(counts), rownames(meta)))

meta$Hours   <- CONFIG$time_map[as.character(meta$Group)]
meta$TimeVar <- log1p(meta$Hours)   # log-spaced: hours to weeks in one design

cat("Samples per group (after occasion pooling):\n")
print(table(meta$Group))
cat("Draws per bird:\n")
print(table(table(meta$ID_Band)))
cat("Samples retained:", nrow(meta),
    "| Birds:", length(unique(meta$ID_Band)), "\n")

stopifnot(all(table(meta$Group) >= CONFIG$min_n_per_group))
if (is.finite(CONFIG$expected_n)) stopifnot(nrow(meta) == CONFIG$expected_n)
## Repeated measures only exist if some bird contributes > 1 sample
stopifnot(any(table(meta$ID_Band) > 1))
```

# Filter and TMM normalize

```{r normalize}

stopifnot(!any(c("group", "lib.size", "norm.factors") %in% colnames(meta)))

dge <- edgeR::DGEList(counts = counts, samples = meta, group = meta$Group)
keep_genes <- edgeR::filterByExpr(dge, group = dge$samples$group)
dge <- dge[keep_genes, , keep.lib.sizes = FALSE]
dge <- edgeR::calcNormFactors(dge, method = "TMM")
cat("Genes retained:", nrow(dge), "| Samples:", ncol(dge), "\n")

logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 2)

## Standardize limma topTable output to the liver script's column names
tidy_tt <- function(tt) {
  out <- data.frame(row.names = rownames(tt))
  if ("logFC" %in% colnames(tt)) out$logFC <- tt$logFC
  out$logCPM <- tt$AveExpr
  out$PValue <- tt$P.Value
  out$FDR    <- tt$adj.P.Val
  out
}
```

# QC: PCA with candidate outliers labelled

```{r pca, fig.width = 7, fig.height = 5}
pca <- stats::prcomp(t(logcpm), center = TRUE, scale. = FALSE)
n_pc <- min(5, ncol(pca$x))
pct  <- round(100 * summary(pca)$importance[2, 1:2], 1)

pca_df <- data.frame(
  PC1       = pca$x[, 1],
  PC2       = pca$x[, 2],
  Treatment = dge$samples$group,
  Band      = dge$samples$ID_Band
)
pca_df$Flag <- ifelse(pca_df$Band %in% c(4531, 4817),
                      as.character(pca_df$Band), NA_character_)

p_pca <- ggplot2::ggplot(pca_df, ggplot2::aes(PC1, PC2, colour = Treatment)) +
  ggplot2::geom_point(size = 3) +
  ggplot2::stat_ellipse(type = "norm", linewidth = 0.4, na.rm = TRUE) +
  ggplot2::geom_text(ggplot2::aes(label = Flag), colour = "black",
                     vjust = -1.1, size = 3, na.rm = TRUE) +
  ggplot2::labs(x = paste0("PC1 (", pct[1], "%)"),
                y = paste0("PC2 (", pct[2], "%)"),
                title = "Blood, female timecourse (candidate outliers labelled)") +
  ggplot2::theme_bw()
p_pca

ggplot2::ggsave(file.path(CONFIG$out_dir, "PCA_timecourse_with_outliers.png"),
                p_pca, width = 7, height = 5, dpi = 300)
```

```{r centroid-distance}
## Distance from each sample to its own treatment-group centroid across the
## first n_pc components — a numeric companion to the visual check.
scores <- pca$x[, seq_len(n_pc), drop = FALSE]
grp    <- as.character(dge$samples$group)

centroids <- t(vapply(split(seq_len(nrow(scores)), grp),
                      function(idx) colMeans(scores[idx, , drop = FALSE]),
                      numeric(n_pc)))

d_cent <- sqrt(rowSums((scores - centroids[grp, , drop = FALSE])^2))

dist_tab <- data.frame(
  Sample    = rownames(scores),
  Band      = dge$samples$ID_Band,
  Occasion  = dge$samples$Occasion,
  Treatment = dge$samples$group,
  DistToCentroid = round(d_cent, 2)
)
dist_tab <- dist_tab[order(-dist_tab$DistToCentroid), ]
print(utils::head(dist_tab, 10))

utils::write.csv(dist_tab,
                 file.path(CONFIG$out_dir, "PCA_centroid_distances.csv"),
                 row.names = FALSE)
```

# Spline time-series model (bird as random ID)

Omnibus test per gene: does expression change smoothly with time in any way?
`voomLmFit(block = ID_Band)` runs voom, estimates the consensus intra-bird
correlation with `duplicateCorrelation`, then refits with that correlation —
the repeated-measures equivalent of a random intercept per bird.

```{r spline-model}
spl <- splines::ns(dge$samples$TimeVar, df = CONFIG$spline_df)
design_spline <- stats::model.matrix(~ spl)
colnames(design_spline) <- c("Intercept",
                             paste0("spl", seq_len(CONFIG$spline_df)))
stopifnot(qr(design_spline)$rank == ncol(design_spline))

fit_spline <- edgeR::voomLmFit(dge, design_spline,
                               block = dge$samples$ID_Band)
if (!is.null(fit_spline$correlation)) {
  cat("Consensus intra-bird correlation (spline design):",
      round(fit_spline$correlation, 3), "\n")
}
fit_spline <- limma::eBayes(fit_spline, robust = TRUE)

res_spline <- tidy_tt(limma::topTable(fit_spline,
                                      coef = 2:(CONFIG$spline_df + 1),
                                      number = Inf, sort.by = "none"))
cat("Spline-significant genes (FDR <", CONFIG$fdr_cut, "):",
    sum(res_spline$FDR < CONFIG$fdr_cut), "\n")
```

# Cross-check: factor-based ANODEV

```{r factor-anodev}
grp_factor <- dge$samples$group
design_grp <- stats::model.matrix(~ grp_factor)
colnames(design_grp) <- gsub("^grp_factor", "", colnames(design_grp))

fit_grp <- edgeR::voomLmFit(dge, design_grp, block = dge$samples$ID_Band)
fit_grp <- limma::eBayes(fit_grp, robust = TRUE)
res_grp <- tidy_tt(limma::topTable(fit_grp, coef = 2:ncol(design_grp),
                                   number = Inf, sort.by = "none"))

stopifnot(setequal(rownames(res_grp), rownames(res_spline)))
overlap <- table(
  spline = res_spline$FDR < CONFIG$fdr_cut,
  factor = res_grp[rownames(res_spline), "FDR"] < CONFIG$fdr_cut
)
print(overlap)
```

# Adjacent-timepoint contrasts

Locates where along the trajectory the change occurs. Level names stay as A–E
here so `makeContrasts` can parse them. The blocked fit (`fit_0`) is shared by
this chunk and the reversal scatter.

```{r contrasts}
design_0 <- stats::model.matrix(~ 0 + grp_factor)
colnames(design_0) <- levels(grp_factor)

fit_0 <- edgeR::voomLmFit(dge, design_0, block = dge$samples$ID_Band)
if (!is.null(fit_0$correlation)) {
  cat("Consensus intra-bird correlation (group design):",
      round(fit_0$correlation, 3), "\n")
}

con <- limma::makeContrasts(
  Heat_vs_Control   = B - A,
  Rec2h_vs_Heat     = C - B,
  Rec24h_vs_Rec2h   = D - C,
  Rec2wk_vs_Rec24h  = E - D,
  Rec2wk_vs_Control = E - A,
  levels = design_0
)

fit_con <- limma::eBayes(limma::contrasts.fit(fit_0, con), robust = TRUE)

step_res <- lapply(colnames(con), function(cn) {
  tt <- tidy_tt(limma::topTable(fit_con, coef = cn,
                                number = Inf, sort.by = "none"))
  data.frame(logFC = tt$logFC, FDR = tt$FDR, row.names = rownames(tt))
})
names(step_res) <- colnames(con)

step_tab <- do.call(cbind, lapply(names(step_res), function(cn) {
  x <- step_res[[cn]][rownames(res_spline), , drop = FALSE]
  stats::setNames(x, paste0(c("logFC_", "FDR_"), cn))
}))
stopifnot(nrow(step_tab) == nrow(res_spline))

print(vapply(step_res, function(x) sum(x$FDR < CONFIG$fdr_cut), numeric(1)))
```

# Reversal scatter: recovery logFC vs heat logFC

```{r reversal-scatter, fig.width = 10, fig.height = 4}

## Recovery-vs-Control contrasts (B - A already exists in step_res)
con_rev <- limma::makeContrasts(
  Rec2h  = C - A,
  Rec24h = D - A,
  Rec2wk = E - A,
  levels = design_0
)

fit_rev <- limma::eBayes(limma::contrasts.fit(fit_0, con_rev), robust = TRUE)

rev_lfc <- vapply(colnames(con_rev), function(cn) {
  tt <- limma::topTable(fit_rev, coef = cn, number = Inf, sort.by = "none")
  tt[rownames(res_spline), "logFC"]
}, numeric(nrow(res_spline)))

sig      <- res_spline$FDR < CONFIG$fdr_cut
heat_lfc <- step_res$Heat_vs_Control[rownames(res_spline), "logFC"]

panel_labs <- c(Rec2h = "+2 h", Rec24h = "+24 h", Rec2wk = "+2 wk")

rev_df <- do.call(rbind, lapply(colnames(con_rev), function(cn) {
  data.frame(
    GeneID   = rownames(res_spline)[sig],
    Heat_LFC = heat_lfc[sig],
    Rec_LFC  = rev_lfc[sig, cn],
    Panel    = factor(panel_labs[cn], levels = panel_labs)
  )
}))

## Slope through the origin: fraction of heat response retained
slope_tab <- do.call(rbind, lapply(split(rev_df, rev_df$Panel), function(d) {
  f <- stats::lm(Rec_LFC ~ 0 + Heat_LFC, data = d)
  data.frame(
    Panel = d$Panel[1],
    Slope = round(stats::coef(f)[1], 3),
    SE    = round(summary(f)$coefficients[1, 2], 3),
    R2    = round(summary(f)$r.squared, 3),
    row.names = NULL
  )
}))
print(slope_tab)

slope_tab$Label <- sprintf("slope = %.2f \u00b1 %.2f\nR\u00b2 = %.2f",
                           slope_tab$Slope, slope_tab$SE, slope_tab$R2)

p_rev <- ggplot2::ggplot(rev_df, ggplot2::aes(Heat_LFC, Rec_LFC)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey60") +
  ggplot2::geom_abline(slope = 1, intercept = 0,
                       linetype = 2, colour = "grey60") +
  ggplot2::geom_point(alpha = 0.25, size = 1, colour = "steelblue") +
  ggplot2::geom_smooth(method = "lm", formula = y ~ 0 + x,
                       colour = "firebrick", linewidth = 0.8, se = TRUE) +
  ggplot2::geom_text(data = slope_tab,
                     ggplot2::aes(x = -Inf, y = Inf, label = Label),
                     hjust = -0.1, vjust = 1.3, size = 3) +
  ggplot2::facet_wrap(~ Panel, nrow = 1) +
  ggplot2::labs(
    x = "logFC Heat vs Control",
    y = "logFC Recovery vs Control",
  ) +
  ggplot2::theme_bw()
p_rev

ggplot2::ggsave(file.path(CONFIG$out_dir, "Reversal_scatter.png"),
                p_rev, width = 10, height = 4, dpi = 300)

utils::write.csv(slope_tab[, c("Panel", "Slope", "SE", "R2")],
                 file.path(CONFIG$out_dir, "Reversal_slopes.csv"),
                 row.names = FALSE)
```

# Trajectory clustering

Groups significant genes by response *shape* rather than direction, so
"spiked and recovered" and "spiked and stayed up" do not share a gene list.

```{r clustering, fig.width = 9, fig.height = 6}
sig_genes <- rownames(res_spline)[res_spline$FDR < CONFIG$fdr_cut]
stopifnot(length(sig_genes) > CONFIG$n_clusters)

grp_means <- t(apply(logcpm[sig_genes, , drop = FALSE], 1, function(y)
  tapply(y, dge$samples$group, mean)))
grp_z <- t(scale(t(grp_means)))
grp_z <- grp_z[stats::complete.cases(grp_z), , drop = FALSE]
cat("Genes clustered:", nrow(grp_z), "of", length(sig_genes), "significant\n")

km <- stats::kmeans(grp_z, centers = CONFIG$n_clusters,
                    nstart = 50, iter.max = 100)

clust_df <- do.call(rbind, lapply(seq_len(CONFIG$n_clusters), function(k) {
  m  <- grp_z[km$cluster == k, , drop = FALSE]
  se <- if (nrow(m) > 1) {
    apply(m, 2, function(v) stats::sd(v) / sqrt(length(v)))
  } else {
    rep(0, ncol(m))          # single-gene cluster: no spread to show
  }
  data.frame(
    Treatment = factor(CONFIG$treat_labels[colnames(grp_z)],
                       levels = CONFIG$treat_labels),
    Mean      = colMeans(m),
    SE        = se,
    Cluster   = paste0("Cluster ", k, " (n=", nrow(m), ")")
  )
}))

p_clust <- ggplot2::ggplot(clust_df, ggplot2::aes(Treatment, Mean, group = 1)) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = Mean - SE, ymax = Mean + SE),
                       fill = "steelblue", alpha = 0.25) +
  ggplot2::geom_line(colour = "steelblue", linewidth = 0.9) +
  ggplot2::geom_point(colour = "steelblue", size = 2) +
  ggplot2::facet_wrap(~ Cluster, scales = "free_y") +
  ggplot2::labs(x = NULL, y = "z-scored mean logCPM") +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
p_clust

ggplot2::ggsave(file.path(CONFIG$out_dir, "Trajectory_clusters.png"),
                p_clust, width = 9, height = 6, dpi = 300)
```

# Outputs

```{r outputs}
out <- data.frame(
  GeneID        = rownames(res_spline),
  Symbol        = sub("^gene-([^|]+)\\|.*$", "\\1", rownames(res_spline)),
  logCPM        = res_spline$logCPM,
  PValue_spline = res_spline$PValue,
  FDR_spline    = res_spline$FDR,
  FDR_factor    = res_grp[rownames(res_spline), "FDR"],
  step_tab,
  Cluster       = unname(km$cluster[rownames(res_spline)]),
  row.names = NULL, check.names = FALSE
)
out <- out[order(out$FDR_spline), ]

utils::write.csv(out,
                 file.path(CONFIG$out_dir, "Timeseries_spline_results.csv"),
                 row.names = FALSE)

grp_means_out <- as.data.frame(grp_means)
colnames(grp_means_out) <- CONFIG$treat_labels[colnames(grp_means_out)]
utils::write.csv(grp_means_out,
                 file.path(CONFIG$out_dir, "Sig_gene_group_means_logCPM.csv"))

## Per-cluster gene lists for downstream GSEA/GO (feed into Script 7)
for (k in seq_len(CONFIG$n_clusters)) {
  utils::write.csv(out[which(out$Cluster == k), c("GeneID", "Symbol")],
                   file.path(CONFIG$out_dir, paste0("Cluster_", k, "_genes.csv")),
                   row.names = FALSE)
}
```

```{r session}
sessionInfo()
```
