---
title: "PathwayEnrichment"
author: "Ali Amer"
date: "2026-07-02"
output: html_document
---

# Script Goal:
# 1. Load exported DESeq2 DEG results (male & female) from CSV
# 2. Clean Gene_ID -> Symbol and map to Entrez ID via biomaRt
# 3. Build ranked gene lists (log2FoldChange) for GSEA
# 4. Run gseKEGG (organism = "tgu") for Male and Female separately
# 5. Export KEGG GSEA results to CSV and produce dotplots (Up in Heat / Down in Heat)
# 6  Run WGCNA with and without CORT 
# 7  Run Gene Ontology

# Clear Memory and Check working directory

```{r}

rm(list = ls(all = TRUE))
getwd()

```

# Load packages

```{r}

library(dplyr)
library(biomaRt)
library(clusterProfiler)
library(ggplot2)

```

# Load Data, Define Output Paths & Create Directories

```{r}

CONFIG <- list(
  # Input CSVs exported from the DESeq2 chunk of 6_DGE_DESeq2_LimmaVoom_EdgeR_Comparison
  male_csv    = "DEG_male_deseq_allgenes.csv",
  female_csv  = "DEG_female_deseq_allgenes.csv",

  # Ensembl BioMart
  biomart_dataset = "tguttata_gene_ensembl",
  biomart_version = 115,

  # KEGG
  kegg_organism = "tgu",
  pvalue_cutoff = 0.05,

  # Output paths
  out_dir_male   = "Results/GSEA/DESeq2_Male",
  out_dir_female = "Results/GSEA/DESeq2_Female"
)

dir.create(CONFIG$out_dir_male,   recursive = TRUE, showWarnings = FALSE)
dir.create(CONFIG$out_dir_female, recursive = TRUE, showWarnings = FALSE)
```

# Load exported DESeq2 DEG CSVs

```{r}
DEG_male_deseq   <- read.csv(CONFIG$male_csv,   stringsAsFactors = FALSE)
DEG_female_deseq <- read.csv(CONFIG$female_csv, stringsAsFactors = FALSE)

cat("Male DEG rows loaded:",   nrow(DEG_male_deseq),   "\n")
cat("Female DEG rows loaded:", nrow(DEG_female_deseq), "\n\n")

stopifnot("log2FoldChange" %in% colnames(DEG_male_deseq))
stopifnot("log2FoldChange" %in% colnames(DEG_female_deseq))
stopifnot("Gene_ID" %in% colnames(DEG_male_deseq))
stopifnot("Gene_ID" %in% colnames(DEG_female_deseq))
```

# Clean Gene IDs -> Symbol

# Gene_ID format inherited from the raw count matrix rownames, e.g. "gene-NR3C1|NR3C1"

```{r}
DEG_male_deseq$Symbol   <- sub("\\|.*", "", sub("^gene-", "", DEG_male_deseq$Gene_ID))
DEG_female_deseq$Symbol <- sub("\\|.*", "", sub("^gene-", "", DEG_female_deseq$Gene_ID))

cat("Sample male symbols:\n")
print(head(DEG_male_deseq$Symbol, 5))
cat("\nSample female symbols:\n")
print(head(DEG_female_deseq$Symbol, 5))
```

# BioMart: map Symbol -> Entrez 

```{r}
mart_finch <- useEnsembl(
  biomart = "ensembl",
  dataset = CONFIG$biomart_dataset,
  version = CONFIG$biomart_version
)

all_symbols <- unique(c(DEG_male_deseq$Symbol, DEG_female_deseq$Symbol))
cat("Unique symbols to map (male + female union):", length(all_symbols), "\n\n")

map_ZF <- getBM(
  attributes = c("external_gene_name", "entrezgene_id"),
  filters    = "external_gene_name",
  values     = all_symbols,
  mart       = mart_finch
)
```

# Map Entrez onto each sex's DEG table

```{r}

DEG_male_deseq$Entrez <- map_ZF$entrezgene_id[match(DEG_male_deseq$Symbol, map_ZF$external_gene_name)]
DEG_female_deseq$Entrez <- map_ZF$entrezgene_id[match(DEG_female_deseq$Symbol, map_ZF$external_gene_name)]

cat("Male mapping results:\n")
cat("  Total DEGs:", nrow(DEG_male_deseq), "\n")
cat("  With Entrez IDs:", sum(!is.na(DEG_male_deseq$Entrez)), "\n")
cat("  Success rate:", round(sum(!is.na(DEG_male_deseq$Entrez)) / nrow(DEG_male_deseq) * 100, 1), "%\n\n")

cat("Female mapping results:\n")
cat("  Total DEGs:", nrow(DEG_female_deseq), "\n")
cat("  With Entrez IDs:", sum(!is.na(DEG_female_deseq$Entrez)), "\n")
cat("  Success rate:", round(sum(!is.na(DEG_female_deseq$Entrez)) / nrow(DEG_female_deseq) * 100, 1), "%\n\n")

```

### GSEA: Males ###

```{r}

cat(" GSEA: Males \n")

gene_list_male <- DEG_male_deseq %>%
  filter(!is.na(Entrez)) %>%
  arrange(desc(log2FoldChange)) %>%
  { setNames(.$log2FoldChange, as.character(.$Entrez)) } %>%
  .[!duplicated(names(.))]

gene_list_male <- sort(gene_list_male, decreasing = TRUE)
cat("  Ranked genes:", length(gene_list_male), "\n\n")

gseKEGG_male <- gseKEGG(
  gene_list_male,
  organism     = CONFIG$kegg_organism,
  pvalueCutoff = CONFIG$pvalue_cutoff,
  verbose      = FALSE
)

if (!is.null(gseKEGG_male) && nrow(gseKEGG_male@result) > 0) {
  cat("  \u2713 Found", nrow(gseKEGG_male@result), "pathways\n\n")

  write.csv(
    gseKEGG_male@result,
    file.path(CONFIG$out_dir_male, "GSE_KEGG_Male.csv"),
    row.names = FALSE
  )

  p_male <- dotplot(
    gseKEGG_male,
    showCategory = 20,
    split = ".sign",
    title = "Male DESeq2 GSEA KEGG"
  ) +
    facet_grid(
      . ~ .sign,
      labeller = labeller(.sign = c(activated = "Up in Heat", suppressed = "Down in Heat"))
    )

  ggsave(
    file.path(CONFIG$out_dir_male, "GSE_KEGG_Male.png"),
    p_male, width = 14, height = 10, dpi = 300
  )
  print(p_male)
} else {
  cat("  No significant pathways found for Males.\n\n")
}
```

### GSEA: Females ###

```{r}

cat("GSEA: Females\n")

gene_list_female <- DEG_female_deseq %>%
  filter(!is.na(Entrez)) %>%
  arrange(desc(log2FoldChange)) %>%
  { setNames(.$log2FoldChange, as.character(.$Entrez)) } %>%
  .[!duplicated(names(.))]

gene_list_female <- sort(gene_list_female, decreasing = TRUE)
cat("  Ranked genes:", length(gene_list_female), "\n\n")

gseKEGG_female <- gseKEGG(
  gene_list_female,
  organism     = CONFIG$kegg_organism,
  pvalueCutoff = CONFIG$pvalue_cutoff,
  verbose      = FALSE
)

if (!is.null(gseKEGG_female) && nrow(gseKEGG_female@result) > 0) {
  cat("  \u2713 Found", nrow(gseKEGG_female@result), "pathways\n\n")

  write.csv(
    gseKEGG_female@result,
    file.path(CONFIG$out_dir_female, "GSE_KEGG_Female.csv"),
    row.names = FALSE
  )

  p_female <- dotplot(
    gseKEGG_female,
    showCategory = 20,
    split = ".sign",
    title = "Female DESeq2 GSEA KEGG"
  ) +
    facet_grid(
      . ~ .sign,
      labeller = labeller(.sign = c(activated = "Up in Heat", suppressed = "Down in Heat"))
    )

  ggsave(
    file.path(CONFIG$out_dir_female, "GSE_KEGG_Female.png"),
    p_female, width = 14, height = 10, dpi = 300
  )
  print(p_female)
} else {
  cat("  No significant pathways found for Females.\n\n")
}
```
