---
title: "PathwayEnrichment"
author: "Ali Amer"
date: "2026-07-02"
output: html_document
---

# Script Goal:
# 1. Load exported DESeq2 DEG results (all-genes, male & female) from CSV
# 2. Clean Gene_ID -> Symbol and map to Entrez ID via biomaRt
# 3. Build ranked gene lists (sign(log2FoldChange) * -log10(pvalue)) for GSEA
# 4. Run gseKEGG (organism = "tgu") for Male and Female separately
# 5. Export KEGG GSEA results to CSV and produce dotplots (Up in Heat / Down in Heat)
# 6. Run GO for Male and Females (Enrich_GO for Over-Representation and GO-GSEA)

# Clear Memory and Check working directory

```{r}

rm(list = ls(all = TRUE))
getwd()
```

# Load packages

```{r}
library(dplyr)
library(biomaRt)
library(clusterProfiler)
library(AnnotationHub)
library(ggplot2)
library(patchwork)
library(WGCNA)
library(DESeq2)
```

# CONFIG

```{r}

CONFIG <- list(
  # Input CSVs exported from the DESeq2 chunk of 6_DGE_DESeq2_LimmaVoom_EdgeR_Comparison
  male_csv    = "DEG_male_deseq_allgenes.csv",
  female_csv  = "DEG_female_deseq_allgenes.csv",

  # Ensembl BioMart
  biomart_dataset = "tguttata_gene_ensembl",
  biomart_version = 115,

  # KEGG
  kegg_organism = "tgu",
  pvalue_cutoff = 0.05,

  # GO — via AnnotationHub OrgDb (org.Taeniopygia_guttata)
  go_ontologies = c("BP", "MF", "CC"),
  go_key_type   = "SYMBOL",
  go_min_gs     = 10,
  go_max_gs     = 500,
  go_show_cat   = 10,   

  # WGCNA 
  wgcna_raw_counts_csv    = "ZF_Liver_rawgene_count_matrix.csv",
  wgcna_pheno_txt         = "PHENO_DATA_Liver_Control_Heat.txt",
  wgcna_master_csv        = "AHDB_Exp1_MasterSpreadSheet.csv",
  wgcna_cort_col          = "CORT.ng.mL.",
  wgcna_include_cort      = TRUE,
  wgcna_min_count         = 10,
  wgcna_min_samples       = 5,
  wgcna_power_test_range  = c(1:20),
  wgcna_network_type      = "signed",
  wgcna_min_module_size   = 30,
  wgcna_merge_cut_height  = 0.25,


  # Output paths
  out_dir_male   = "Results/GSEA/DESeq2_Male",
  out_dir_female = "Results/GSEA/DESeq2_Female",
  out_dir_wgcna_male   = "Results/WGCNA/Male",
  out_dir_wgcna_female = "Results/WGCNA/Female"
)

dir.create(CONFIG$out_dir_male,         recursive = TRUE, showWarnings = FALSE)
dir.create(CONFIG$out_dir_female,       recursive = TRUE, showWarnings = FALSE)
dir.create(CONFIG$out_dir_wgcna_male,   recursive = TRUE, showWarnings = FALSE)
dir.create(CONFIG$out_dir_wgcna_female, recursive = TRUE, showWarnings = FALSE)
```

# Load exported DESeq2 DEG CSVs

```{r}
DEG_male_deseq   <- read.csv(CONFIG$male_csv,   stringsAsFactors = FALSE)
DEG_female_deseq <- read.csv(CONFIG$female_csv, stringsAsFactors = FALSE)

cat("Male DEG rows loaded:",   nrow(DEG_male_deseq),   "\n")
cat("Female DEG rows loaded:", nrow(DEG_female_deseq), "\n\n")

stopifnot("log2FoldChange" %in% colnames(DEG_male_deseq))
stopifnot("log2FoldChange" %in% colnames(DEG_female_deseq))
```

# Clean Gene IDs -> Symbol

# Gene_ID format inherited from the raw count matrix rownames, e.g. "gene-NR3C1|NR3C1"

```{r}
DEG_male_deseq$Symbol   <- sub("\\|.*", "", sub("^gene-", "", DEG_male_deseq$Gene_ID))
DEG_female_deseq$Symbol <- sub("\\|.*", "", sub("^gene-", "", DEG_female_deseq$Gene_ID))

cat("Sample male symbols:\n")
print(head(DEG_male_deseq$Symbol, 5))
cat("\nSample female symbols:\n")
print(head(DEG_female_deseq$Symbol, 5))
```

# BioMart: map Symbol -> Entrez

```{r}
 
mart_finch <- useEnsembl(
  biomart = "ensembl",
  dataset = CONFIG$biomart_dataset,
  version = CONFIG$biomart_version
)

all_symbols <- unique(c(DEG_male_deseq$Symbol, DEG_female_deseq$Symbol))
cat("Unique symbols to map (male + female together):", length(all_symbols), "\n\n")

map_ZF <- getBM(
  attributes = c("external_gene_name", "entrezgene_id"),
  filters    = "external_gene_name",
  values     = all_symbols,
  mart       = mart_finch
)
```

# Map Entrez onto each sex's DEG table

```{r}

DEG_male_deseq$Entrez <- map_ZF$entrezgene_id[match(DEG_male_deseq$Symbol, map_ZF$external_gene_name)]
DEG_female_deseq$Entrez <- map_ZF$entrezgene_id[match(DEG_female_deseq$Symbol, map_ZF$external_gene_name)]

cat("Male mapping results:\n")
cat("  Total DEGs:", nrow(DEG_male_deseq), "\n")
cat("  With Entrez IDs:", sum(!is.na(DEG_male_deseq$Entrez)), "\n")
cat("  Success rate:", round(sum(!is.na(DEG_male_deseq$Entrez)) / nrow(DEG_male_deseq) * 100, 1), "%\n\n")

cat("Female mapping results:\n")
cat("  Total DEGs:", nrow(DEG_female_deseq), "\n")
cat("  With Entrez IDs:", sum(!is.na(DEG_female_deseq$Entrez)), "\n")
cat("  Success rate:", round(sum(!is.na(DEG_female_deseq$Entrez)) / nrow(DEG_female_deseq) * 100, 1), "%\n\n")
```

# Build ranking metric: sign(log2FoldChange) * -log10(pvalue)

```{r}
DEG_male_deseq$Rank   <- sign(DEG_male_deseq$log2FoldChange)   * -log10(DEG_male_deseq$pvalue)
DEG_female_deseq$Rank <- sign(DEG_female_deseq$log2FoldChange) * -log10(DEG_female_deseq$pvalue)

cat("Sample male ranks:\n")
print(head(DEG_male_deseq[, c("Symbol", "log2FoldChange", "pvalue", "Rank")]))
cat("\nSample female ranks:\n")
print(head(DEG_female_deseq[, c("Symbol", "log2FoldChange", "pvalue", "Rank")]))

```


### GSEA: Males ###

```{r}

cat(" GSEA: Males \n")

gene_list_male <- DEG_male_deseq %>%
  filter(!is.na(Entrez)) %>%
  arrange(desc(Rank)) %>%
  { setNames(.$Rank, as.character(.$Entrez)) } %>%
  .[!duplicated(names(.))]

gene_list_male <- sort(gene_list_male, decreasing = TRUE)
cat("  Ranked genes:", length(gene_list_male), "\n\n")

gseKEGG_male <- gseKEGG(
  gene_list_male,
  organism     = CONFIG$kegg_organism,
  pvalueCutoff = CONFIG$pvalue_cutoff,
  verbose      = FALSE
)

if (!is.null(gseKEGG_male) && nrow(gseKEGG_male@result) > 0) {
  cat("  \u2713 Found", nrow(gseKEGG_male@result), "pathways\n\n")

  write.csv(
    gseKEGG_male@result,
    file.path(CONFIG$out_dir_male, "GSE_KEGG_Male.csv"),
    row.names = FALSE
  )

  p_male <- dotplot(
    gseKEGG_male,
    showCategory = 20,
    split = ".sign",
    title = "Male DESeq2 GSEA KEGG"
  ) +
    facet_grid(
      . ~ .sign,
      labeller = labeller(.sign = c(activated = "Up in Heat", suppressed = "Down in Heat"))
    )

  ggsave(
    file.path(CONFIG$out_dir_male, "GSE_KEGG_Male.png"),
    p_male, width = 14, height = 10, dpi = 300
  )
  print(p_male)
} else {
  cat("  No significant pathways found for Males.\n\n")
}
```

### GSEA: Females ###

```{r}

cat("GSEA: Females \n")

gene_list_female <- DEG_female_deseq %>%
  filter(!is.na(Entrez)) %>%
  arrange(desc(Rank)) %>%
  { setNames(.$Rank, as.character(.$Entrez)) } %>%
  .[!duplicated(names(.))]

gene_list_female <- sort(gene_list_female, decreasing = TRUE)
cat("  Ranked genes:", length(gene_list_female), "\n\n")

gseKEGG_female <- gseKEGG(
  gene_list_female,
  organism     = CONFIG$kegg_organism,
  pvalueCutoff = CONFIG$pvalue_cutoff,
  verbose      = FALSE
)

if (!is.null(gseKEGG_female) && nrow(gseKEGG_female@result) > 0) {
  cat("  \u2713 Found", nrow(gseKEGG_female@result), "pathways\n\n")

  write.csv(
    gseKEGG_female@result,
    file.path(CONFIG$out_dir_female, "GSE_KEGG_Female.csv"),
    row.names = FALSE
  )

  p_female <- dotplot(
    gseKEGG_female,
    showCategory = 20,
    split = ".sign",
    title = "Female DESeq2 GSEA KEGG"
  ) +
    facet_grid(
      . ~ .sign,
      labeller = labeller(.sign = c(activated = "Up in Heat", suppressed = "Down in Heat"))
    )

  ggsave(
    file.path(CONFIG$out_dir_female, "GSE_KEGG_Female.png"),
    p_female, width = 14, height = 10, dpi = 300
  )
  print(p_female)
} else {
  cat("  No significant pathways found for Females.\n\n")
}
```

# AnnotationHub: fetch the NCBI-built OrgDb for Taeniopygia guttata

```{r}

ah <- AnnotationHub()
zf_orgdb_hits <- query(ah, c("OrgDb", "Taeniopygia guttata"))
print(zf_orgdb_hits)

orgdb_ZF <- zf_orgdb_hits[[length(zf_orgdb_hits)]]

cat("Available keytypes:\n")
print(keytypes(orgdb_ZF))

```

# GO GSEA helper: runs BP, MF, and CC for one sex, combines the three dotplots
# with patchwork, and exports the combined figure

```{r}

run_go_gsea_sex <- function(gene_list, out_dir, label) {

  plots       <- list()
  results_all <- list()

  for (ont in CONFIG$go_ontologies) {
    cat("=== GO GSEA:", label, "-", ont, "===\n")

    go_res <- gseGO(
      gene_list,
      OrgDb        = orgdb_ZF,
      keyType      = CONFIG$go_key_type,
      ont          = ont,
      minGSSize    = CONFIG$go_min_gs,
      maxGSSize    = CONFIG$go_max_gs,
      pvalueCutoff = CONFIG$pvalue_cutoff,
      verbose      = FALSE
    )

    results_all[[ont]] <- go_res

    if (!is.null(go_res) && nrow(go_res@result) > 0) {
      cat("  \u2713 Found", nrow(go_res@result), ont, "terms\n\n")

      write.csv(
        go_res@result,
        file.path(out_dir, paste0("GO_GSEA_", label, "_", ont, ".csv")),
        row.names = FALSE
      )

      plots[[ont]] <- dotplot(
        go_res,
        showCategory = CONFIG$go_show_cat,
        split = ".sign",
        title = paste0(label, " - ", ont)
      ) +
        facet_grid(
          . ~ .sign,
          labeller = labeller(.sign = c(activated = "Up in Heat", suppressed = "Down in Heat"))
        )
    } else {
      cat("  No significant", ont, "terms found for", label, ".\n\n")
    }
  }

  invisible(list(results = results_all, plots = plots))
}
```

### GO GSEA: Males (BP + MF + CC, combined) ###

```{r}

gene_list_male_symbol <- DEG_male_deseq %>%
  filter(!is.na(Symbol), Symbol != "") %>%
  arrange(desc(Rank)) %>%
  { setNames(.$Rank, .$Symbol) } %>%
  .[!duplicated(names(.))]

gene_list_male_symbol <- sort(gene_list_male_symbol, decreasing = TRUE)
cat("  Ranked genes:", length(gene_list_male_symbol), "\n\n")

go_male <- run_go_gsea_sex(gene_list_male_symbol, CONFIG$out_dir_male, "Male")

if (length(go_male$plots) > 0) {
  combined_go_male <- wrap_plots(go_male$plots, ncol = 1) +
    plot_annotation(title = "Male DESeq2 GO GSEA — BP / MF / CC")

  ggsave(
    file.path(CONFIG$out_dir_male, "GO_GSEA_Male_combined.png"),
    combined_go_male, width = 14, height = 8 * length(go_male$plots), dpi = 300
  )
  print(combined_go_male)
} else {
  cat("  No GO plots to combine for Males.\n\n")
}
```

### GO GSEA: Females (BP + MF + CC, combined) ###

```{r}

gene_list_female_symbol <- DEG_female_deseq %>%
  filter(!is.na(Symbol), Symbol != "") %>%
  arrange(desc(Rank)) %>%
  { setNames(.$Rank, .$Symbol) } %>%
  .[!duplicated(names(.))]

gene_list_female_symbol <- sort(gene_list_female_symbol, decreasing = TRUE)
cat("  Ranked genes:", length(gene_list_female_symbol), "\n\n")

go_female <- run_go_gsea_sex(gene_list_female_symbol, CONFIG$out_dir_female, "Female")

if (length(go_female$plots) > 0) {
  combined_go_female <- wrap_plots(go_female$plots, ncol = 1) +
    plot_annotation(title = "Female DESeq2 GO GSEA — BP / MF / CC")

  ggsave(
    file.path(CONFIG$out_dir_female, "GO_GSEA_Female_combined.png"),
    combined_go_female, width = 14, height = 8 * length(go_female$plots), dpi = 300
  )
  print(combined_go_female)
} else {
  cat("  No GO plots to combine for Females.\n\n")
}

```

### GSEA_KEGG: Males ###

```{r}

cat(" GSEA: Males \n")

gene_list_male <- DEG_male_deseq %>%
  filter(!is.na(Entrez)) %>%
  arrange(desc(Rank)) %>%
  { setNames(.$Rank, as.character(.$Entrez)) } %>%
  .[!duplicated(names(.))]

gene_list_male <- sort(gene_list_male, decreasing = TRUE)
cat("  Ranked genes:", length(gene_list_male), "\n\n")

gseKEGG_male <- gseKEGG(
  gene_list_male,
  organism     = CONFIG$kegg_organism,
  pvalueCutoff = CONFIG$pvalue_cutoff,
  verbose      = FALSE
)

if (!is.null(gseKEGG_male) && nrow(gseKEGG_male@result) > 0) {
  cat("  \u2713 Found", nrow(gseKEGG_male@result), "pathways\n\n")

  write.csv(
    gseKEGG_male@result,
    file.path(CONFIG$out_dir_male, "GSE_KEGG_Male.csv"),
    row.names = FALSE
  )

  p_male <- dotplot(
    gseKEGG_male,
    showCategory = 20,
    split = ".sign",
    title = "Male DESeq2 GSEA KEGG"
  ) +
    facet_grid(
      . ~ .sign,
      labeller = labeller(.sign = c(activated = "Up in Heat", suppressed = "Down in Heat"))
    )

  ggsave(
    file.path(CONFIG$out_dir_male, "GSE_KEGG_Male.png"),
    p_male, width = 14, height = 10, dpi = 300
  )
  print(p_male)
} else {
  cat("  No significant pathways found for Males.\n\n")
}
```

### GSEA_KEGG: Females ###

```{r}

cat("GSEA: Females \n")

gene_list_female <- DEG_female_deseq %>%
  filter(!is.na(Entrez)) %>%
  arrange(desc(Rank)) %>%
  { setNames(.$Rank, as.character(.$Entrez)) } %>%
  .[!duplicated(names(.))]

gene_list_female <- sort(gene_list_female, decreasing = TRUE)
cat("  Ranked genes:", length(gene_list_female), "\n\n")

gseKEGG_female <- gseKEGG(
  gene_list_female,
  organism     = CONFIG$kegg_organism,
  pvalueCutoff = CONFIG$pvalue_cutoff,
  verbose      = FALSE
)

if (!is.null(gseKEGG_female) && nrow(gseKEGG_female@result) > 0) {
  cat("  \u2713 Found", nrow(gseKEGG_female@result), "pathways\n\n")

  write.csv(
    gseKEGG_female@result,
    file.path(CONFIG$out_dir_female, "GSE_KEGG_Female.csv"),
    row.names = FALSE
  )

  p_female <- dotplot(
    gseKEGG_female,
    showCategory = 20,
    split = ".sign",
    title = "Female DESeq2 GSEA KEGG"
  ) +
    facet_grid(
      . ~ .sign,
      labeller = labeller(.sign = c(activated = "Up in Heat", suppressed = "Down in Heat"))
    )

  ggsave(
    file.path(CONFIG$out_dir_female, "GSE_KEGG_Female.png"),
    p_female, width = 14, height = 10, dpi = 300
  )
  print(p_female)
} else {
  cat("  No significant pathways found for Females.\n\n")
}

```

# Load Data Files for WGCNA 

# Built from the raw count matrix + Control/Heat pheno file, subset to Treatment A & B (Control/Heat) and split by sex. CORT is pulled in separately from the master spreadsheet, joined by bird ID extracted from the sample column names.

```{r}

# Raw counts + Control/Heat pheno data 

Raw_Counts <- as.matrix(read.csv(CONFIG$wgcna_raw_counts_csv, row.names = "gene_id"))
ZF_Liver_sample_list <- read.table(
  CONFIG$wgcna_pheno_txt, header = TRUE, row.names = 1, sep = "\t"
)

common_samples <- intersect(rownames(ZF_Liver_sample_list), colnames(Raw_Counts))
cat("Samples in pheno file:", nrow(ZF_Liver_sample_list),
    "| Samples in count matrix:", ncol(Raw_Counts),
    "| Common:", length(common_samples), "\n\n")

Raw_Counts_AB <- Raw_Counts[, common_samples]
ZF_Liver_sample_list <- ZF_Liver_sample_list[common_samples, ]
ZF_Liver_sample_list <- ZF_Liver_sample_list[colnames(Raw_Counts_AB), ]
stopifnot(identical(rownames(ZF_Liver_sample_list), colnames(Raw_Counts_AB)))

##  Set reference levels (Control/Heat only)
ZF_Liver_sample_list$sex       <- relevel(factor(ZF_Liver_sample_list$sex), ref = "female")
ZF_Liver_sample_list$treatment <- relevel(factor(ZF_Liver_sample_list$treatment), ref = "Control")

##  Pull CORT from the master spreadsheet, joined by bird ID

master_sheet <- read.csv(CONFIG$wgcna_master_csv, stringsAsFactors = FALSE)

bird_id <- sub("^(ZF_[0-9]+)_.*$", "\\1", rownames(ZF_Liver_sample_list))
ZF_Liver_sample_list$ID_Band <- bird_id

ZF_Liver_sample_list$ID_Band <- sub("^ZF_", "", ZF_Liver_sample_list$ID_Band)

ZF_Liver_sample_list[[CONFIG$wgcna_cort_col]] <- master_sheet[[CONFIG$wgcna_cort_col]][
  match(ZF_Liver_sample_list$ID_Band, master_sheet$ID_Band)
]

cat("CORT values matched:",
    sum(!is.na(ZF_Liver_sample_list[[CONFIG$wgcna_cort_col]])),
    "/", nrow(ZF_Liver_sample_list), "\n\n")

##  Split by sex 
AHDB1_AB_Male   <- ZF_Liver_sample_list[!grepl("female", ZF_Liver_sample_list$sex), ]
AHDB1_AB_Female <- ZF_Liver_sample_list[!grepl("^male$", ZF_Liver_sample_list$sex), ]

Counts_AB_Male   <- Raw_Counts_AB[, rownames(AHDB1_AB_Male)]
Counts_AB_Female <- Raw_Counts_AB[, rownames(AHDB1_AB_Female)]

stopifnot(identical(rownames(AHDB1_AB_Male),   colnames(Counts_AB_Male)))
stopifnot(identical(rownames(AHDB1_AB_Female), colnames(Counts_AB_Female)))

cat("Male samples:",   ncol(Counts_AB_Male),   "| genes:", nrow(Counts_AB_Male),   "\n")
cat("Female samples:", ncol(Counts_AB_Female), "| genes:", nrow(Counts_AB_Female), "\n\n")
```

# WGCNA helper function 

```{r}

run_wgcna <- function(counts, meta, out_dir, label) {

  cat(" WGCNA:", label, "\n")

## Filter low-expression genes
  keep <- rowSums(counts >= CONFIG$wgcna_min_count) >= CONFIG$wgcna_min_samples
  counts_filt <- counts[keep, ]
  cat("Genes after filtering:", nrow(counts_filt), "/", nrow(counts), "\n")

  ## Variance-stabilizing transformation
  dds <- DESeqDataSetFromMatrix(
    countData = counts_filt,
    colData   = meta,
    design    = ~ 1
  )
  vsd <- vst(dds, blind = TRUE)
  expr_mat <- t(assay(vsd))   # WGCNA wants samples-in-rows, genes-in-columns

  ## QC
  gsg <- goodSamplesGenes(expr_mat, verbose = 3)
  if (!gsg$allOK) {
    cat("  Removing", sum(!gsg$goodGenes), "bad genes and",
        sum(!gsg$goodSamples), "bad samples\n")
    expr_mat <- expr_mat[gsg$goodSamples, gsg$goodGenes]
  }
  cat("  Final matrix:", nrow(expr_mat), "samples x", ncol(expr_mat), "genes\n\n")

  ##  Sample clustering 
  sample_tree <- hclust(dist(expr_mat), method = "average")
  pdf(file.path(out_dir, paste0(label, "_sample_clustering.pdf")), width = 10, height = 6)
  plot(sample_tree, main = paste(label, "- Sample clustering to detect outliers"),
       sub = "", xlab = "")
  dev.off()

  ##  Pick soft-thresholding power 
  sft <- pickSoftThreshold(
    expr_mat,
    powerVector = CONFIG$wgcna_power_test_range,
    networkType = CONFIG$wgcna_network_type,
    verbose     = 0
  )

  pdf(file.path(out_dir, paste0(label, "_soft_threshold.pdf")), width = 10, height = 5)
  par(mfrow = c(1, 2))
  plot(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
       xlab = "Soft Threshold (power)", ylab = "Scale Free Topology Model Fit, R^2",
       type = "n", main = paste(label, "- Scale independence"))
  text(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
       labels = CONFIG$wgcna_power_test_range, col = "red")
  abline(h = 0.90, col = "red")
  plot(sft$fitIndices[, 1], sft$fitIndices[, 5],
       xlab = "Soft Threshold (power)", ylab = "Mean Connectivity",
       type = "n", main = paste(label, "- Mean connectivity"))
  text(sft$fitIndices[, 1], sft$fitIndices[, 5], labels = CONFIG$wgcna_power_test_range, col = "red")
  dev.off()

  chosen_power <- sft$powerEstimate
  if (is.na(chosen_power)) {
    cat("  No power reached R^2 > 0.9 automatically \n")
    chosen_power <- 6 # Need to choose   
  }
  cat("  Chosen soft power:", chosen_power, "\n\n")

  ##  Network construction
  net <- blockwiseModules(
    expr_mat,
    power              = chosen_power,
    networkType        = CONFIG$wgcna_network_type,
    TOMType             = CONFIG$wgcna_network_type,
    minModuleSize      = CONFIG$wgcna_min_module_size,
    mergeCutHeight     = CONFIG$wgcna_merge_cut_height,
    numericLabels      = TRUE,
    pamRespectsDendro  = FALSE,
    saveTOMs           = FALSE,
    verbose            = 0
  )

  module_labels  <- net$colors
  module_colors  <- labels2colors(module_labels)
  module_eigengenes <- net$MEs

  cat("  Modules found:", length(unique(module_colors)), "\n")
  print(table(module_colors))

  pdf(file.path(out_dir, paste0(label, "_dendrogram.pdf")), width = 10, height = 6)
  plotDendroAndColors(
    net$dendrograms[[1]], module_colors[net$blockGenes[[1]]],
    "Module colors", dendroLabels = FALSE, hang = 0.03,
    addGuide = TRUE, guideHang = 0.05,
    main = paste(label, "- Gene dendrogram and module colors")
  )
  dev.off()

  ##  Export gene -> module assignment
  gene_module_df <- data.frame(
    Gene_ID = colnames(expr_mat),
    Module  = module_colors
  )
  write.csv(
    gene_module_df,
    file.path(out_dir, paste0(label, "_gene_modules.csv")),
    row.names = FALSE
  )

  ##  Module-trait correlations: WITHOUT CORT
  treatment_numeric <- as.numeric(factor(meta$treatment)) - 1  # Control=0, Heat=1 
  traits_no_cort <- data.frame(Treatment = treatment_numeric)
  rownames(traits_no_cort) <- rownames(meta)
  traits_no_cort <- traits_no_cort[rownames(expr_mat), , drop = FALSE]

  mod_trait_cor_no_cort <- cor(module_eigengenes, traits_no_cort, use = "p")
  mod_trait_p_no_cort   <- corPvalueStudent(mod_trait_cor_no_cort, nrow(expr_mat))

  write.csv(
    as.data.frame(mod_trait_cor_no_cort),
    file.path(out_dir, paste0(label, "_module_trait_cor_noCORT.csv"))
  )
  write.csv(
    as.data.frame(mod_trait_p_no_cort),
    file.path(out_dir, paste0(label, "_module_trait_pval_noCORT.csv"))
  )

  ##  Module-trait correlations: WITH CORT
  if (CONFIG$wgcna_include_cort && CONFIG$wgcna_cort_col %in% colnames(meta)) {
    traits_with_cort <- data.frame(
      Treatment = treatment_numeric,
      CORT      = as.numeric(meta[[CONFIG$wgcna_cort_col]])
    )
    rownames(traits_with_cort) <- rownames(meta)
    traits_with_cort <- traits_with_cort[rownames(expr_mat), , drop = FALSE]

    mod_trait_cor_cort <- cor(module_eigengenes, traits_with_cort, use = "p")
    mod_trait_p_cort   <- corPvalueStudent(mod_trait_cor_cort, nrow(expr_mat))

    write.csv(
      as.data.frame(mod_trait_cor_cort),
      file.path(out_dir, paste0(label, "_module_trait_cor_withCORT.csv"))
    )
    write.csv(
      as.data.frame(mod_trait_p_cort),
      file.path(out_dir, paste0(label, "_module_trait_pval_withCORT.csv"))
    )

    ## Module-trait heatmap (with CORT)
    text_matrix <- paste0(
      signif(mod_trait_cor_cort, 2), "\n(", signif(mod_trait_p_cort, 1), ")"
    )
    dim(text_matrix) <- dim(mod_trait_cor_cort)

    pdf(file.path(out_dir, paste0(label, "_module_trait_heatmap_withCORT.pdf")),
        width = 8, height = max(6, 0.3 * nrow(mod_trait_cor_cort)))
    labeledHeatmap(
      Matrix         = mod_trait_cor_cort,
      xLabels        = colnames(traits_with_cort),
      yLabels        = rownames(mod_trait_cor_cort),
      ySymbols       = rownames(mod_trait_cor_cort),
      colorLabels    = FALSE,
      colors         = blueWhiteRed(50),
      textMatrix     = text_matrix,
      setStdMargins  = FALSE,
      cex.text       = 0.6,
      zlim           = c(-1, 1),
      main           = paste(label, "- Module-trait relationships (with CORT)")
    )
    dev.off()
  } else {
    cat("  Skipping CORT run — CORT column not found or include_cort = FALSE\n")
  }

  ##  Hub gene identification 
  kme <- signedKME(expr_mat, module_eigengenes, outputColumnName = "kME_")
  hub_df <- data.frame(
    Gene_ID = colnames(expr_mat),
    Module  = module_colors,
    kme
  )
  write.csv(
    hub_df,
    file.path(out_dir, paste0(label, "_module_membership_kME.csv")),
    row.names = FALSE
  )

  invisible(list(
    net = net,
    module_colors = module_colors,
    module_eigengenes = module_eigengenes,
    expr_mat = expr_mat
  ))
}
```

# Run: Males

```{r}

wgcna_male <- run_wgcna(
  counts  = Counts_AB_Male,
  meta    = AHDB1_AB_Male,
  out_dir = CONFIG$out_dir_male,
  label   = "Male"
)

```

# Run: Females

```{r}
wgcna_female <- run_wgcna(
  counts  = Counts_AB_Female,
  meta    = AHDB1_AB_Female,
  out_dir = CONFIG$out_dir_female,
  label   = "Female"
)
```


