---
title: "6_DGE_DESeq2"
author: "Brynleigh Payne"
date: "2026-06-16"
output: html_document
---

####  DESeq2 Script for Differential Gene Expression Analysis in Functional Genomics BIOL: 6850
### Resources and Citations:
# Love et al 2016 DESeq2 GenomeBiology
# https://bioconductor.riken.jp/packages/2.14/bioc/vignettes/DESeq2/inst/doc/DESeq2.pdf
# http://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html

#### Install the DESeq2 package if you have not already

```{r}
#if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
#BiocManager::install("DESeq2")
library(DESeq2)
```

#### Input the count data, the gene(/transcript) count matrix and labels

```{r}
countdata <- as.matrix(read.csv("gene_count_matrix (3).csv", row.names="gene_id"))
dim(countdata)
head(countdata)
```

#### Input the meta data or phenotype data
# Note: The PHENO_DATA file contains information on each sample, e.g., sex or population. The exact way to import this depends on the format of the file.
##  Make sure the individual names match between the count data and the PHENO_DATA

```{r}
coldata <-(read.table("PHENO_DATA_Blood_Control_Heat.txt", header=TRUE, row.names=1, sep = "\t"))
dim(coldata)
head(coldata)

```

## Remove Baseline samples

```{r}
coldata_filtered <- coldata[!grepl("^Baseline", coldata$treatment), ]
head(coldata_filtered)
```

## Subset the countdata to only include Control (A) and Heat (B) samples 

```{r}
samples_to_keep <- rownames(coldata_filtered)

length(intersect(samples_to_keep, colnames(countdata)))

all(samples_to_keep %in% colnames(countdata))

setdiff(samples_to_keep, colnames(countdata))

setdiff(colnames(countdata), samples_to_keep)

common_samples <- intersect(samples_to_keep, colnames(countdata))

countdata_subset <- countdata[, common_samples]
coldata_subset <- coldata_filtered[common_samples, ]

coldata_subset <- coldata_subset[colnames(countdata_subset), ]

identical(rownames(coldata_subset), colnames(countdata_subset))
```

# Check all sample IDs in colData are also in CountData and match their orders

```{r}
all(rownames(coldata_subset) %in% colnames(countdata_subset))
countdata_subset <- countdata_subset[, rownames(coldata_subset)]
all(rownames(coldata_subset) == colnames(countdata_subset))
```

## Manually set reference levels for sex and treatment

```{r}
coldata_subset$sex <- relevel(factor(coldata_subset$sex), ref = "female")
coldata_subset$treatment <- relevel(factor(coldata_subset$treatment), ref = "Control")
```

#### Create the DESEQ dataset and define the statistical model
# Interaction Model

```{r}
dds <- DESeqDataSetFromMatrix(countData = countdata_subset, colData=coldata_subset,  design = ~ 0 + sex + treatment + sex:treatment)
dds
```

##   Prefiltering 
# Here we perform a minimal pre-filtering to remove rows that have less than 20 reads mapped.
# You can play around with this number to see how it affects your results!

```{r}
dds <- dds[ rowSums(counts(dds)) > 20, ]

# How many genes were filtered out?
dds
```

## Set factors for statistical analyses

```{r}
dds$sex <- factor(dds$sex, levels=c("female","male"))
dds$treatment <- factor(dds$treatment, levels=c("Control","Heat"))
```

## Differential expression analysis of the Interaction Model

```{r}
dds <- DESeq(dds)
res <- results(dds)
res
```

# Order our results table by the smallest adjusted p value:

```{r}
resOrdered <- res[order(res$padj),]
resOrdered
```

# We can summarize some basic tallies using the summary function the default is p<0.1.

```{r}
summary(res)
```

# How many adjusted p-values were less than 0.1?

```{r}
sum(res$padj < 0.1, na.rm=TRUE)
```

# If the adjusted p value will be a value other than 0.1, alpha should be set to that value:

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
sum(res05$padj < 0.05, na.rm=TRUE)
```

# plotMA shows the log2 fold changes attributable to a given variable over the meanof normalized counts. 
# Points will be colored red if the adjusted p value is less than 0.1. 
# Points which fall out of the window are plotted as open triangles pointing either up or down
  
```{r}
plotMA(res, main="DESeq2 Interaction Model", ylim=c(-8,8))
```
  
#### Create the DESEQ dataset and define the statistical model
# Female Model
```{r}
# Subset females from pheno data
coldata_female_subset <- coldata_subset[coldata_subset$sex == "female", ]
head(coldata_female_subset)

# Subset counts to match
keep_samples <- rownames(coldata_female_subset)

countdata_female_subset <- countdata_subset[, keep_samples]

# Sanity check
coldata_female_subset <- coldata_female_subset[colnames(countdata_female_subset), ]

identical(rownames(coldata_female_subset), colnames(countdata_female_subset))

# Create dds2 for females
dds2 <- DESeqDataSetFromMatrix(countData = countdata_female_subset, colData=coldata_female_subset,  design = ~ 0 + treatment)
dds2
```

##   Prefiltering 
# Here we perform a minimal pre-filtering to remove rows that have less than 20 reads mapped.
# You can play around with this number to see how it affects your results!

```{r}
dds2 <- dds2[ rowSums(counts(dds2)) > 20, ]

# How many genes were filtered out?
dds2
```

## Set factors for statistical analyses

```{r}
dds2$treatment <- factor(dds2$treatment, levels=c("Control","Heat"))
```

## Differential expression analysis of the Female model

```{r}
dds2 <- DESeq(dds2)
res2 <- results(dds2)
res2
```

# Order our results table by the smallest adjusted p value:

```{r}
resOrdered2 <- res2[order(res2$padj),]
resOrdered2
```

# We can summarize some basic tallies using the summary function the default is p<0.1.

```{r}
summary(res2)
```

# How many adjusted p-values were less than 0.1?

```{r}
sum(res2$padj < 0.1, na.rm=TRUE)
```

# How many adjusted p-values were less than 0.05?

```{r}
sum(res2$padj < 0.05, na.rm=TRUE)
```

# plotMA shows the log2 fold changes attributable to a given variable over the meanof normalized counts. 
# Points will be colored red if the adjusted p value is less than 0.1. 
# Points which fall out of the window are plotted as open triangles pointing either up or down
  
```{r}
plotMA(res2, main="DESeq2 Female Model", ylim=c(-8,8))
```

#### Create the DESEQ dataset and define the statistical model
# Male Model
```{r}
# Subset males from pheno data
coldata_male_subset <- coldata_subset[coldata_subset$sex == "male", ]
head(coldata_male_subset)

# Subset counts to match
keep_samples <- rownames(coldata_male_subset)

countdata_male_subset <- countdata_subset[, keep_samples]

# Sanity check
coldata_male_subset <- coldata_male_subset[colnames(countdata_male_subset), ]

identical(rownames(coldata_male_subset), colnames(countdata_male_subset))

# Create dd32 for males
dds3 <- DESeqDataSetFromMatrix(countData = countdata_male_subset, colData=coldata_male_subset,  design = ~ 0 + treatment)
dds3
```

##   Prefiltering 
# Here we perform a minimal pre-filtering to remove rows that have less than 20 reads mapped.
# You can play around with this number to see how it affects your results!

```{r}
dds3 <- dds3[ rowSums(counts(dds3)) > 20, ]

# How many genes were filtered out?
dds3
```

## Set factors for statistical analyses

```{r}
dds3$treatment <- factor(dds3$treatment, levels=c("Control","Heat"))
```

## Differential expression analysis of the Male model

```{r}
dds3 <- DESeq(dds3)
res3 <- results(dds3)
res3
```

# We can summarize some basic tallies using the summary function the default is p<0.1.

```{r}
summary(res3)
```

# How many adjusted p-values were less than 0.1?

```{r}
sum(res3$padj < 0.1, na.rm=TRUE)
```

# How many adjusted p-values were less than 0.1?

```{r}
sum(res3$padj < 0.05, na.rm=TRUE)
```

# plotMA shows the log2 fold changes attributable to a given variable over the meanof normalized counts. 
# Points will be colored red if the adjusted p value is less than 0.1. 
# Points which fall out of the window are plotted as open triangles pointing either up or down
  
```{r}
plotMA(res3, main="DESeq2 Male Model", ylim=c(-8,8))
```













#After calling plotMA, one can use the function identify to interactively detect the row number of individual genes by clicking on the plot. 
# One can then recover the gene identiers by saving the resulting indices:
  idx <- identify(res$baseMean, res$log2FoldChange)
# after selecting a gene. You need to press escape to move on
  rownames(res)[idx]

    
##  1.5.2 Plot counts - sanity check!
  
# You can select the gene to plot by rowname or by numeric index.
  plotCounts(dds, gene="FUN_025750", intgroup="treatment")
# You can plot the gene with the lowest adjuated P-value
  plotCounts(dds, gene=which.min(res$padj), intgroup="treatment")
  dds
  
##  Write your results to a file 
  write.csv(as.data.frame(resOrdered), file="DGESeq_results.csv")  
  
## 2.1.2 Extracting transformed values
  rld <- rlog(dds)
  vsd <- varianceStabilizingTransformation(dds)
  head(assay(rld), 3)
  
### Heatmap of the count matrix
#library("genefilter")
  topVarGenes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 50)
  
  library("pheatmap")
  mat  <- assay(vsd)[ topVarGenes, ]
  mat  <- mat - rowMeans(mat)
  anno <- as.data.frame(colData(vsd)[, c("treatment", "type")])
  df <- as.data.frame(colData(dds)[,c("treatment","type")])
    pheatmap(mat, annotation_col = anno)
  
  
## 2.2.1 Heatmap of the count matrix
#  library("pheatmap")
#  select <- order(rowMeans(counts(dds,normalized=TRUE)),decreasing=TRUE)[1:50]
#  nt <- normTransform(dds) # defaults to log2(x+1)
#  log2.norm.counts <- assay(nt)[select,]
#   df <- as.data.frame(colData(dds)[,c("treatment","type")])
#  pheatmap(assay(vsd)[mat,], cluster_rows=TRUE, show_rownames=TRUE,
#          cluster_cols=TRUE, annotation_col=df)
  
# pheatmap(log2.norm.counts, cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df)
#  pheatmap(assay(rld)[select,], cluster_rows=FALSE, show_rownames=FALSE,  cluster_cols=FALSE, annotation_col=df) 
  
  
#2.2.2 Heatmap of the sample-to-sample distances
  sampleDists <- dist(t(assay(rld)))
  library("RColorBrewer")
  sampleDistMatrix <- as.matrix(sampleDists)
  rownames(sampleDistMatrix) <- paste(rld$treatment)
  colnames(sampleDistMatrix) <- NULL
  colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
  pheatmap(sampleDistMatrix,
           clustering_distance_rows=sampleDists,
           clustering_distance_cols=sampleDists,
           col=colors)
  
  
# 2.2.3 Principal component plot of the samples
  plotPCA(rld, intgroup=c("treatment"))
  


############ Preparing Data for GSEA and Cytoscape.  #############

### Merge 'gene names' with DGE results by Gene Model

## Import Annotation file with results from Blast to databases
Anno <- read.csv("Daphnia_pulex.annotations_Name.csv", stringsAsFactors = FALSE, na.strings=c(""))
summary(Anno)
dim(Anno)

## Import the DGE results file make sure the gene model name is 'gene_id' to match annotation file
DGEresults <- read.csv("DGESeq_results.csv", stringsAsFactors = FALSE)
summary(DGEresults)
dim(DGEresults)

## Rename first column so it matches "gene_id" in annotation file
names(DGEresults)[1]<- "gene_id" 

#Merge anno with DGE results
DGE_Anno <- merge(Anno,DGEresults,by="gene_id",all.y = FALSE)
dim(DGE_Anno)
summary(DGE_Anno)

  

############################# Make ranked list for GSEA ####################
## example aa <- within(resOrdered, z <- x + y - 2)
DGE_Anno_Rank <-  within(DGE_Anno, rank <- sign(log2FoldChange) * -log10(pvalue))
DGE_Anno_Rank 

#subset the results so only Gene Name and rank
DGErank = subset(DGE_Anno_Rank, select = c(Name,rank) )
DGErank

#subset the results so only Gene Name and rank
DGErank_withName <- na.omit(DGErank)
DGErank_withName
dim(DGErank_withName)



### IF NEEDED....remove the "gene-" from row names
## https://stackoverflow.com/questions/39897155/replace-and-remove-part-of-string-in-rownames/39897315  "URS000075AF9C-snoRNA_GTATGTGTGGACAGCACTGAGACTGAGTCT"    to   "snoRNA"
## We can use gsub to match one of more characters that are not a - ([^-]+) from the start (^) of the string followed by 
## a - or (|) one or more characters that are not an underscore ([^_]+) until the end of the string ($)  and replace it with blanks ("").
## gsub("^[^-]+-|_[^_]+$", "", rownames(df))

#gene <-gsub("^[^-]+-", "", rownames(DGErank))
#DGErankIDs  <-cbind(gene,DGErank)
#head(DGErankIDs)
#summary(DGErankIDs)  

#write.csv(as.data.frame(DGErankIDs), file="DGErankIDs.csv", row.names=FALSE)  
write.table(as.data.frame(DGErank_withName), file="DGErankName.rnk", quote=FALSE, row.names=FALSE, sep = "\t")  

####  We also need the normalized expression DATA
nt <- normTransform(dds) # defaults to log2(x+1)
head(assay(nt))
# compare to original count data
head(countdata)
# make it a new dataframe
NormTransExp<-assay(nt)
summary(NormTransExp)
head(NormTransExp)

## Rename first column so it matches "gene_id" in annotation file
gene_id <-gsub("^[^-]+-", "", rownames(NormTransExp))
NormTransExpIDs  <-cbind(gene_id,NormTransExp)
head(NormTransExpIDs)

## Rename first column so it matches "gene_id" in annotation file
#names(NormTransExp)[1]<- "gene_id"
#head(NormTransExp)

#Merge anno with DGE results
NormTransExp_Anno <- merge(Anno,NormTransExpIDs,by="gene_id",all.y = FALSE)
dim(NormTransExp_Anno)
summary(NormTransExp_Anno)
head(NormTransExp_Anno)

#remove FunIDs
#NormTransExp_Anno = NormTransExp_Anno[,-1]
# Move the FunIDs to column 2

## Swap column 1 and column 2
cols <- colnames(NormTransExp_Anno)  # get all column names
cols[c(1, 2)] <- cols[c(2, 1)]  # Swap column names
NormTransExp_Anno <- NormTransExp_Anno[, cols]  # Reorder the dataframe based on the new column order
head(NormTransExp_Anno)

#subset the results so only Genes with Name
NormTransExp_Anno_withName <- na.omit(NormTransExp_Anno)
head(NormTransExp_Anno_withName)
dim(NormTransExp_Anno_withName)

#write.csv(as.data.frame(NormTransExpIDs), file="NormTransExpressionData.csv", row.names=FALSE)  
write.table(as.data.frame(NormTransExp_Anno_withName), file="NormTransExp_Anno_Names.txt", quote=FALSE, row.names=FALSE, sep = "\t")  