---
title: "6_DGE_DESeq2_LimmaVoom_EdgeR_Comparison"
author: "Ali Amer, modified by Brynleigh Payne"
date: "2026-06-22"
output: html_document
---

# Script Goal: 
# 1. Convert raw count reads of samples sent for liver RNA-Seq (A vs B) from prep.dy script   
# 2. Check library sizes across A vs B & Normalize raw count data
# 3. Run DGE with edgeR, limma and DESeq2 (Males Only & Females Only)
# 4. Run Pathway Enrichment & WGCNA

#### Input the count data, the gene(/transcript) count matrix and labels

```{r}
Raw_Counts <- as.matrix(read.csv("gene_count_matrix (3).csv", row.names="gene_id"))
dim(Raw_Counts)
head(Raw_Counts)
```

#### Input the meta data or phenotype data
# Note: The PHENO_DATA file contains information on each sample, e.g., sex or population. The exact way to import this depends on the format of the file.
##  Make sure the individual names match between the count data and the PHENO_DATA

```{r}
ZF_Blood_sample_list <-(read.table("PHENO_DATA_Blood_Control_Heat.txt", header=TRUE, row.names=1, sep = "\t"))
dim(ZF_Blood_sample_list)
head(ZF_Blood_sample_list)
```

## Remove Baseline samples

```{r}
ZF_Blood_sample_list <- ZF_Blood_sample_list[!grepl("^Baseline", ZF_Blood_sample_list$treatment), ]
head(ZF_Blood_sample_list)
```

## Subset the countdata to only include Control (A) and Heat (B) samples 

```{r}
samples_to_keep <- rownames(ZF_Blood_sample_list)

length(intersect(samples_to_keep, colnames(Raw_Counts)))

all(samples_to_keep %in% colnames(Raw_Counts))

setdiff(samples_to_keep, colnames(Raw_Counts))

setdiff(colnames(Raw_Counts), samples_to_keep)

common_samples <- intersect(samples_to_keep, colnames(Raw_Counts))

Raw_Counts_AB <- Raw_Counts[, common_samples]
ZF_Blood_sample_list <- ZF_Blood_sample_list[common_samples, ]

ZF_Blood_sample_list <- ZF_Blood_sample_list[colnames(Raw_Counts_AB), ]

identical(rownames(ZF_Blood_sample_list), colnames(Raw_Counts_AB))
```

# Check all sample IDs in colData are also in CountData and match their orders

```{r}
all(rownames(ZF_Blood_sample_list) %in% colnames(Raw_Counts_AB))
Raw_Counts_AB <- Raw_Counts_AB[, rownames(ZF_Blood_sample_list)]
all(rownames(ZF_Blood_sample_list) == colnames(Raw_Counts_AB))
```

## Manually set reference levels for sex and treatment

```{r}
ZF_Blood_sample_list$sex <- relevel(factor(ZF_Blood_sample_list$sex), ref = "female")
ZF_Blood_sample_list$treatment <- relevel(factor(ZF_Blood_sample_list$treatment), ref = "Control")
```

# Construct DGEList object
```{r}
DGE_AB <- DGEList(
  counts  = Raw_Counts_AB,
  samples = ZF_Blood_sample_list
)
```

# Calculate Library Size & Make BoxPlots (A vs B)

```{r}

cpm_AB <- cpm(DGE_AB)
lcpm_AB <- cpm(DGE_AB, log=TRUE)
summary(lcpm_AB)

# Boxplot by Treatment

boxplot(
  DGE_AB$samples$lib.size/1e6 ~ DGE_AB$samples$treatment,
  ylab = "Library size (millions)",
  main = "Library Size by Treatment (Control vs Heat)",
  col = c("lightblue", "coral"),
  outline = FALSE
)


# Boxplot by Treatment & Sex

boxplot(
  DGE_AB$samples$lib.size/1e6 ~ interaction(DGE_AB$samples$treatment, DGE_AB$samples$sex),
  ylab = "Library size (millions)",
  main = "Library Size by Treatment and Sex (Control vs Heat)",
  las = 3,
  col = rep(c("lightblue", "coral"), 2),
  outline = FALSE
)

# Display
par(mfrow = c(1, 2))
boxplot(
  DGE_AB$samples$lib.size/1e6 ~ DGE_AB$samples$treatment,
  ylab = "Library size (millions)",
  main = "Library Size by Treatment",
  col = c("lightblue", "coral"),
  outline = FALSE
)

boxplot(
  DGE_AB$samples$lib.size/1e6 ~ interaction(DGE_AB$samples$treatment, DGE_AB$samples$sex),
  ylab = "Library size (millions)",
  main = "Library Size by Treatment and Sex",
  xlab = "",
  las = 3,
  col = rep(c("lightblue", "coral"), 2),
  outline = FALSE
)

```

### DGE with limma-voom ###

```{r}

library(limma)

### Males only

## Remove females samples
AHDB1_AB_Male <- ZF_Blood_sample_list[!grepl("female", ZF_Blood_sample_list$sex), ]
head(AHDB1_AB_Male)

## Subset the countdata to only include Control (A) and Heat (B) samples 
samples_to_keep <- rownames(AHDB1_AB_Male)

length(intersect(samples_to_keep, colnames(Raw_Counts_AB)))

all(samples_to_keep %in% colnames(Raw_Counts_AB))

setdiff(samples_to_keep, colnames(Raw_Counts_AB))

setdiff(colnames(Raw_Counts_AB), samples_to_keep)

common_samples <- intersect(samples_to_keep, colnames(Raw_Counts_AB))

Counts_AB_Male <- Raw_Counts_AB[, common_samples]
AHDB1_AB_Male <- AHDB1_AB_Male[common_samples, ]

AHDB1_AB_Male <- AHDB1_AB_Male[colnames(Counts_AB_Male), ]

identical(rownames(AHDB1_AB_Male), colnames(Counts_AB_Male))

dge_AB_Male_voom <- DGEList(
  counts  = Counts_AB_Male,
  samples = AHDB1_AB_Male
)

# Sanity Check
table(dge_AB_Male_voom$samples$treatment)
cat("Male-Only Subset\n")
cat("Total genes (before filtering):", nrow(dge_AB_Male_voom), "\n\n")

dge_AB_Male_voom$samples$treatment <- droplevels(factor(dge_AB_Male_voom$samples$treatment))

design_male_voom <- model.matrix(~ 0 + treatment, data = dge_AB_Male_voom$samples)
colnames(design_male_voom) <- levels(dge_AB_Male_voom$samples$treatment)

keep_male_voom <- filterByExpr(dge_AB_Male_voom, design_male_voom)
dge_AB_Male_voom <- dge_AB_Male_voom[keep_male_voom, , keep.lib.sizes = FALSE]

cat("Genes after filtering:", nrow(dge_AB_Male_voom), "\n\n")

dge_AB_Male_voom <- calcNormFactors(dge_AB_Male_voom, method = "TMM")

v_male <- voom(dge_AB_Male_voom, design_male_voom, plot = TRUE)

fit_male_voom <- lmFit(v_male, design_male_voom)

contrast_male_voom <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_male_voom
)
cat("Contrast:\n")
print(contrast_male_voom)

fit_male_voom <- contrasts.fit(fit_male_voom, contrast_male_voom)
fit_male_voom <- eBayes(fit_male_voom)

DEG_male_voom <- topTable(fit_male_voom, coef = "Heat_vs_Control", n = Inf, sort.by = "P")
DEG_male_voom$Gene_ID <- rownames(DEG_male_voom)

sig_male_voom <- DEG_male_voom[DEG_male_voom$adj.P.Val < 0.1, ]
sig_male_voom_up <- sig_male_voom[sig_male_voom$logFC > 0, ]
sig_male_voom_down <- sig_male_voom[sig_male_voom$logFC < 0, ]

cat("Male-Only DEG Results 0.1 (limma-voom)\n")
cat("Total significant DEGs:", nrow(sig_male_voom), "\n")
cat("Upregulated in Heat:", nrow(sig_male_voom_up), "\n")
cat("Downregulated in Heat:", nrow(sig_male_voom_down), "\n\n")

stringent_sig_male_voom <- DEG_male_voom[DEG_male_voom$adj.P.Val < 0.05, ]
stringent_sig_male_voom_up <- stringent_sig_male_voom[stringent_sig_male_voom$logFC > 0, ]
stringent_sig_male_voom_down <- stringent_sig_male_voom[stringent_sig_male_voom$logFC < 0, ]

cat("Male-Only DEG Results 0.05 (limma-voom)\n")
cat("Total significant DEGs:", nrow(stringent_sig_male_voom), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_male_voom_up), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_male_voom_down), "\n\n")


# Plot mean variance trend
v_male <- voom(dge_AB_Male_voom, design_male_voom, plot = TRUE)


### Females only

## Remove males samples
AHDB1_AB_Female <- ZF_Blood_sample_list[!grepl("^male$", ZF_Blood_sample_list$sex), ]
head(AHDB1_AB_Female)

## Subset the countdata to only include female Control (A) and Heat (B) samples 
samples_to_keep_female <- rownames(AHDB1_AB_Female)

length(intersect(samples_to_keep_female, colnames(Raw_Counts_AB)))

all(samples_to_keep_female %in% colnames(Raw_Counts_AB))

setdiff(samples_to_keep_female, colnames(Raw_Counts_AB))

setdiff(colnames(Raw_Counts_AB), samples_to_keep_female)

common_samples <- intersect(samples_to_keep_female, colnames(Raw_Counts_AB))

Counts_AB_Female <- Raw_Counts_AB[, common_samples]
AHDB1_AB_Female <- AHDB1_AB_Female[common_samples, ]

AHDB1_AB_Female <- AHDB1_AB_Female[colnames(Counts_AB_Female), ]

identical(rownames(AHDB1_AB_Female), colnames(Counts_AB_Female))

dge_AB_Female_voom <- DGEList(
  counts  = Counts_AB_Female,
  samples = AHDB1_AB_Female
)

# Sanity Check
table(dge_AB_Female_voom$samples$treatment)
cat("Female-Only Subset\n")
cat("Total genes (before filtering):", nrow(dge_AB_Female_voom), "\n\n")

dge_AB_Female_voom$samples$treatment <- droplevels(factor(dge_AB_Female_voom$samples$treatment))

design_female_voom <- model.matrix(~ 0 + treatment, data = dge_AB_Female_voom$samples)
colnames(design_female_voom) <- levels(dge_AB_Female_voom$samples$treatment)

keep_female_voom <- filterByExpr(dge_AB_Female_voom, design_female_voom)
dge_AB_Female_voom <- dge_AB_Female_voom[keep_female_voom, , keep.lib.sizes = FALSE]

cat("Genes after filtering:", nrow(dge_AB_Female_voom), "\n\n")

dge_AB_Female_voom <- calcNormFactors(dge_AB_Female_voom, method = "TMM")

v_female <- voom(dge_AB_Female_voom, design_female_voom, plot = TRUE)

fit_female_voom <- lmFit(v_female, design_female_voom)

contrast_female_voom <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_female_voom
)
cat("Contrast:\n")
print(contrast_female_voom)

fit_female_voom <- contrasts.fit(fit_female_voom, contrast_female_voom)
fit_female_voom <- eBayes(fit_female_voom)

DEG_female_voom <- topTable(fit_female_voom, coef = "Heat_vs_Control", n = Inf, sort.by = "P")
DEG_female_voom$Gene_ID <- rownames(DEG_female_voom)

sig_female_voom <- DEG_female_voom[DEG_female_voom$adj.P.Val < 0.1, ]
sig_female_voom_up <- sig_female_voom[sig_female_voom$logFC > 0, ]
sig_female_voom_down <- sig_female_voom[sig_female_voom$logFC < 0, ]

cat("Female-Only DEG Results 0.1 (limma-voom)\n")
cat("Total significant DEGs:", nrow(sig_female_voom), "\n")
cat("Upregulated in Heat:", nrow(sig_female_voom_up), "\n")
cat("Downregulated in Heat:", nrow(sig_female_voom_down), "\n\n")

stringent_sig_female_voom <- DEG_female_voom[DEG_female_voom$adj.P.Val < 0.05, ]
stringent_sig_female_voom_up <- stringent_sig_female_voom[stringent_sig_female_voom$logFC > 0, ]
stringent_sig_female_voom_down <- stringent_sig_female_voom[stringent_sig_female_voom$logFC < 0, ]

cat("Female-Only DEG Results 0.05 (limma-voom)\n")
cat("Total significant DEGs:", nrow(stringent_sig_female_voom), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_female_voom_up), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_female_voom_down), "\n\n")

# Plot mean variance trend
v_female <- voom(dge_AB_Female_voom, design_female_voom, plot = TRUE)

```

### DGE with DESeq2 ###

```{r}
library(DESeq2)


## Males only

# Build DESeq2 object
dds_male <- DESeqDataSetFromMatrix(
  countData = Counts_AB_Male,
  colData   = AHDB1_AB_Male,
  design    = ~ 0 + treatment
)

dds_male <- DESeq(dds_male)

res_male <- results(
  dds_male,
  contrast = c("treatment", "Heat", "Control"),
  alpha    = 0.05
)

summary(res_male)

DEG_male_deseq <- as.data.frame(res_male)
DEG_male_deseq$Gene_ID <- rownames(DEG_male_deseq)

# Remove NA's
DEG_male_deseq <- DEG_male_deseq[
  !is.na(DEG_male_deseq$padj) & !is.na(DEG_male_deseq$log2FoldChange),
]
cat("Genes after NA removal:", nrow(DEG_male_deseq), "\n\n")

# Subset DEG"s into upregulated and downregulated 
sig_male_deseq      <- DEG_male_deseq[DEG_male_deseq$padj < 0.1, ]
sig_male_deseq_up   <- sig_male_deseq[sig_male_deseq$log2FoldChange > 0, ]
sig_male_deseq_down <- sig_male_deseq[sig_male_deseq$log2FoldChange < 0, ]

cat("Male-Only DEG Results 0.1 (DESeq2)\n")
cat("Total significant DEGs:", nrow(sig_male_deseq), "\n")
cat("Upregulated in Heat:", nrow(sig_male_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(sig_male_deseq_down), "\n\n")

stringent_sig_male_deseq      <- DEG_male_deseq[DEG_male_deseq$padj < 0.05, ]
stringent_sig_male_deseq_up   <- stringent_sig_male_deseq[stringent_sig_male_deseq$log2FoldChange > 0, ]
stringent_sig_male_deseq_down <- stringent_sig_male_deseq[stringent_sig_male_deseq$log2FoldChange < 0, ]

cat("Male-Only DEG Results 0.05 (DESeq2)\n")
cat("Total significant DEGs:", nrow(stringent_sig_male_deseq), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_male_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_male_deseq_down), "\n\n")


## Females only

# Build DESeq2 object
dds_female <- DESeqDataSetFromMatrix(
  countData = Counts_AB_Female,
  colData   = AHDB1_AB_Female,
  design    = ~ 0 + treatment
)

keep_female_deseq <- rowSums(counts(dds_female)) >= 10
dds_female <- dds_female[keep_female_deseq, ]
cat("Genes after filtering:", nrow(dds_female), "\n\n")

dds_female <- DESeq(dds_female)

res_female <- results(
  dds_female,
  contrast = c("treatment", "Heat", "Control"),
  alpha    = 0.05
)

summary(res_female)

DEG_female_deseq <- as.data.frame(res_female)
DEG_female_deseq$Gene_ID <- rownames(DEG_female_deseq)

# Remove NA's
DEG_female_deseq <- DEG_female_deseq[
  !is.na(DEG_female_deseq$padj) & !is.na(DEG_female_deseq$log2FoldChange),
]
cat("Genes after NA removal:", nrow(DEG_female_deseq), "\n\n")

# Subset DEG's into upregulated and downregulated
sig_female_deseq      <- DEG_female_deseq[DEG_female_deseq$padj < 0.1, ]
sig_female_deseq_up   <- sig_female_deseq[sig_female_deseq$log2FoldChange > 0, ]
sig_female_deseq_down <- sig_female_deseq[sig_female_deseq$log2FoldChange < 0, ]

cat("Female-Only DEG Results 0.1 (DESeq2)\n")
cat("Total significant DEGs:", nrow(sig_female_deseq), "\n")
cat("Upregulated in Heat:", nrow(sig_female_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(sig_female_deseq_down), "\n\n")

stringent_sig_female_deseq      <- DEG_female_deseq[DEG_female_deseq$padj < 0.05, ]
stringent_sig_female_deseq_up   <- stringent_sig_female_deseq[stringent_sig_female_deseq$log2FoldChange > 0, ]
stringent_sig_female_deseq_down <- stringent_sig_female_deseq[stringent_sig_female_deseq$log2FoldChange < 0, ]

cat("Female-Only DEG Results 0.05 (DESeq2)\n")
cat("Total significant DEGs:", nrow(stringent_sig_female_deseq), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_female_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_female_deseq_down), "\n\n")

```

# Volcano Plots for all three methods

```{r}
library(patchwork)
library(ggVennDiagram)

```

