---
title: "6_DGE_EdgeRxVoom"
author: "Dasia Simpson, modified by Brynleigh Payne"
date: "2026-06-17"
output: html_document
---
#rmarkdown::render("DGanalyses.R")
#rmarkdown::render("DGanalyses.R", "pdf_document")

# Created by Dasia Simpson in the T. Schwartz Lab for Male Sceloporus Multi Stress Experiment

# Literature mostly R code is from 
# Law et al. (2018). RNA-seq analysis is easy as 1-2-3 with limma, Glimma and edgeR. F1000Research, 5, 1408. https://doi.org/10.12688/f1000research.9005.3

# Literature of which the gene count data was generated from (Bioinformatic Pipeline)
# Pertea et al (2016). Transcript-level expression analysis of RNA-seq experiments with HISAT, StringTie and Ballgown. Nature Protocols, 11(9), 1650–1667. https://doi.org/10.1038/nprot.2016.095

# Literatrue for general RNAseq terminolgy 
#https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7873980/

# Introduction of Differential Gene Expression

#Differential gene expression (DE) refers to the analyses of genes being expressed within a transcriptome. 
#A transcriptome is the collection of RNAs (or mRNA) in a cell or a collection of cells. 
#DE analyses are important because it provides perspective on what genes are being expressed under an experimental condition. 
#For example, what genes are be expressed when an individual is being heat stressed relative to the control (exposure to no stressor)? 
#In this code, we are looking at normalized read count data generated from the HiSAT2 protocol (Pertea et al 2016). 
#Data was taken from a multi stress experiment where male eastern fence lizards were exposed to ecologically relevant stressors, 
#acute heat stress and fire ant envenomation (including a control group). In this code, we will use statistical analysis
#to determine changes in expression levels between experimental groups. 


# This installs Bioconductor packages if havent done so already
# Bioconductor is a free, open source software for the analysis of genomic data and is primarly used in R

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("edgeR")
BiocManager::install("Glimma")
```

#BiocManager::install("Biostrings", version = "3.8")
#BiocManager::install("biomaRt", version = "3.8")
#install.packages("biomartr", dependencies = TRUE)
#BiocManager::install("edgeR")

###########################
## Load  required packages
#edgeR an R packages used for differential expression analyses of read counts from RNA-Seq data 
#edgeR is used to import,organize,filter and normalize the data 
```{r}
library(edgeR) 
```
###########################

###########################
#Limma an R package that uses "voom" method, linear modeling, Bayesian modeling. 
# This package is used to assess differential gene expression
```{r}
library(limma)
```
###########################


###########################
#Glimma is an R package that creates a interactive plot of your results 
```{r}
library(Glimma)
```
###########################


###########################
############################
# Read in data files 
# "gene_count_matrix_anno.txt (converted it into a text file on personal computer) was generated from
#  using a python script, "PrepDE.py", directly from files generated by StringTie to make files readable in Edge R
#  this text file includes  gene level counts from each sample
# more detailed methodology in this literature(Pertea et al., 2016)

# "read.table" reads a file in table format and creates a data frame from it
# "row.names" indicates that the rows in this text file are "Genes_ID"
# "header = TRUE" indicating whether the file contains the names of the variables as its first line

# creates a data from "gene_count_matrix_anno.txt" file and places it into an object called "countdata"
```{r}
countdata <- as.matrix(read.csv("gene_count_matrix (3).csv", row.names="gene_id"))

coldata <-(read.table("PHENO_DATA_Blood_Control_Heat.txt", header=TRUE, row.names=1, sep = "\t"))
dim(coldata)
head(coldata)

## Remove Baseline samples
coldata_filtered <- coldata[!grepl("^Baseline", coldata$treatment), ]
head(coldata_filtered)

## Subset the countdata to only include Control (A) and Heat (B) samples 
samples_to_keep <- rownames(coldata_filtered)

length(intersect(samples_to_keep, colnames(countdata)))

all(samples_to_keep %in% colnames(countdata))

setdiff(samples_to_keep, colnames(countdata))

setdiff(colnames(countdata), samples_to_keep)

common_samples <- intersect(samples_to_keep, colnames(countdata))

countdata_subset <- countdata[, common_samples]
coldata_subset <- coldata_filtered[common_samples, ]

coldata_subset <- coldata_subset[colnames(countdata_subset), ]

identical(rownames(coldata_subset), colnames(countdata_subset))
```

###########################
###########################
# "dim" function tells you how many rows and columns
```{r}
dim(countdata_subset)
```

# "head" previews the first few lines of your data
```{r}
head(countdata_subset)
```

# Make a DGEList object from the data table (countdata_subset) then puts DGEList object into a variable "x"
```{r}
x <- DGEList(counts=countdata_subset, samples=coldata_subset)
class(x)
```

###########################

# Defining Treatment Groups
# Organizing sample information
# DEGList function because all the samples are stored into 1 file 
# This line of code associates the treatments (FA-Fire Ant, HS-HEat Stress and C-Control) with the columns of the count matrix 
```{r}
coldata_subset$sex <- factor(coldata_subset$sex)

coldata_subset$treatment <- factor(coldata_subset$treatment)

```

###########################
#Library Size and Normalization Factors 

#Calculates library size (the total number of reads for that individial) for each sample and sets normalization factors to 1
#all samples
```{r}
x$samples

x
```

## Generate barplot of sample library size for each treatment/control
```{r}
barplot(x$samples$lib.size*1e-6, ylab="Library size (millions)")
```

#write.csv (x$samples, file="sampleslibrarysize.csv")

## Add sample names to the dataframe
```{r}
samplenames <- colnames(countdata_subset)
samplenames
```


#Raw counts are converted to CPM (counts per million) and log-CPM (log2 -counts per million)
# CPM and log-CPM transformations are used regularly in RNAseq analyses, however they do not account for  gene length differences

# "cpm" function = CPM (counts per million)
# "lcpm function = log-CPM (log2 -counts per million)
## for all treatments 
```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
summary(lcpm)
```

#average library size is 84.33366 
```{r}
L <- mean(x$samples$lib.size) * 1e-6
L
```

#median library size is 83.94554
```{r}
M <- median(x$samples$lib.size) * 1e-6
M
```


# Filtering Low-Expression Genes in R

# Shows that ? genes across all 49 samples does NOT zero counts 
FALSE  TRUE 
21372  2139 
```{r}
table(rowSums(x$counts==0)==49)
```


#Filtering genes that are lowly expressed
#The cut-off is cpm is >1 expressed across 3 conditions were kept
# 12489 genes were passed this filter
```{r}
keep.exprs <- rowSums(cpm > 1) >= 3
x <- x[keep.exprs, keep.lib.sizes = FALSE]
x
```

#write.csv (x$samples, file="sampleslibrarysize_filtered.csv")

# gives a number of columns and rows, should be reduced number because we filtered through there count data in the previous lines 
```{r}
dim(x)
```


## Visual represention of logCPM values for raw data count files and filterd data
## count files. Using this pipeline, number of genes reduced to 54,149 to 16,877 which is 
## 68.83% total loss 
```{r}
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.21), las=2,
     main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=0, lty=3)
for (i in 2:nsamples){
  den <- density(lcpm[,i])
  lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.21), las=2,
     main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=0, lty=3)
for (i in 2:nsamples){
  den <- density(lcpm[,i])
  lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```
###################



######Normalization (EdgeR uses TTM normalization)
## Correction factors computed internally 
```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
```

##### Gives a visual representation of Normalization 
```{r}
x2 <- x
x2$samples$norm.factors <- 1
x2$counts[,1] <- ceiling(x2$counts[,1]*0.05)
x2$counts[,2] <- x2$counts[,2]*5


lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```

##################
### Can be used for a Heatmap
# Normalized count data read into a csv file to make heatmap
```{r}
write.csv(lcpm,file = "NormalisedCount_data.csv" )
```

#########################
########################


## Create Multi-dimensional Scaling (MDS) plot
## This plots shows the similarites and differences between treatment groups
## # the colour grouping defined by cell types.
# Not currently separating out the sexes
```{r}
library(RColorBrewer)

group <- factor(x$samples$treatment)

cols <- brewer.pal(3, "Set1")[1:nlevels(group)]

plotMDS(lcpm,
        labels=group,
        col=cols[as.numeric(group)])

```

########################
########################

#Quick peek at data
```{r}
x$samples
x$counts
```

# Glimma makes interactive web plot based on log counts per million (lcpm)

```{r}
glMDSPlot(lcpm, labels=paste(x$samples$treatment), treatment=x$samples[,c(2)],launch=TRUE)
```

###	Statistical Model to test for DE genes
## Create a design to compare by group factors

```{r}
# Interaction
interaction_design <- model.matrix(~ 0 + sex + treatment + sex:treatment, data = coldata_subset)

# Female Model
# Subset females from pheno data
coldata_female_subset <- coldata_subset[coldata_subset$sex == "female", ]
head(coldata_female_subset)

# Subset counts to match
keep_samples <- rownames(coldata_female_subset)

countdata_female_subset <- countdata_subset[, keep_samples]

# Sanity check
coldata_female_subset <- coldata_female_subset[colnames(countdata_female_subset), ]

identical(rownames(coldata_female_subset), colnames(countdata_female_subset))

# Create design for females
x.female <- x[, rownames(coldata_female_subset)]
female_design <- model.matrix(
    ~0 + treatment,
    data = coldata_female_subset
)

colnames(female_design) <- c("Control", "Heat")

female_design

# Male Model
# Subset males from pheno data
coldata_male_subset <- coldata_subset[coldata_subset$sex == "male", ]
head(coldata_male_subset)

# Subset counts to match
keep_samples <- rownames(coldata_male_subset)

countdata_male_subset <- countdata_subset[, keep_samples]

# Sanity check
coldata_male_subset <- coldata_male_subset[colnames(countdata_male_subset), ]

identical(rownames(coldata_male_subset), colnames(countdata_male_subset))

# Create design for males
x.male <- x[, rownames(coldata_male_subset)]
male_design <- model.matrix(
    ~0 + treatment,
    data = coldata_male_subset
)

colnames(male_design) <- c("Control", "Heat")

male_design
```

##############
## the left side is the name of the contrast, the right side of the "=" are the two Treatments you want to contrast 
#(make sure the names of the treatments match what you added earlier)
#contr.matrix <- makeContrasts(
#  Heat = C-HS, 
#  FireAnt = C-FA, 
#  FireAnt_vs_Heat = FA-HS, 
# levels=design)
#contr.matrix

#(make sure the names of the treatments match what you added earlier)
contr.matrix <- makeContrasts(
  Heat = HS-C, 
  FireAnt = FA-C, 
  FireAnt_vs_Heat = FA-HS, 
  levels=design)
contr.matrix
```{r}
contr.matrix <- makeContrasts(
    Heat_vs_Control = Heat - Control,
    levels = female_design
)

contr.matrix2 <- makeContrasts(
    Heat_vs_Control = Heat - Control,
    levels = male_design
)
```

############Removing heteroscedascity from count data
#In limma, linear modelling is carried out on the log-CPM values which are assumed to be normally distributed and the mean-variance relationship is accommodated using precision weights calculated by the voom function.
#voom converts raw counts to log-CPM values by automatically extracting library sizes and normalisation factors from x itself. Additional normalisation to log-CPM values can be specified within voom using the normalize.method argument.
```{r}
v.female <- voom(x.female, female_design, plot=TRUE)
v.female

v.male <- voom(x.male, male_design, plot=TRUE)
v.male
```


######### Linear Modeling using Limma #########
## number of differentially expressed genes  
## Significance is defined using an adjusted p-value cutoff that is set at 5% by default.
## "lmFit" Fit linear model for each gene given a series of arrays
## "contrasts.fit" compute and test arbitrary contrasts for regression objects (in this case "vfit")
## "eBayes" Empirical Bayes Statistics
```{r}
vfit.female <- lmFit(v.female, female_design)
vfit.female <- contrasts.fit(vfit.female, contrasts=contr.matrix)
efit.female <- eBayes(vfit.female)
plotSA(efit.female)

vfit.male <- lmFit(v.male, male_design)
vfit.male <- contrasts.fit(vfit.male, contrasts=contr.matrix2)
efit.male <- eBayes(vfit.male)
plotSA(efit.male)
```


# DE genes can be extracted using the results from the "decideTests" function
## This function gives you the opportunity look at DE genes summarized in a table
```{r}
et.female <-(decideTests(efit.female))
summary(et.female)

et.male <-(decideTests(efit.male))
summary(et.male)
```


#  For a stricter  significance, you can change the logFC value to be above a minimum value
# "treat" function used to calculate p-values from empirical Bayes moderated t-statistics with a minimum log-FC requirement (Law et al. 2018)
```{r}
tfit.female <- treat(vfit.female, lfc=1)
dt.female <- decideTests(tfit.female)
summary(dt.female)

tfit.male <- treat(vfit.male, lfc=1)
dt.male <- decideTests(tfit.male)
summary(dt.male)
```

####  Genes that are in common among the contrasts - use the columns in the contrast matrix
## if use contrasts above, 1 would be control vs heat, 2 would be control vs fire ant, 3 would be heat vs fire ant
# Only do if more than 1 contracts, ex. interaction model
```{r}
de.common.female <- which(et.female[,1]!=0 & et.female[,2]!=0)
length(de.common.female)
```

# Gives list of the commonly expressed DE genes
head(de.common)


# venn diagram 
vennDiagram(et[,1:2], circle.col=c("turquoise", "salmon"))


# "write.fit" function can pull the results each genes from all experimental condtions and output it into a single file (for example p-values)
write.fit(efit, et, file="results.txt")


# Give you a table of the DE genes
# By default topTreat arranges genes from smallest to largest adjusted p-value

# this table gives you all DE genes both statiscally signifcant and not significant !!!
# heat and control 
#1 would be control vs heat
```{r}
female_TopDEgenes <- topTable(efit.female, coef = "Heat_vs_Control", n = Inf, sort.by = "P")
summary(female_TopDEgenes)
head(female_TopDEgenes)

male_TopDEgenes <- topTable(efit.male, coef = "Heat_vs_Control", n = Inf, sort.by = "P")
summary(male_TopDEgenes)
head(male_TopDEgenes)
```

## creates a .csv file 
```{r}
write.csv(female_TopDEgenes,file = "Female_EdgeRxLimmaVoom_TopDE_genes_Heat_vs_Control.csv" )

write.csv(male_TopDEgenes,file = "Male_EdgeRxLimmaVoom_TopDE_genes_Heat_vs_Control.csv" )
```

#fireant and control
# this table gives you all DE genes both statisTcally signifcant and not significant 
#2 would be control vs fire ant
fireant_TopDEgenes<- topTreat(efit, coef=2, n=Inf)
summary(fireant_TopDEgenes)
head(fireant_TopDEgenes)

## creates a .csv file
write.csv(fireant_TopDEgenes,file = "TopDE_genes_Fireant_vs_Control.csv" )

#fire ant and heat 
FAvsHS_TopDEgenes<- topTreat(efit, coef=3, n=Inf)
summary(FAvsHS_TopDEgenes)
head(FAvsHS_TopDEgenes)

write.csv(FAvsHS_TopDEgenes,file = "Top_FAvsHS_TopDEgenes.csv" )

# THIS LINE OF COMMAND GIEVES YOU ALL "statistically significant" genes
# Gives list of all differentially expressed genes with a cutoff value of 0.05
deg <- topTreat(efit, n = Inf, p = 0.05)
write.csv(deg, file = "deg_ALLTREATMENTS.csv")

#give list of DE genes in heat stress relative to the control with a cutoff value of 0.05
degheat <-topTreat(efit,coef = 1,n = Inf,p = 0.05)
write.csv(degheat, file = "deg_HEATONLY.csv")

#give list of DE genes in fire ant relative to the control with a cutoff value of 0.05
degfire <-topTreat(efit,coef = 2,n = Inf,p = 0.05)


#Gives list of upregulated genes using "deg" object
up <- row.names(deg[deg$logFC > 0,])
up
write.csv(up, file = "up_ALLtreatments.csv")

# Gives a list of downregulated genes using "deg" object
down <- row.names(deg[deg$logFC < 0,])
down
write.csv(down, file = "down_ALLtreatments.csv")

# gives list of upregulated genes using "degheat" object (heat stress DE genes)
upregulated_heat <- row.names(degheat[degheat$logFC > 0,])
upregulated_heat
write.csv(upregulated_heat, file = "up_heatONLY.csv")

# gives list of downregulated genes using "degheat" object (heat stress DE genes) 
downregulated_heat <- row.names(degheat[degheat$logFC < 0,])
downregulated_heat
write.csv(downregulated_heat, file = "down_heatONLY.csv")


# gives list of upregulated genes using "degFIRE" object (FIRE ANT DE genes)

upregulated_fire <- row.names(degfire[degfire$logFC > 0,])
upregulated_fire
write.csv(upregulated_fire, file = "up_fireONLY.csv")

# gives list of downregulated genes using "degFIRE" object (FIRE ANT DE genes) 


downregulated_fire <- row.names(degfire[degfire$logFC < 0,])
downregulated_fire
write.csv(downregulated_fire, file = "down_fireONLY.csv")

# MD plots based off DE genes

#Heat
plotMD(efit, column=1, status=et[,1], main=colnames(efit)[1], xlim=c(-8,13))

# Fire Ant
plotMD(efit, column=2, status=et[,2], main=colnames(efit)[2], xlim=c(-8,13))

#########################################################################################
###### CORT DATA ######
#########################################################################################


## Need to set working directory 

library(ggplot2)
library(dplyr)

Data <- read.csv(file = "CORT_HeatSeries_Data_DYS.csv", header = TRUE, stringsAsFactors = FALSE)
str(Data)



MultistressCORT <- Data[Data$Experiment == "MultiStress",]
MultiStressBLEED <- read.csv(file = "MultiStress_Males_ExpDayData_26-27July2017_FINAL.csv", header = TRUE, stringsAsFactors = FALSE)
Capture <- read.csv(file = "Sceloporus_CollectionData_2017.csv", header = TRUE, stringsAsFactors = FALSE)

MultiStress <- merge(MultiStressBLEED, MultistressCORT, by = "ID")
MultiStress <- merge(MultiStress, Capture[,c(1,3,14,15)], by = "ID")
MultiStress$Condition <- resid(lm(MultiStress$Mass_Capture_g~MultiStress$SVL_Capture_mm))

# preview data
head(MultiStress)

# list the column names using "colnames" function
colnames(MultiStress)



# filtering dataset to only include FA, HS, Control groups
MultiStress_NOCORT <- filter(MultiStress, Treatment == "FA"|Treatment ==  "HS" |Treatment == "CONTROL")

head(MultiStress_NOCORT)

##Barplot
ggplot(data = MultiStress_NOCORT,aes (x= Treatment, y = CorrectedCORT)) +
  geom_boxplot() +
  theme(axis.text=element_text(size=14),axis.title=element_text(size=16,face="bold")) +
  stat_summary(fun.data = mean_cl_boot) +
  geom_point(size = 2, position = position_jitter(w=.02), shape = 21, alpha = 0.8, col = "black", fill = "black") +
  ylab ("Plasma CORT(ng/ml)")


ggplot(data = MultiStress, aes (x= Treatment, y = CorrectedCORT)) +
  stat_summary(fun.data = mean_cl_boot, geom = "pointrange", size = 2) + 
  geom_point(size = 2, position = position_jitter(w=.02), shape = 21, alpha = 0.8, col = "black", fill = "white") +
  ylab("Plasma CORT (ng/ml)") + 
  rory_theme

ggplot(data = MultiStress, aes (x= Treatment, y = CorrectedCORT)) +
  stat_summary(fun.data = mean_cl_boot, geom = "pointrange", size = 2) + 
  geom_point(size = 2, position = position_jitter(w=.02), shape = 21, alpha = 0.8, col = "black", fill = "white") +
  ylab("Plasma CORT (ng/ml)") + 
  rory_theme


MultiStress_M1 <- lm(CorrectedCORT ~ Condition * Treatment, data = MultiStress)
MultiStress_M2 <- lm(CorrectedCORT ~ Condition + Treatment, data = MultiStress)
MultiStress_M3 <- lm(CorrectedCORT ~ Treatment, data = MultiStress)
aictab(list(MultiStress_M1, MultiStress_M2, MultiStress_M3), 
       modnames = c("MultiStress_M1", "MultiStress_M2", "MultiStress_M3")) #M3 is preferred
Anova(MultiStress_M3, type = 3) #quite significant P < 0.0001
lsmeans(MultiStress_M3, pairwise~Treatment) #confirms Control and CORT are the same, and below FA and HS which are indistinguishable
