---
title: "ZebraFinch Acute Heat Damage Biomarker Experiment 1: Liver RNAseq (Heat vs Control)"
author: "Ali Amer, modified by Brynleigh"
output:
  html_document:
    df_print: paged

---

# Script Goal: 
# 1. Convert raw count reads of samples sent for blood RNA-Seq (A vs B) from prep.dy script   
# 2. Check library sizes across A vs B & Normalize raw count data
# 3. Run DGE with edgeR, limma and DESeq2 (Males Only & Females Only)
# 4. Run Pathway Enrichment & WGCNA

```{r setup, include=FALSE}
# Check working directory 
getwd()

```

```{r}

# Read in Acute Heat Damage Biomarker Experiment 1 MasterSpreadSheet
MasterSpreadsheet <- read.csv(
  "~/Downloads/ZebraFinch_DamageFitnessProject/LiverRNAseq/AHDB_Exp1_MasterSpreadSheet.csv",
  header = TRUE,
  stringsAsFactors = FALSE
)
str(MasterSpreadsheet)

# Read in Sample List of Samples Sent for RNA Seq
ZF_Blood_sample_list <- read.csv(
  "ZF_Blood_sample_list.csv",
  header = TRUE,
  stringsAsFactors = FALSE
)
str(ZF_Blood_sample_list)
dim(ZF_Blood_sample_list)

# Read in Raw Count Data
Raw_Counts <- read.csv(
  "gene_count_matrix (3).csv",
  row.names = 1,
  header = TRUE,
  check.names = FALSE  
)

head(colnames(Raw_Counts))
dim(Raw_Counts)

# Store initial gene count
initial_gene_count <- nrow(Raw_Counts)

cat("Initial raw gene count:", initial_gene_count, "\n")
cat("Number of samples:", ncol(Raw_Counts), "\n\n")

```

# Subset & Align AHDB Exp1 Master File with RNASeq Sample List 

```{r}

# Extract numeric band ID from Individual_ID
ZF_Blood_sample_list$ID_Band <- as.numeric(
  sub("^ZF_", "", ZF_Blood_sample_list$Sample_ID)
)
head(ZF_Blood_sample_list)

# Subset data from master file
Master_Spreadsheet_RNAseq_subset <- MasterSpreadsheet[
  MasterSpreadsheet$ID_Band %in% ZF_Blood_sample_list$ID_Band,
]

# Remove Baseline Timepoint Columns
Master_Spreadsheet_RNAseq_subset <- subset(
  Master_Spreadsheet_RNAseq_subset,
  TimePoint != "Baseline")

# Remove Treatment Timepoint for E Treatment Columns
Master_Spreadsheet_RNAseq_subset <- subset(
  Master_Spreadsheet_RNAseq_subset,
  !(Treatment == "Heat_2-WK_Repair" & TimePoint == "Treatment")
)

# Final check
all(Master_Spreadsheet_RNAseq_subset$ID_Band %in% ZF_Blood_sample_list$ID_Band)
# All ID_Bands match

# Re-write treatment column
Master_Spreadsheet_RNAseq_subset$Treatment <- dplyr::recode(
  Master_Spreadsheet_RNAseq_subset$Treatment,
  "A" = "Control",
  "B" = "Heat",
  "C" = "Heat_2-HR_Repair",
  "D" = "Heat_24-HR_Repair",
  "E" = "Heat_2-WK_Repair"
)

# Confirm correct treatment distribution - 70 total (20 Individuals in A & B each, 10 Indviduals in C through E) 
cat("\n TREATMENT GROUP DISTRIBUTION \n")
print(table(Master_Spreadsheet_RNAseq_subset$Treatment))
head(Master_Spreadsheet_RNAseq_subset) 

```

# Merge Raw Count Data with MasterSpreadsheet that was aligned with RNA-Seq Sample ID List

```{r}

# Merge all RNAseq samples with metadata
Master_Spreadsheet_Full <- merge(
  Master_Spreadsheet_RNAseq_subset,
  ZF_Blood_sample_list[, c("ID_Band", "RNAseq_Sample_ID")],
  by = "ID_Band",
  all.x = TRUE
)

rownames(Master_Spreadsheet_Full) <- Master_Spreadsheet_Full$RNAseq_Sample_ID

# Add Count Data for all samples
Counts_Full <- Raw_Counts[, rownames(Master_Spreadsheet_Full)]

# Verify alignment
cat("Counts columns match metadata rows:", 
    all(colnames(Counts_Full) == rownames(Master_Spreadsheet_Full)), "\n")

```

# Subset Data for A (Control) vs B (Heat) Comparison

```{r}

# Subset experimental data for Control vs Heat
AHDB1_AB <- subset(
  Master_Spreadsheet_RNAseq_subset,
  Treatment %in% c("Control", "^Heat")
)
cat("Subset dimensions:", nrow(AHDB1_AB), "samples\n")

AHDB1_AB <- merge(
  AHDB1_AB,
  ZF_Blood_sample_list[, c("ID_Band", "RNAseq_Sample_ID")],
  by = "ID_Band",
  all.x = TRUE
)

rownames(AHDB1_AB) <- AHDB1_AB$RNAseq_Sample_ID

# Subset Count Data
Counts_AB <- Raw_Counts[, rownames(AHDB1_AB)]

# Construct DGEList object
dge_AB <- DGEList(
  counts  = Counts_AB,
  samples = AHDB1_AB
)

# Sanity check
cat("\nTreatment by Sex distribution:\n")
print(table(dge_AB$samples$Treatment, dge_AB$samples$Sex))
cat("\nDGEList object created:", class(dge_AB), "\n")
cat("Samples:", ncol(dge_AB), "\n")
cat("Genes (before filtering):", nrow(dge_AB), "\n\n")

```

# Calculate Library Size & Make BoxPlots (A vs B)

```{r}

cpm_AB <- cpm(dge_AB)
lcpm_AB <- cpm(dge_AB, log=TRUE)
summary(lcpm_AB)

# Boxplot by Treatment
boxplot(
  dge_AB$samples$lib.size/1e6 ~ dge_AB$samples$Treatment,
  ylab = "Library size (millions)",
  main = "Library Size by Treatment (Control vs Heat)",
  col = c("lightblue", "coral"),
  outline = FALSE
)

# Boxplot by Treatment & Sex
boxplot(
  dge_AB$samples$lib.size/1e6 ~ interaction(dge_AB$samples$Treatment, dge_AB$samples$Sex),
  ylab = "Library size (millions)",
  main = "Library Size by Treatment and Sex (Control vs Heat)",
  las = 3,
  col = rep(c("lightblue", "coral"), 2),
  outline = FALSE
)

# Display
par(mfrow = c(1, 2))
boxplot(
  dge_AB$samples$lib.size/1e6 ~ dge_AB$samples$Treatment,
  ylab = "Library size (millions)",
  main = "Library Size by Treatment",
  col = c("lightblue", "coral"),
  outline = FALSE
)
boxplot(
  dge_AB$samples$lib.size/1e6 ~ interaction(dge_AB$samples$Treatment, dge_AB$samples$Sex),
  ylab = "Library size (millions)",
  main = "Library Size by Treatment and Sex",
  xlab = "",
  las = 3,
  col = rep(c("lightblue", "coral"), 2),
  outline = FALSE
)

```

# Differential Gene Expression w/ EdgeR, Limma, DESeq2 (Male ZEFI's Only)

```{r}

library(ggplot2)

### DGE with Edge R ###

library(edgeR)

AHDB1_AB_Male <- subset(
  AHDB1_AB,
  Sex == "M"
)

dim(AHDB1_AB_Male)

Counts_AB_Male <- Raw_Counts[, rownames(AHDB1_AB_Male)]

dge_AB_Male <- DGEList(
  counts  = Counts_AB_Male,
  samples = AHDB1_AB_Male
)

# Set factor and build design matrix
dge_AB_Male$samples$Treatment <- droplevels(factor(dge_AB_Male$samples$Treatment))

design_male <- model.matrix(~ 0 + Treatment, data = dge_AB_Male$samples)
colnames(design_male) <- levels(dge_AB_Male$samples$Treatment)
table(dge_AB_Male$samples$Treatment)

cat("Total genes (before filtering):", nrow(dge_AB_Male), "\n\n")

# filter using the design matrix
keep_male <- filterByExpr(dge_AB_Male, design_male)
dge_AB_Male <- dge_AB_Male[keep_male, , keep.lib.sizes = FALSE]
cat("Total genes (after filtering):", nrow(dge_AB_Male), "\n\n")

# TMM normalization
dge_AB_Male <- calcNormFactors(dge_AB_Male, method = "TMM")

dge_AB_Male <- estimateDisp(dge_AB_Male, design_male)
fit_male <- glmQLFit(dge_AB_Male, design_male)

contrast_male <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_male
)
cat("Contrast")
print(contrast_male)

qlf_male <- glmQLFTest(fit_male, contrast = contrast_male)

DEG_male <- topTags(qlf_male, n = Inf)$table
DEG_male$Gene_ID <- rownames(DEG_male)

sig_male <- DEG_male[DEG_male$FDR < 0.05, ]
sig_male_up <- sig_male[sig_male$logFC > 0, ]
sig_male_down <- sig_male[sig_male$logFC < 0, ]

cat("Male-Only DEG Results - EdgeR")
cat("Total significant DEGs:", nrow(sig_male), "\n")
cat("Upregulated in Heat:", nrow(sig_male_up), "\n")
cat("Downregulated in Heat:", nrow(sig_male_down), "\n\n")

# This needs to be fixed....

#write.csv(DEG_male, "Results/DGE/Male_Only/DEG_Male_Heat_vs_Control_All.csv", 
#          row.names = FALSE)
#write.csv(sig_male, "Results/DGE/Male_Only/DEG_Male_Heat_vs_Control_Significant.csv", 
#          row.names = FALSE)
#write.csv(sig_male_up, "Results/DGE/Male_Only/DEG_Male_Upregulated.csv", 
#          row.names = FALSE)
#write.csv(sig_male_down, "Results/DGE/Male_Only/DEG_Male_Downregulated.csv", 
#          row.names = FALSE)

### DGE with limma-voom ###

library(limma)

dge_AB_Male_voom <- DGEList(
  counts  = Counts_AB_Male,
  samples = AHDB1_AB_Male
)

# Sanity Check
table(dge_AB_Male_voom$samples$Treatment)
cat("Male-Only Subset\n")
cat("Total genes (before filtering):", nrow(dge_AB_Male_voom), "\n\n")

dge_AB_Male_voom$samples$Treatment <- droplevels(factor(dge_AB_Male_voom$samples$Treatment))

design_male_voom <- model.matrix(~ 0 + Treatment, data = dge_AB_Male_voom$samples)
colnames(design_male_voom) <- levels(dge_AB_Male_voom$samples$Treatment)

keep_male_voom <- filterByExpr(dge_AB_Male_voom, design_male_voom)
dge_AB_Male_voom <- dge_AB_Male_voom[keep_male_voom, , keep.lib.sizes = FALSE]

cat("Genes after filtering:", nrow(dge_AB_Male_voom), "\n\n")

dge_AB_Male_voom <- calcNormFactors(dge_AB_Male_voom, method = "TMM")

v_male <- voom(dge_AB_Male_voom, design_male_voom, plot = TRUE)

fit_male_voom <- lmFit(v_male, design_male_voom)

contrast_male_voom <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_male_voom
)
cat("Contrast:\n")
print(contrast_male_voom)

fit_male_voom <- contrasts.fit(fit_male_voom, contrast_male_voom)
fit_male_voom <- eBayes(fit_male_voom)

DEG_male_voom <- topTable(fit_male_voom, coef = "Heat_vs_Control", n = Inf, sort.by = "P")
DEG_male_voom$Gene_ID <- rownames(DEG_male_voom)

sig_male_voom <- DEG_male_voom[DEG_male_voom$adj.P.Val < 0.05, ]
sig_male_voom_up <- sig_male_voom[sig_male_voom$logFC > 0, ]
sig_male_voom_down <- sig_male_voom[sig_male_voom$logFC < 0, ]

cat("Male-Only DEG Results (limma-voom)\n")
cat("Total significant DEGs:", nrow(sig_male_voom), "\n")
cat("Upregulated in Heat:", nrow(sig_male_voom_up), "\n")
cat("Downregulated in Heat:", nrow(sig_male_voom_down), "\n\n")

# Need to fix (i.e,. make correct directories)
#write.csv(DEG_male_voom, "Results/DGE/Male_Only/Voom_DEG_Male_Heat_vs_Control_All.csv",
#          row.names = FALSE)
#write.csv(sig_male_voom,"Results/DGE/Male_Only/Voom_DEG_Male_Heat_vs_Control_Significant.csv",
#          row.names = FALSE)
#write.csv(sig_male_voom_up,"Results/DGE/Male_Only/Voom_DEG_Male_Upregulated.csv",
#          row.names = FALSE)
#write.csv(sig_male_voom_down, #"Results/DGE/Male_Only/Voom_DEG_Male_Downregulated.csv",
#          row.names = FALSE)

# Plot mean variance trend
v_male <- voom(dge_AB_Male_voom, design_male_voom, plot = TRUE)

### DGE with DESeq2 ###

library(DESeq2)

AHDB1_AB_Male_deseq <- subset(AHDB1_AB, Sex == "M")
Counts_AB_Male_deseq <- Raw_Counts[, rownames(AHDB1_AB_Male_deseq)]

Counts_AB_Male_int <- round(Counts_AB_Male_deseq)

# Build colData and set reference level
col_data_male <- AHDB1_AB_Male_deseq
col_data_male$Treatment <- droplevels(factor(col_data_male$Treatment))
col_data_male$Treatment <- relevel(col_data_male$Treatment, ref = "Control")

cat("Male-Only Subset\n")
cat("Total genes (before filtering):", nrow(Counts_AB_Male_int), "\n\n")

# Build DESeq2 object
dds_male <- DESeqDataSetFromMatrix(
  countData = Counts_AB_Male_int,
  colData   = col_data_male,
  design    = ~ Treatment
)

keep_male_deseq <- rowSums(counts(dds_male)) >= 10
dds_male <- dds_male[keep_male_deseq, ]
cat("Genes after filtering:", nrow(dds_male), "\n\n")

dds_male <- DESeq(dds_male)

res_male <- results(
  dds_male,
  contrast = c("Treatment", "Heat", "Control"),
  alpha    = 0.05
)

summary(res_male)

DEG_male_deseq <- as.data.frame(res_male)
DEG_male_deseq$Gene_ID <- rownames(DEG_male_deseq)

# Remove NA's
DEG_male_deseq <- DEG_male_deseq[
  !is.na(DEG_male_deseq$padj) & !is.na(DEG_male_deseq$log2FoldChange),
]
cat("Genes after NA removal:", nrow(DEG_male_deseq), "\n\n")

# Subset DEG"s into upregulated and downregulated 

sig_male_deseq      <- DEG_male_deseq[DEG_male_deseq$padj < 0.05, ]
sig_male_deseq_up   <- sig_male_deseq[sig_male_deseq$log2FoldChange > 0, ]
sig_male_deseq_down <- sig_male_deseq[sig_male_deseq$log2FoldChange < 0, ]

cat("Male-Only DEG Results (DESeq2)\n")
cat("Total significant DEGs:", nrow(sig_male_deseq), "\n")
cat("Upregulated in Heat:", nrow(sig_male_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(sig_male_deseq_down), "\n\n")

#write.csv(DEG_male_deseq,      #"Results/DGE/Male_Only/DESeq2_DEG_Male_Heat_vs_Control_All.csv",         #row.names = FALSE)
#write.csv(sig_male_deseq,      #"Results/DGE/Male_Only/DESeq2_DEG_Male_Heat_vs_Control_Significant.csv", #row.names = FALSE)
#write.csv(sig_male_deseq_up,   #"Results/DGE/Male_Only/DESeq2_DEG_Male_Upregulated.csv",                 #row.names = FALSE)
#write.csv(sig_male_deseq_down, #"Results/DGE/Male_Only/DESeq2_DEG_Male_Downregulated.csv",               #row.names = FALSE)

# Volcano Plots for all three methods

library(patchwork)
library(ggVennDiagram)

# edgeR volcano 
make_volcano <- function(deg_df, title) {
  deg_df$Significance <- "NS"
  deg_df$Significance[deg_df$FDR < 0.05 & deg_df$logFC > 0] <- "Up in Heat"
  deg_df$Significance[deg_df$FDR < 0.05 & deg_df$logFC < 0] <- "Down in Heat"
  deg_df$Significance <- factor(deg_df$Significance,
                                levels = c("Up in Heat", "Down in Heat", "NS"))
  ggplot(deg_df, aes(x = logFC, y = -log10(FDR), color = Significance)) +
    geom_point(alpha = 0.6, size = 1.2) +
    scale_color_manual(values = c(
      "Up in Heat"   = "#F08080",
      "Down in Heat" = "#6BAED6",
      "NS"           = "#AAAAAA"
    )) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed",
               color = "black", linewidth = 0.5) +
    labs(title = title, x = "log2 Fold Change", y = "-log10(FDR)", color = NULL) +
    theme_bw(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", hjust = 0.5),
      legend.position  = "bottom",
      legend.text      = element_text(size = 11),
      panel.grid.minor = element_blank()
    ) +
    guides(color = guide_legend(override.aes = list(size = 3)))
}

# limma-voom volcano 
make_volcano_voom <- function(deg_df, title) {
  deg_df$Significance <- "NS"
  deg_df$Significance[deg_df$adj.P.Val < 0.05 & deg_df$logFC > 0] <- "Up in Heat"
  deg_df$Significance[deg_df$adj.P.Val < 0.05 & deg_df$logFC < 0] <- "Down in Heat"
  deg_df$Significance <- factor(deg_df$Significance,
                                levels = c("Up in Heat", "Down in Heat", "NS"))
  ggplot(deg_df, aes(x = logFC, y = -log10(adj.P.Val), color = Significance)) +
    geom_point(alpha = 0.6, size = 1.2) +
    scale_color_manual(values = c(
      "Up in Heat"   = "#F08080",
      "Down in Heat" = "#6BAED6",
      "NS"           = "#AAAAAA"
    )) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed",
               color = "black", linewidth = 0.5) +
    labs(title = title, x = "log2 Fold Change", y = "-log10(adj.P.Val)", color = NULL) +
    theme_bw(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", hjust = 0.5),
      legend.position  = "bottom",
      legend.text      = element_text(size = 11),
      panel.grid.minor = element_blank()
    ) +
    guides(color = guide_legend(override.aes = list(size = 3)))
}

# DESeq2 volcano 
make_volcano_deseq <- function(deg_df, title) {
  deg_df$Significance <- "NS"
  deg_df$Significance[deg_df$padj < 0.05 & deg_df$log2FoldChange > 0] <- "Up in Heat"
  deg_df$Significance[deg_df$padj < 0.05 & deg_df$log2FoldChange < 0] <- "Down in Heat"
  deg_df$Significance <- factor(deg_df$Significance,
                                levels = c("Up in Heat", "Down in Heat", "NS"))
  ggplot(deg_df, aes(x = log2FoldChange, y = -log10(padj), color = Significance)) +
    geom_point(alpha = 0.6, size = 1.2) +
    scale_color_manual(values = c(
      "Up in Heat"   = "#F08080",
      "Down in Heat" = "#6BAED6",
      "NS"           = "#AAAAAA"
    )) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed",
               color = "black", linewidth = 0.5) +
    labs(title = title, x = "log2 Fold Change", y = "-log10(padj)", color = NULL) +
    theme_bw(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", hjust = 0.5),
      legend.position  = "bottom",
      legend.text      = element_text(size = 11),
      panel.grid.minor = element_blank()
    ) +
    guides(color = guide_legend(override.aes = list(size = 3)))
}

# Build and combine
p_edger <- make_volcano(DEG_male,             "edgeR")
p_voom  <- make_volcano_voom(DEG_male_voom,   "limma-voom")
p_deseq <- make_volcano_deseq(DEG_male_deseq, "DESeq2")

combined_volcano <- (p_edger | p_voom | p_deseq) +
  plot_annotation(
    title    = "Heat vs Control — Males Only",
    subtitle = "edgeR  |  limma-voom  |  DESeq2",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 15, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray40")
    )
  ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

print(combined_volcano)

ggsave("Results/DGE/Male_Only/Volcano_Male_All_Methods.png",
       combined_volcano, width = 18, height = 6, dpi = 300)

venn_list <- list(
  edgeR        = rownames(sig_male),
  `limma-voom` = rownames(sig_male_voom),
  DESeq2       = sig_male_deseq$Gene_ID
)

venn_up <- list(
  edgeR        = rownames(sig_male_up),
  `limma-voom` = rownames(sig_male_voom_up),
  DESeq2       = sig_male_deseq_up$Gene_ID
)

venn_down <- list(
  edgeR        = rownames(sig_male_down),
  `limma-voom` = rownames(sig_male_voom_down),
  DESeq2       = sig_male_deseq_down$Gene_ID
)

# All DEGs
p_venn <- ggVennDiagram(venn_list, label_alpha = 0, label = "count") +
  scale_fill_gradient(low = "#F7FBFF", high = "#2171B5") +
  scale_color_manual(values = c("black", "black", "black")) +
  labs(title = "All DEGs") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

# Upregulated
p_venn_up <- ggVennDiagram(venn_up, label_alpha = 0, label = "count") +
  scale_fill_gradient(low = "#FFF5F0", high = "#CB181D") +
  scale_color_manual(values = c("black", "black", "black")) +
  labs(title = "Upregulated in Heat") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

# Downregulated
p_venn_down <- ggVennDiagram(venn_down, label_alpha = 0, label = "count") +
  scale_fill_gradient(low = "#F7FBFF", high = "#2171B5") +
  scale_color_manual(values = c("black", "black", "black")) +
  labs(title = "Downregulated in Heat") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

# Combined Venn
combined_venn <- (p_venn / (p_venn_up | p_venn_down)) +
  plot_annotation(
    title = "DEG Overlap Across Methods — Males (Heat vs Control)",
    theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  )

print(combined_venn)

ggsave("Results/DGE/Male_Only/Venn_Male_Combined.png",
       combined_venn, width = 12, height = 12, dpi = 300)

consensus_all  <- Reduce(intersect, venn_list)
consensus_up   <- Reduce(intersect, venn_up)
consensus_down <- Reduce(intersect, venn_down)

cat("Genes significant in all 3 methods:", length(consensus_all), "\n")
cat("Upregulated in all 3 methods:", length(consensus_up), "\n")
cat("Downregulated in all 3 methods:", length(consensus_down), "\n")

# Need to fix
write.csv(data.frame(Gene_ID = consensus_all),
          "Results/DGE/Male_Only/Consensus_DEGs_All_Methods.csv",    row.names = FALSE)
write.csv(data.frame(Gene_ID = consensus_up),
          "Results/DGE/Male_Only/Consensus_DEGs_Upregulated.csv",    row.names = FALSE)
write.csv(data.frame(Gene_ID = consensus_down),
          "Results/DGE/Male_Only/Consensus_DEGs_Downregulated.csv",  row.names = FALSE)

# Compare top 100 DEG's across three methods

# edgeR — ranked by FDR
top100_edger <- DEG_male[order(DEG_male$FDR), ][1:100, ]
genes_top100_edger <- rownames(top100_edger)

# limma-voom — ranked by adj.P.Val
top100_voom <- DEG_male_voom[order(DEG_male_voom$adj.P.Val), ][1:100, ]
genes_top100_voom <- rownames(top100_voom)

# DESeq2 — ranked by padj
top100_deseq <- DEG_male_deseq[order(DEG_male_deseq$padj), ][1:100, ]
genes_top100_deseq <- top100_deseq$Gene_ID

overlap_er_voom  <- intersect(genes_top100_edger, genes_top100_voom)
overlap_er_deseq <- intersect(genes_top100_edger, genes_top100_deseq)
overlap_voom_deseq <- intersect(genes_top100_voom, genes_top100_deseq)
overlap_all      <- Reduce(intersect, list(genes_top100_edger, genes_top100_voom, genes_top100_deseq))

cat("edgeR ∩ limma-voom:          ", length(overlap_er_voom), "\n")
cat("edgeR ∩ DESeq2:              ", length(overlap_er_deseq), "\n")
cat("limma-voom ∩ DESeq2:         ", length(overlap_voom_deseq), "\n")
cat("All three methods:           ", length(overlap_all), "\n")

# 77 genes in top 100 common across all three methods. 

cat("Genes in all 3 top 100 lists:\n")
print(overlap_all)

venn_top100 <- list(
  edgeR        = genes_top100_edger,
  `limma-voom` = genes_top100_voom,
  DESeq2       = genes_top100_deseq
)

p_venn_top100 <- ggVennDiagram(venn_top100, label_alpha = 0, label = "count") +
  scale_fill_gradient(low = "#F7FBFF", high = "#2171B5") +
  scale_color_manual(values = c("black", "black", "black")) +
  labs(title = "Top 100 DEG Overlap Across Methods — Males (Heat vs Control)") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

print(p_venn_top100)

ggsave("Results/DGE/Male_Only/Venn_Top100_Male_All_Methods.png",
       p_venn_top100, width = 7, height = 6, dpi = 300)


# Build a summary table showing rank of each consensus gene in each method
rank_edger <- data.frame(
  Gene_ID   = rownames(DEG_male[order(DEG_male$FDR), ]),
  Rank_edgeR = seq_len(nrow(DEG_male))
)

rank_voom <- data.frame(
  Gene_ID    = rownames(DEG_male_voom[order(DEG_male_voom$adj.P.Val), ]),
  Rank_voom  = seq_len(nrow(DEG_male_voom))
)

rank_deseq <- data.frame(
  Gene_ID     = DEG_male_deseq$Gene_ID[order(DEG_male_deseq$padj)],
  Rank_DESeq2 = seq_len(nrow(DEG_male_deseq))
)

# Merge on consensus top 100 genes
rank_table <- data.frame(Gene_ID = overlap_all) %>%
  merge(rank_edger,  by = "Gene_ID") %>%
  merge(rank_voom,   by = "Gene_ID") %>%
  merge(rank_deseq,  by = "Gene_ID") %>%
  arrange(Rank_edgeR)

cat("\nRank comparison for genes in all 3 top 100 lists:\n")
print(rank_table)

write.csv(rank_table,
          "Results/DGE/Male_Only/Top100_Consensus_RankComparison.csv",
          row.names = FALSE)

```

# Differential Gene Expression w/ EdgeR, Limma, DESeq2 (Female ZEFI's Only)

```{r}

library(ggplot2)

### DGE with Edge R ###

library(edgeR)

AHDB1_AB_Female <- subset(AHDB1_AB, Sex == "F")

dim(AHDB1_AB_Female)

Counts_AB_Female <- Raw_Counts[, rownames(AHDB1_AB_Female)]

dge_AB_Female <- DGEList(
  counts  = Counts_AB_Female,
  samples = AHDB1_AB_Female
)

# Set factor and build design matrix
dge_AB_Female$samples$Treatment <- droplevels(factor(dge_AB_Female$samples$Treatment))

design_female <- model.matrix(~ 0 + Treatment, data = dge_AB_Female$samples)
colnames(design_female) <- levels(dge_AB_Female$samples$Treatment)
table(dge_AB_Female$samples$Treatment)

cat("Total genes (before filtering):", nrow(dge_AB_Female), "\n\n")

# filter using the design matrix
keep_female <- filterByExpr(dge_AB_Female, design_female)
dge_AB_Female <- dge_AB_Female[keep_female, , keep.lib.sizes = FALSE]
cat("Total genes (after filtering):", nrow(dge_AB_Female), "\n\n")

# TMM normalization
dge_AB_Female <- calcNormFactors(dge_AB_Female, method = "TMM")

dge_AB_Female <- estimateDisp(dge_AB_Female, design_female)
fit_female <- glmQLFit(dge_AB_Female, design_female)

contrast_female <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_female
)
cat("Contrast")
print(contrast_female)

qlf_female <- glmQLFTest(fit_female, contrast = contrast_female)

DEG_female <- topTags(qlf_female, n = Inf)$table
DEG_female$Gene_ID <- rownames(DEG_female)

sig_female      <- DEG_female[DEG_female$FDR < 0.05, ]
sig_female_up   <- sig_female[sig_female$logFC > 0, ]
sig_female_down <- sig_female[sig_female$logFC < 0, ]

cat("Female-Only DEG Results - EdgeR")
cat("Total significant DEGs:", nrow(sig_female), "\n")
cat("Upregulated in Heat:", nrow(sig_female_up), "\n")
cat("Downregulated in Heat:", nrow(sig_female_down), "\n\n")

#write.csv(DEG_female,      #"Results/DGE/Female_Only/DEG_Female_Heat_vs_Control_All.csv",         row.names = #FALSE)
#write.csv(sig_female,      #"Results/DGE/Female_Only/DEG_Female_Heat_vs_Control_Significant.csv", row.names = #FALSE)
#write.csv(sig_female_up,   "Results/DGE/Female_Only/DEG_Female_Upregulated.csv", #                row.names = FALSE)
#write.csv(sig_female_down, #"Results/DGE/Female_Only/DEG_Female_Downregulated.csv",               row.names = #FALSE)

### DGE with limma-voom ###

library(limma)

dge_AB_Female_voom <- DGEList(
  counts  = Counts_AB_Female,
  samples = AHDB1_AB_Female
)

# Sanity Check
table(dge_AB_Female_voom$samples$Treatment)
cat("Female-Only Subset\n")
cat("Total genes (before filtering):", nrow(dge_AB_Female_voom), "\n\n")

dge_AB_Female_voom$samples$Treatment <- droplevels(factor(dge_AB_Female_voom$samples$Treatment))

design_female_voom <- model.matrix(~ 0 + Treatment, data = dge_AB_Female_voom$samples)
colnames(design_female_voom) <- levels(dge_AB_Female_voom$samples$Treatment)

keep_female_voom <- filterByExpr(dge_AB_Female_voom, design_female_voom)
dge_AB_Female_voom <- dge_AB_Female_voom[keep_female_voom, , keep.lib.sizes = FALSE]

cat("Genes after filtering:", nrow(dge_AB_Female_voom), "\n\n")

dge_AB_Female_voom <- calcNormFactors(dge_AB_Female_voom, method = "TMM")

v_female <- voom(dge_AB_Female_voom, design_female_voom, plot = TRUE)

fit_female_voom <- lmFit(v_female, design_female_voom)

contrast_female_voom <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_female_voom
)
cat("Contrast:\n")
print(contrast_female_voom)

fit_female_voom <- contrasts.fit(fit_female_voom, contrast_female_voom)
fit_female_voom <- eBayes(fit_female_voom, trend = TRUE)

DEG_female_voom <- topTable(fit_female_voom, coef = "Heat_vs_Control", n = Inf, sort.by = "P")
DEG_female_voom$Gene_ID <- rownames(DEG_female_voom)

sig_female_voom      <- DEG_female_voom[DEG_female_voom$adj.P.Val < 0.05, ]
sig_female_voom_up   <- sig_female_voom[sig_female_voom$logFC > 0, ]
sig_female_voom_down <- sig_female_voom[sig_female_voom$logFC < 0, ]

cat("Female-Only DEG Results (limma-voom)\n")
cat("Total significant DEGs:", nrow(sig_female_voom), "\n")
cat("Upregulated in Heat:", nrow(sig_female_voom_up), "\n")
cat("Downregulated in Heat:", nrow(sig_female_voom_down), "\n\n")

#write.csv(DEG_female_voom,      #"Results/DGE/Female_Only/Voom_DEG_Female_Heat_vs_Control_All.csv",         #row.names = FALSE)
#write.csv(sig_female_voom,      #"Results/DGE/Female_Only/Voom_DEG_Female_Heat_vs_Control_Significant.csv", #row.names = FALSE)
#write.csv(sig_female_voom_up,   #"Results/DGE/Female_Only/Voom_DEG_Female_Upregulated.csv",                 #row.names = FALSE)
#write.csv(sig_female_voom_down, #"Results/DGE/Female_Only/Voom_DEG_Female_Downregulated.csv",               #row.names = FALSE)

# Plot mean variance trend
v_female <- voom(dge_AB_Female_voom, design_female_voom, plot = TRUE)

### DGE with DESeq2 ###

library(DESeq2)

AHDB1_AB_Female_deseq <- subset(AHDB1_AB, Sex == "F")
Counts_AB_Female_deseq <- Raw_Counts[, rownames(AHDB1_AB_Female_deseq)]

Counts_AB_Female_int <- round(Counts_AB_Female_deseq)

# Build colData and set reference level
col_data_female <- AHDB1_AB_Female_deseq
col_data_female$Treatment <- droplevels(factor(col_data_female$Treatment))
col_data_female$Treatment <- relevel(col_data_female$Treatment, ref = "Control")

cat("Female-Only Subset\n")
cat("Total genes (before filtering):", nrow(Counts_AB_Female_int), "\n\n")

# Build DESeq2 object
dds_female <- DESeqDataSetFromMatrix(
  countData = Counts_AB_Female_int,
  colData   = col_data_female,
  design    = ~ Treatment
)

keep_female_deseq <- rowSums(counts(dds_female)) >= 10
dds_female <- dds_female[keep_female_deseq, ]
cat("Genes after filtering:", nrow(dds_female), "\n\n")

dds_female <- DESeq(dds_female)

res_female <- results(
  dds_female,
  contrast = c("Treatment", "Heat", "Control"),
  alpha    = 0.05
)

summary(res_female)

DEG_female_deseq <- as.data.frame(res_female)
DEG_female_deseq$Gene_ID <- rownames(DEG_female_deseq)

# Remove NA's
DEG_female_deseq <- DEG_female_deseq[
  !is.na(DEG_female_deseq$padj) & !is.na(DEG_female_deseq$log2FoldChange),
]
cat("Genes after NA removal:", nrow(DEG_female_deseq), "\n\n")

# Subset DEG's into upregulated and downregulated
sig_female_deseq      <- DEG_female_deseq[DEG_female_deseq$padj < 0.05, ]
sig_female_deseq_up   <- sig_female_deseq[sig_female_deseq$log2FoldChange > 0, ]
sig_female_deseq_down <- sig_female_deseq[sig_female_deseq$log2FoldChange < 0, ]

cat("Female-Only DEG Results (DESeq2)\n")
cat("Total significant DEGs:", nrow(sig_female_deseq), "\n")
cat("Upregulated in Heat:", nrow(sig_female_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(sig_female_deseq_down), "\n\n")

#write.csv(DEG_female_deseq,      #"Results/DGE/Female_Only/DESeq2_DEG_Female_Heat_vs_Control_All.csv",         #row.names = FALSE)
#write.csv(sig_female_deseq,      #"Results/DGE/Female_Only/DESeq2_DEG_Female_Heat_vs_Control_Significant.csv", #row.names = FALSE)
#write.csv(sig_female_deseq_up,   #"Results/DGE/Female_Only/DESeq2_DEG_Female_Upregulated.csv",                 #row.names = FALSE)
#write.csv(sig_female_deseq_down, #"Results/DGE/Female_Only/DESeq2_DEG_Female_Downregulated.csv",               #row.names = FALSE)

# Volcano Plots for all three methods

library(patchwork)
library(ggVennDiagram)

p_edger_f <- make_volcano(DEG_female,             "edgeR")
p_voom_f  <- make_volcano_voom(DEG_female_voom,   "limma-voom")
p_deseq_f <- make_volcano_deseq(DEG_female_deseq, "DESeq2")

combined_volcano_female <- (p_edger_f | p_voom_f | p_deseq_f) +
  plot_annotation(
    title    = "Heat vs Control — Females Only",
    subtitle = "edgeR  |  limma-voom  |  DESeq2",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 15, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray40")
    )
  ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

print(combined_volcano_female)

ggsave("Results/DGE/Female_Only/Volcano_Female_All_Methods.png",
       combined_volcano_female, width = 18, height = 6, dpi = 300)

venn_list_f <- list(
  edgeR        = rownames(sig_female),
  `limma-voom` = rownames(sig_female_voom),
  DESeq2       = sig_female_deseq$Gene_ID
)

venn_up_f <- list(
  edgeR        = rownames(sig_female_up),
  `limma-voom` = rownames(sig_female_voom_up),
  DESeq2       = sig_female_deseq_up$Gene_ID
)

venn_down_f <- list(
  edgeR        = rownames(sig_female_down),
  `limma-voom` = rownames(sig_female_voom_down),
  DESeq2       = sig_female_deseq_down$Gene_ID
)

# All DEGs
p_venn_f <- ggVennDiagram(venn_list_f, label_alpha = 0, label = "count") +
  scale_fill_gradient(low = "#F7FBFF", high = "#2171B5") +
  scale_color_manual(values = c("black", "black", "black")) +
  labs(title = "All DEGs") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

# Upregulated
p_venn_up_f <- ggVennDiagram(venn_up_f, label_alpha = 0, label = "count") +
  scale_fill_gradient(low = "#FFF5F0", high = "#CB181D") +
  scale_color_manual(values = c("black", "black", "black")) +
  labs(title = "Upregulated in Heat") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

# Downregulated
p_venn_down_f <- ggVennDiagram(venn_down_f, label_alpha = 0, label = "count") +
  scale_fill_gradient(low = "#F7FBFF", high = "#2171B5") +
  scale_color_manual(values = c("black", "black", "black")) +
  labs(title = "Downregulated in Heat") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

# Combined Venn
combined_venn_female <- (p_venn_f / (p_venn_up_f | p_venn_down_f)) +
  plot_annotation(
    title = "DEG Overlap Across Methods — Females (Heat vs Control)",
    theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  )

print(combined_venn_female)

ggsave("Results/DGE/Female_Only/Venn_Female_Combined.png",
       combined_venn_female, width = 12, height = 12, dpi = 300)

consensus_all_f  <- Reduce(intersect, venn_list_f)
consensus_up_f   <- Reduce(intersect, venn_up_f)
consensus_down_f <- Reduce(intersect, venn_down_f)

cat("Genes significant in all 3 methods:", length(consensus_all_f), "\n")
cat("Upregulated in all 3 methods:", length(consensus_up_f), "\n")
cat("Downregulated in all 3 methods:", length(consensus_down_f), "\n")

write.csv(data.frame(Gene_ID = consensus_all_f),
          "Results/DGE/Female_Only/Consensus_DEGs_All_Methods.csv",   row.names = FALSE)
write.csv(data.frame(Gene_ID = consensus_up_f),
          "Results/DGE/Female_Only/Consensus_DEGs_Upregulated.csv",   row.names = FALSE)
write.csv(data.frame(Gene_ID = consensus_down_f),
          "Results/DGE/Female_Only/Consensus_DEGs_Downregulated.csv", row.names = FALSE)

# Compare top 100 DEG's across three methods

# edgeR — ranked by FDR
top100_edger_f <- DEG_female[order(DEG_female$FDR), ][1:100, ]
genes_top100_edger_f <- rownames(top100_edger_f)

# limma-voom — ranked by adj.P.Val
top100_voom_f <- DEG_female_voom[order(DEG_female_voom$adj.P.Val), ][1:100, ]
genes_top100_voom_f <- rownames(top100_voom_f)

# DESeq2 — ranked by padj
top100_deseq_f <- DEG_female_deseq[order(DEG_female_deseq$padj), ][1:100, ]
genes_top100_deseq_f <- top100_deseq_f$Gene_ID

overlap_er_voom_f    <- intersect(genes_top100_edger_f, genes_top100_voom_f)
overlap_er_deseq_f   <- intersect(genes_top100_edger_f, genes_top100_deseq_f)
overlap_voom_deseq_f <- intersect(genes_top100_voom_f,  genes_top100_deseq_f)
overlap_all_f        <- Reduce(intersect, list(genes_top100_edger_f, genes_top100_voom_f, genes_top100_deseq_f))

cat("edgeR ∩ limma-voom:          ", length(overlap_er_voom_f), "\n")
cat("edgeR ∩ DESeq2:              ", length(overlap_er_deseq_f), "\n")
cat("limma-voom ∩ DESeq2:         ", length(overlap_voom_deseq_f), "\n")
cat("All three methods:           ", length(overlap_all_f), "\n")

cat("Genes in all 3 top 100 lists:\n")
print(overlap_all_f)

venn_top100_f <- list(
  edgeR        = genes_top100_edger_f,
  `limma-voom` = genes_top100_voom_f,
  DESeq2       = genes_top100_deseq_f
)

p_venn_top100_f <- ggVennDiagram(venn_top100_f, label_alpha = 0, label = "count") +
  scale_fill_gradient(low = "#F7FBFF", high = "#2171B5") +
  scale_color_manual(values = c("black", "black", "black")) +
  labs(title = "Top 100 DEG Overlap Across Methods — Females (Heat vs Control)") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

print(p_venn_top100_f)

ggsave("Results/DGE/Female_Only/Venn_Top100_Female_All_Methods.png",
       p_venn_top100_f, width = 7, height = 6, dpi = 300)

# Build a summary table showing rank of each consensus gene in each method
rank_edger_f <- data.frame(
  Gene_ID    = rownames(DEG_female[order(DEG_female$FDR), ]),
  Rank_edgeR = seq_len(nrow(DEG_female))
)

rank_voom_f <- data.frame(
  Gene_ID   = rownames(DEG_female_voom[order(DEG_female_voom$adj.P.Val), ]),
  Rank_voom = seq_len(nrow(DEG_female_voom))
)

rank_deseq_f <- data.frame(
  Gene_ID     = DEG_female_deseq$Gene_ID[order(DEG_female_deseq$padj)],
  Rank_DESeq2 = seq_len(nrow(DEG_female_deseq))
)

# Merge on consensus top 100 genes
rank_table_f <- data.frame(Gene_ID = overlap_all_f) %>%
  merge(rank_edger_f,  by = "Gene_ID") %>%
  merge(rank_voom_f,   by = "Gene_ID") %>%
  merge(rank_deseq_f,  by = "Gene_ID") %>%
  arrange(Rank_edgeR)

cat("\nRank comparison for genes in all 3 top 100 lists:\n")
print(rank_table_f)

write.csv(rank_table_f,
          "Results/DGE/Female_Only/Top100_Consensus_RankComparison.csv",
          row.names = FALSE)

```

# (1) Ignore:  Normalize Count Data for A vs B, (2) Run DGE with EdgeR, Limma, & DESeq2 on Male & Females Combined

```{r}

dge_AB <- calcNormFactors(dge_AB, method = "TMM")
cat("Normalization factors:\n")
print(dge_AB$samples$norm.factors)

# Calculate log-CPM for normalized data
lcpm_AB_norm <- cpm(dge_AB, log=TRUE, prior.count = 1)
dim(lcpm_AB_norm)

# Export normalized counts
# write.csv(lcpm_AB_norm, file = "Results/NormalizedCounts/NormalisedCount_AB_CombinedSexes.csv")

# Set factors
dge_AB$samples$Sex <- factor(dge_AB$samples$Sex)
dge_AB$samples$Treatment <- factor(dge_AB$samples$Treatment)

# Combined group
dge_AB$samples$group <- interaction(
  dge_AB$samples$Treatment,
  dge_AB$samples$Sex,
  sep = "_"
)

# Sanity Check for group distributions
cat("Group distribution:\n")
print(table(dge_AB$samples$group))

# Design matrix
design <- model.matrix(~ 0 + group, data = dge_AB$samples)
colnames(design) <- levels(dge_AB$samples$group)
cat("Design matrix:\n")
print(design)

# Filter low-expressed genes
keep <- filterByExpr(dge_AB, design)
genes_after_filter <- sum(keep)

cat("Genes before filtering:", nrow(dge_AB), "\n")
cat("Genes after filtering:", genes_after_filter, "\n")
cat("Genes removed:", nrow(dge_AB) - genes_after_filter, "\n")
cat("Retention rate:", round(genes_after_filter / nrow(dge_AB) * 100, 2), "%\n\n")

dge_AB <- dge_AB[keep, , keep.lib.sizes = FALSE]

# Re-normalize after filtering
dge_AB <- calcNormFactors(dge_AB)

# Estimate dispersions and fit model
dge_AB <- estimateDisp(dge_AB, design)
fit <- glmQLFit(dge_AB, design)

# Define contrasts
my.contrasts <- makeContrasts(
  # Heat effect (averaged over sex)
  Heat_vs_Control =
    (Heat_M + Heat_F) - (Control_M + Control_F),
  
  # Sex effect (averaged over treatment)
  Sex_effect =
    (Control_M + Heat_M) - (Control_F + Heat_F),
  
  # Sex-specific heat response
  Interaction =
    (Heat_M - Control_M) - (Heat_F - Control_F),
  
  levels = design
)

cat("Contrasts defined:\n")
print(my.contrasts)

# Run QLF-tests
qlf_heat <- glmQLFTest(fit, contrast = my.contrasts[, "Heat_vs_Control"])
qlf_sex  <- glmQLFTest(fit, contrast = my.contrasts[, "Sex_effect"])
qlf_int  <- glmQLFTest(fit, contrast = my.contrasts[, "Interaction"])

# Extract results
DEG_heat <- topTags(qlf_heat, n = Inf)$table
DEG_sex  <- topTags(qlf_sex,  n = Inf)$table
DEG_int  <- topTags(qlf_int,  n = Inf)$table

# Add Gene IDs
DEG_heat$Gene_ID <- rownames(DEG_heat)
DEG_sex$Gene_ID  <- rownames(DEG_sex)
DEG_int$Gene_ID  <- rownames(DEG_int)

# Identify significant genes (FDR < 0.05)
sig_heat <- DEG_heat[DEG_heat$FDR < 0.05, ]
sig_sex  <- DEG_sex[DEG_sex$FDR < 0.05, ]
sig_int  <- DEG_int[DEG_int$FDR < 0.05, ]

# Further categorize by direction
sig_heat_up <- sig_heat[sig_heat$logFC > 0, ]
sig_heat_down <- sig_heat[sig_heat$logFC < 0, ]

sig_sex_male <- sig_sex[sig_sex$logFC > 0, ]  # Higher in males
sig_sex_female <- sig_sex[sig_sex$logFC < 0, ]  # Higher in females

sig_int_male <- sig_int[sig_int$logFC > 0, ]  # Stronger heat response in males
sig_int_female <- sig_int[sig_int$logFC < 0, ]  # Stronger heat response in females

cat("Heat vs Control:\n")
cat("  Total significant DEGs (FDR < 0.05):", nrow(sig_heat), "\n")
cat("  Upregulated in Heat:", nrow(sig_heat_up), "\n")
cat("  Downregulated in Heat:", nrow(sig_heat_down), "\n\n")

cat("Sex Effect (Male vs Female):\n")
cat("  Total significant DEGs (FDR < 0.05):", nrow(sig_sex), "\n")
cat("  Higher in Males:", nrow(sig_sex_male), "\n")
cat("  Higher in Females:", nrow(sig_sex_female), "\n\n")

cat("Treatment × Sex Interaction:\n")
cat("  Total significant DEGs (FDR < 0.05):", nrow(sig_int), "\n")
cat("  Stronger heat response in Males:", nrow(sig_int_male), "\n")
cat("  Stronger heat response in Females:", nrow(sig_int_female), "\n\n")

# Export results
write.csv(DEG_heat, "Results/DGE/Male_Female_Combined/Heat_vs_Control_AllGenes.csv", row.names = FALSE)
write.csv(sig_heat, "Results/DGE/Male_Female_Combined/DEG_Heat_vs_Control_Significant.csv", row.names = FALSE)

write.csv(sig_heat_up, "Results/DGE/Male_Female_Combined/DEG_Heat_Upregulated.csv", row.names = FALSE)
write.csv(sig_heat_down, "Results/DGE/Male_Female_Combined/DEG_Heat_Downregulated.csv", row.names = FALSE)

write.csv(DEG_sex, "Results/DGE/Male_Female_Combined/Sex_Effect_AllGenes.csv", row.names = FALSE)
write.csv(sig_sex, "Results/DGE/Male_Female_Combined/DEG_Sex_Effect_Significant.csv", row.names = FALSE)

write.csv(sig_sex_male, "Results/DGE/Male_Female_Combined/DEG_Sex_Male_Higher.csv", row.names = FALSE)
write.csv(sig_sex_female, "Results/DGE/Male_Female_Combined/DEG_Sex_Female_Higher.csv", row.names = FALSE)

write.csv(DEG_int, "Results/DGE/Male_Female_Combined/Interaction_AllGenes.csv", row.names = FALSE)
write.csv(sig_int, "Results/DGE/Male_Female_Combined/DEG_Interaction_Significant.csv", row.names = FALSE)

```


