---
title: "PathwayEnrichment — Liver Time-Series"
author: "Ali Amer"
date: "2026-08-07"
output: html_document
---

# Script Goal:
# 1. Load Timeseries_spline_results.csv from the liver A-E timecourse script
#    (Symbol, spline FDR, per-contrast logFC_/FDR_ columns, Cluster)
# 2. Map Symbol -> Entrez via biomaRt
# 3. Build one ranked list per adjacent/endpoint contrast:
#    sign(logFC) * -log10(FDR), Inf-guarded
# 4. GSEA per contrast: gseKEGG (tgu), gseGO (BP/MF/CC), MSigDB Hallmark,
#    MitoCarta3.0
# 5. ORA per temporal cluster (k-means clusters from the timecourse script)
#    via compareCluster with universe = all tested genes: GO, KEGG, Hallmark

# Clear Memory and Check working directory

```{r}

rm(list = ls(all = TRUE))
getwd()

```

# Load packages

```{r}
library(dplyr)
library(tidyr)
library(readxl)
library(biomaRt)
library(clusterProfiler)
library(AnnotationHub)
library(ggplot2)
library(patchwork)
library(msigdbr)
library(enrichplot)
```

# CONFIG

```{r}

CONFIG <- list(
  tissue      = "Liver",
  results_csv = "DifferentialGeneExpression/Timeseries_spline_results.csv",
  out_root    = "Results/Enrichment_Timecourse_Liver",

  # Contrast columns present as logFC_<name> / FDR_<name> in the results CSV
  contrasts = c("Heat_vs_Control", "Rec2h_vs_Heat", "Rec24h_vs_Rec2h",
                "Rec2wk_vs_Rec24h", "Rec2wk_vs_Control"),

  # Ensembl BioMart
  biomart_dataset = "tguttata_gene_ensembl",
  biomart_version = 115,

  # KEGG
  kegg_organism = "tgu",
  pvalue_cutoff = 0.05,

  # GO — via AnnotationHub OrgDb (org.Taeniopygia_guttata)
  go_ontologies = c("BP", "MF", "CC"),
  go_key_type   = "SYMBOL",
  go_min_gs     = 10,
  go_max_gs     = 500,
  go_show_cat   = 10,

  # MitoCarta
  mitocarta_xls = "Human.MitoCarta3.0.xls",

  # Clusters (from the timecourse script's k-means)
  n_clusters = 6
)

dir.create(CONFIG$out_root, recursive = TRUE, showWarnings = FALSE)
for (cn in CONFIG$contrasts) {
  dir.create(file.path(CONFIG$out_root, cn), recursive = TRUE,
             showWarnings = FALSE)
}
dir.create(file.path(CONFIG$out_root, "Clusters"), recursive = TRUE,
           showWarnings = FALSE)
```

# Load time-series results

```{r}
res <- read.csv(CONFIG$results_csv, stringsAsFactors = FALSE,
                check.names = FALSE)
cat(CONFIG$tissue, "time-series rows loaded:", nrow(res), "\n")

stopifnot("Symbol" %in% colnames(res),
          "Cluster" %in% colnames(res),
          all(paste0("logFC_", CONFIG$contrasts) %in% colnames(res)),
          all(paste0("FDR_",   CONFIG$contrasts) %in% colnames(res)))
```

# BioMart: map Symbol -> Entrez

```{r}

mart_finch <- useEnsembl(
  biomart = "ensembl",
  dataset = CONFIG$biomart_dataset,
  version = CONFIG$biomart_version
)

all_symbols <- unique(res$Symbol)
cat("Unique symbols to map:", length(all_symbols), "\n\n")

map_ZF <- getBM(
  attributes = c("external_gene_name", "entrezgene_id"),
  filters    = "external_gene_name",
  values     = all_symbols,
  mart       = mart_finch
)

res$Entrez <- map_ZF$entrezgene_id[match(res$Symbol, map_ZF$external_gene_name)]

cat("Mapping results:\n")
cat("  Total genes:", nrow(res), "\n")
cat("  With Entrez IDs:", sum(!is.na(res$Entrez)), "\n")
cat("  Success rate:", round(sum(!is.na(res$Entrez)) / nrow(res) * 100, 1),
    "%\n\n")
```

# AnnotationHub: fetch the NCBI-built OrgDb for Taeniopygia guttata

```{r}

ah <- AnnotationHub()
zf_orgdb_hits <- query(ah, c("OrgDb", "Taeniopygia guttata"))
print(zf_orgdb_hits)

orgdb_ZF <- zf_orgdb_hits[[length(zf_orgdb_hits)]]

cat("Available keytypes:\n")
print(keytypes(orgdb_ZF))
```

# Gene-set collections used across contrasts (Hallmark, MitoCarta)

```{r}
hallmark <- msigdbr(species = "Homo sapiens", category = "H")
hallmark_df <- hallmark %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::distinct()

mc <- readxl::read_excel(CONFIG$mitocarta_xls,
                         sheet = "A Human MitoCarta3.0")

mitocarta_paths <- mc %>%
  dplyr::filter(MitoCarta3.0_MitoPathways != "0") %>%
  dplyr::select(Symbol, MitoCarta3.0_MitoPathways) %>%
  tidyr::separate_rows(MitoCarta3.0_MitoPathways, sep = "\\s*\\|\\s*") %>%
  dplyr::mutate(gs_name = trimws(sub(".*>\\s*", "", MitoCarta3.0_MitoPathways))) %>%
  dplyr::select(gs_name, gene_symbol = Symbol) %>%
  dplyr::distinct()

mitocarta_df <- dplyr::bind_rows(
  mitocarta_paths,
  data.frame(gs_name = "MITOCARTA_ALL", gene_symbol = unique(mc$Symbol))
)

cat("MitoCarta genes:", length(unique(mc$Symbol)),
    "| pathway terms:", length(unique(mitocarta_paths$gs_name)), "\n")
```

# Ranked-list builder: sign(logFC) * -log10(FDR), Inf-guarded

```{r}

make_rank <- function(df, contrast, id_col) {
  lfc <- df[[paste0("logFC_", contrast)]]
  fdr <- df[[paste0("FDR_",   contrast)]]

  ## Guard against FDR == 0 producing Inf ranks
  fdr_min <- min(fdr[fdr > 0], na.rm = TRUE)
  fdr     <- pmax(fdr, fdr_min / 10)

  rk  <- sign(lfc) * -log10(fdr)
  ids <- as.character(df[[id_col]])

  keep <- !is.na(rk) & !is.na(ids) & ids != ""
  gl   <- setNames(rk[keep], ids[keep])
  gl   <- gl[!duplicated(names(gl))]
  sort(gl, decreasing = TRUE)
}
```

# GSEA helpers (one call per contrast)

```{r}

save_gsea <- function(res_obj, out_dir, file_stub, title) {
  if (is.null(res_obj) || nrow(res_obj@result) == 0) {
    cat("  No significant sets for", file_stub, "\n\n")
    return(invisible(NULL))
  }
  cat("  \u2713 Found", nrow(res_obj@result), "sets for", file_stub, "\n\n")

  write.csv(res_obj@result,
            file.path(out_dir, paste0(file_stub, ".csv")),
            row.names = FALSE)

  p <- enrichplot::dotplot(res_obj, showCategory = 20, split = ".sign",
                           title = title) +
    ggplot2::facet_grid(. ~ .sign,
      labeller = ggplot2::labeller(.sign = c(activated  = "Up",
                                             suppressed = "Down")))

  ggplot2::ggsave(file.path(out_dir, paste0(file_stub, ".png")),
                  p, width = 14, height = 10, dpi = 300)
  print(p)
  invisible(res_obj)
}

run_contrast_gsea <- function(contrast) {

  out_dir <- file.path(CONFIG$out_root, contrast)
  cat("### GSEA:", CONFIG$tissue, contrast, "###\n")

  gl_entrez <- make_rank(res, contrast, "Entrez")
  gl_symbol <- make_rank(res, contrast, "Symbol")
  cat("  Ranked genes (Entrez):", length(gl_entrez),
      "| (Symbol):", length(gl_symbol), "\n\n")

  ## KEGG
  gk <- gseKEGG(gl_entrez,
                organism     = CONFIG$kegg_organism,
                pvalueCutoff = CONFIG$pvalue_cutoff,
                verbose      = FALSE)
  save_gsea(gk, out_dir, paste0("GSE_KEGG_", contrast),
            paste(CONFIG$tissue, contrast, "GSEA KEGG"))

  ## GO (BP / MF / CC, combined with patchwork)
  go_plots <- list()
  for (ont in CONFIG$go_ontologies) {
    gg <- gseGO(gl_symbol,
                OrgDb        = orgdb_ZF,
                keyType      = CONFIG$go_key_type,
                ont          = ont,
                minGSSize    = CONFIG$go_min_gs,
                maxGSSize    = CONFIG$go_max_gs,
                pvalueCutoff = CONFIG$pvalue_cutoff,
                verbose      = FALSE)
    if (!is.null(gg) && nrow(gg@result) > 0) {
      cat("  \u2713 Found", nrow(gg@result), ont, "terms\n")
      write.csv(gg@result,
                file.path(out_dir, paste0("GO_GSEA_", contrast, "_", ont, ".csv")),
                row.names = FALSE)
      go_plots[[ont]] <- dotplot(gg, showCategory = CONFIG$go_show_cat,
                                 split = ".sign",
                                 title = paste0(contrast, " - ", ont)) +
        facet_grid(. ~ .sign,
          labeller = labeller(.sign = c(activated  = "Up",
                                        suppressed = "Down")))
    } else {
      cat("  No significant", ont, "terms.\n")
    }
  }
  if (length(go_plots) > 0) {
    combined_go <- wrap_plots(go_plots, ncol = 1) +
      plot_annotation(title = paste(CONFIG$tissue, contrast,
                                    "GO GSEA — BP / MF / CC"))
    ggsave(file.path(out_dir, paste0("GO_GSEA_", contrast, "_combined.png")),
           combined_go, width = 14, height = 8 * length(go_plots), dpi = 300)
    print(combined_go)
  }

  ## MSigDB Hallmark
  gh <- GSEA(geneList     = gl_symbol,
             TERM2GENE    = hallmark_df,
             minGSSize    = 10,
             maxGSSize    = 500,
             pvalueCutoff = CONFIG$pvalue_cutoff,
             verbose      = FALSE)
  save_gsea(gh, out_dir, paste0("GSEA_MSigDB_", contrast),
            paste(CONFIG$tissue, contrast, "GSEA MSigDB Hallmark"))

  ## MitoCarta
  gm <- GSEA(geneList     = gl_symbol,
             TERM2GENE    = mitocarta_df,
             minGSSize    = 3,
             maxGSSize    = 500,
             pvalueCutoff = CONFIG$pvalue_cutoff,
             verbose      = FALSE)
  save_gsea(gm, out_dir, paste0("GSEA_MitoCarta_", contrast),
            paste(CONFIG$tissue, contrast, "GSEA MitoCarta3.0"))

  invisible(NULL)
}
```

### GSEA: all contrasts ###

```{r, fig.width = 14, fig.height = 10}
for (cn in CONFIG$contrasts) run_contrast_gsea(cn)
```

# ORA per temporal cluster

# Cluster membership comes from the timecourse script's k-means (Cluster
# column, NA for non-significant genes). Universe = all tested genes, so the
# background matches the filterByExpr-retained set rather than the whole
# genome annotation.

```{r}
cluster_symbols <- lapply(seq_len(CONFIG$n_clusters), function(k) {
  unique(na.omit(res$Symbol[res$Cluster %in% k & res$Symbol != ""]))
})
names(cluster_symbols) <- paste0("Cluster_", seq_len(CONFIG$n_clusters))

cluster_entrez <- lapply(seq_len(CONFIG$n_clusters), function(k) {
  unique(na.omit(as.character(res$Entrez[res$Cluster %in% k])))
})
names(cluster_entrez) <- names(cluster_symbols)

print(lengths(cluster_symbols))

universe_symbol <- unique(res$Symbol[!is.na(res$Symbol) & res$Symbol != ""])
universe_entrez <- unique(na.omit(as.character(res$Entrez)))
cat("Universe:", length(universe_symbol), "symbols |",
    length(universe_entrez), "Entrez\n")

cluster_symbols <- cluster_symbols[lengths(cluster_symbols) >= 10]
cluster_entrez  <- cluster_entrez[lengths(cluster_entrez)  >= 10]
```

### ORA GO per cluster ###

```{r, fig.width = 12, fig.height = 10}
for (ont in CONFIG$go_ontologies) {
  cat("### ORA GO:", CONFIG$tissue, "clusters -", ont, "###\n")

  go_cc <- compareCluster(
    geneCluster  = cluster_symbols,
    fun          = "enrichGO",
    OrgDb        = orgdb_ZF,
    keyType      = CONFIG$go_key_type,
    ont          = ont,
    universe     = universe_symbol,
    minGSSize    = CONFIG$go_min_gs,
    maxGSSize    = CONFIG$go_max_gs,
    pvalueCutoff = CONFIG$pvalue_cutoff
  )

  if (!is.null(go_cc) && nrow(go_cc@compareClusterResult) > 0) {
    cat("  \u2713 Found", nrow(go_cc@compareClusterResult), ont, "rows\n\n")

    write.csv(go_cc@compareClusterResult,
              file.path(CONFIG$out_root, "Clusters",
                        paste0("ORA_GO_Clusters_", ont, ".csv")),
              row.names = FALSE)

    p <- dotplot(go_cc, showCategory = CONFIG$go_show_cat) +
      labs(title = paste(CONFIG$tissue, "temporal clusters — GO", ont))

    ggsave(file.path(CONFIG$out_root, "Clusters",
                     paste0("ORA_GO_Clusters_", ont, ".png")),
           p, width = 12, height = 10, dpi = 300)
    print(p)
  } else {
    cat("  No significant", ont, "terms across clusters.\n\n")
  }
}
```

### ORA KEGG per cluster ###

```{r, fig.width = 12, fig.height = 10}
cat("### ORA KEGG:", CONFIG$tissue, "clusters ###\n")

kegg_cc <- compareCluster(
  geneCluster  = cluster_entrez,
  fun          = "enrichKEGG",
  organism     = CONFIG$kegg_organism,
  universe     = universe_entrez,
  pvalueCutoff = CONFIG$pvalue_cutoff
)

if (!is.null(kegg_cc) && nrow(kegg_cc@compareClusterResult) > 0) {
  cat("  \u2713 Found", nrow(kegg_cc@compareClusterResult), "rows\n\n")

  write.csv(kegg_cc@compareClusterResult,
            file.path(CONFIG$out_root, "Clusters", "ORA_KEGG_Clusters.csv"),
            row.names = FALSE)

  p <- dotplot(kegg_cc, showCategory = CONFIG$go_show_cat) +
    labs(title = paste(CONFIG$tissue, "temporal clusters — KEGG"))

  ggsave(file.path(CONFIG$out_root, "Clusters", "ORA_KEGG_Clusters.png"),
         p, width = 12, height = 10, dpi = 300)
  print(p)
} else {
  cat("  No significant KEGG pathways across clusters.\n\n")
}
```

### ORA MSigDB Hallmark per cluster ###

```{r, fig.width = 12, fig.height = 10}
cat("### ORA Hallmark:", CONFIG$tissue, "clusters ###\n")

hall_cc <- compareCluster(
  geneCluster  = cluster_symbols,
  fun          = "enricher",
  TERM2GENE    = hallmark_df,
  universe     = universe_symbol,
  pvalueCutoff = CONFIG$pvalue_cutoff
)

if (!is.null(hall_cc) && nrow(hall_cc@compareClusterResult) > 0) {
  cat("  \u2713 Found", nrow(hall_cc@compareClusterResult), "rows\n\n")

  write.csv(hall_cc@compareClusterResult,
            file.path(CONFIG$out_root, "Clusters", "ORA_MSigDB_Clusters.csv"),
            row.names = FALSE)

  p <- dotplot(hall_cc, showCategory = CONFIG$go_show_cat, font.size = 12) +
    scale_color_gradient(low = "#2c7bb6", high = "#d7191c") +
    scale_size(range = c(3, 8)) +
    labs(title = paste(CONFIG$tissue, "temporal clusters — MSigDB Hallmark"),
         x = "Gene Ratio", y = NULL) +
    theme_bw(base_size = 12) +
    theme(strip.text = element_text(size = 12, face = "bold"),
          axis.text.y = element_text(size = 10),
          panel.grid.minor = element_blank())

  ggsave(file.path(CONFIG$out_root, "Clusters", "ORA_MSigDB_Clusters.png"),
         p, width = 12, height = 10, dpi = 300)
  print(p)
} else {
  cat("  No significant Hallmark sets across clusters.\n\n")
}
```

```{r}
sessionInfo()
```
