---
title: "6_DGE_DESeq2_LimmaVoom_EdgeR_Comparison"
author: "Ali Amer, modified by Brynleigh Payne"
date: "2026-06-22"
output: md_document
---

# Script Goal: 
# 1. Convert raw count reads of samples sent for Liver RNA-Seq (A vs B) from prep.dy script   
# 2. Check library sizes across A vs B & Normalize raw count data
# 3. Run DGE with edgeR, limma and DESeq2 (Males Only & Females Only)
# 4. Run Pathway Enrichment & WGCNA

#### Clear Memory, Check Directory & Load Libraries
```{r}

# Clear memory
rm(list=ls(all = TRUE))

# Check working directory 
getwd()

library(edgeR)
library(DESeq2)
library(limma)

```

#### Input the count data, the gene(/transcript) count matrix and labels

```{r}
Raw_Counts <- as.matrix(read.csv("ZF_Liver_rawgene_count_matrix.csv", row.names="gene_id"))
dim(Raw_Counts)
head(Raw_Counts)
```

#### Input the meta data or phenotype data
# Note: The PHENO_DATA file contains information on each sample, e.g., sex or population. The exact way to import this depends on the format of the file.
##  Make sure the individual names match between the count data and the PHENO_DATA

```{r}
ZF_Liver_sample_list <-(read.table("PHENO_DATA_Liver_Control_Heat.txt", header=TRUE, row.names=1, sep = "\t"))
dim(ZF_Liver_sample_list)
head(ZF_Liver_sample_list)
```

## Subset the countdata to only include Control (A) and Heat (B) samples 

```{r}
samples_to_keep <- rownames(ZF_Liver_sample_list)

length(intersect(samples_to_keep, colnames(Raw_Counts)))

all(samples_to_keep %in% colnames(Raw_Counts))

setdiff(samples_to_keep, colnames(Raw_Counts))

setdiff(colnames(Raw_Counts), samples_to_keep)

common_samples <- intersect(samples_to_keep, colnames(Raw_Counts))

Raw_Counts_AB <- Raw_Counts[, common_samples]
ZF_Liver_sample_list <- ZF_Liver_sample_list[common_samples, ]

ZF_Liver_sample_list <- ZF_Liver_sample_list[colnames(Raw_Counts_AB), ]

identical(rownames(ZF_Liver_sample_list), colnames(Raw_Counts_AB))
```

# Check all sample IDs in colData are also in CountData and match their orders

```{r}
all(rownames(ZF_Liver_sample_list) %in% colnames(Raw_Counts_AB))
Raw_Counts_AB <- Raw_Counts_AB[, rownames(ZF_Liver_sample_list)]
all(rownames(ZF_Liver_sample_list) == colnames(Raw_Counts_AB))
```

## Manually set reference levels for sex and treatment

```{r}
ZF_Liver_sample_list$sex <- relevel(factor(ZF_Liver_sample_list$sex), ref = "female")
ZF_Liver_sample_list$treatment <- relevel(factor(ZF_Liver_sample_list$treatment), ref = "Control")
```

# Construct DGEList object
```{r}

DGE_AB <- DGEList(
  counts  = Raw_Counts_AB,
  samples = ZF_Liver_sample_list
)
```

# Calculate Library Size & Make BoxPlots (A vs B)

```{r}

cpm_AB <- cpm(DGE_AB)
lcpm_AB <- cpm(DGE_AB, log=TRUE)
summary(lcpm_AB)

# Boxplot by Treatment

boxplot(
  DGE_AB$samples$lib.size/1e6 ~ DGE_AB$samples$treatment,
  ylab = "Library size (millions)",
  main = "Library Size by Treatment (Control vs Heat)",
  col = c("lightblue", "coral"),
  outline = FALSE
)


# Boxplot by Treatment & Sex

boxplot(
  DGE_AB$samples$lib.size/1e6 ~ interaction(DGE_AB$samples$treatment, DGE_AB$samples$sex),
  ylab = "Library size (millions)",
  main = "Library Size by Treatment and Sex (Control vs Heat)",
  las = 3,
  col = rep(c("lightblue", "coral"), 2),
  outline = FALSE
)

# Display
par(mfrow = c(1, 2))
boxplot(
  DGE_AB$samples$lib.size/1e6 ~ DGE_AB$samples$treatment,
  ylab = "Library size (millions)",
  main = "Library Size by Treatment",
  col = c("lightblue", "coral"),
  outline = FALSE
)

boxplot(
  DGE_AB$samples$lib.size/1e6 ~ interaction(DGE_AB$samples$treatment, DGE_AB$samples$sex),
  ylab = "Library size (millions)",
  main = "Library Size by Treatment and Sex",
  xlab = "",
  las = 3,
  col = rep(c("lightblue", "coral"), 2),
  outline = FALSE
)

```

### DGE with limma-voom ###

```{r}

### Males only

## Remove females samples
AHDB1_AB_Male <- ZF_Liver_sample_list[!grepl("female", ZF_Liver_sample_list$sex), ]
head(AHDB1_AB_Male)

## Subset the countdata to only include Control (A) and Heat (B) samples 
samples_to_keep <- rownames(AHDB1_AB_Male)

length(intersect(samples_to_keep, colnames(Raw_Counts_AB)))

all(samples_to_keep %in% colnames(Raw_Counts_AB))

setdiff(samples_to_keep, colnames(Raw_Counts_AB))

setdiff(colnames(Raw_Counts_AB), samples_to_keep)

common_samples <- intersect(samples_to_keep, colnames(Raw_Counts_AB))

Counts_AB_Male <- Raw_Counts_AB[, common_samples]
AHDB1_AB_Male <- AHDB1_AB_Male[common_samples, ]

AHDB1_AB_Male <- AHDB1_AB_Male[colnames(Counts_AB_Male), ]

identical(rownames(AHDB1_AB_Male), colnames(Counts_AB_Male))

dge_AB_Male_voom <- DGEList(
  counts  = Counts_AB_Male,
  samples = AHDB1_AB_Male
)

# Sanity Check
table(dge_AB_Male_voom$samples$treatment)
cat("Male-Only Subset\n")
cat("Total genes (before filtering):", nrow(dge_AB_Male_voom), "\n\n")

dge_AB_Male_voom$samples$treatment <- droplevels(factor(dge_AB_Male_voom$samples$treatment))

design_male_voom <- model.matrix(~ 0 + treatment, data = dge_AB_Male_voom$samples)
colnames(design_male_voom) <- levels(dge_AB_Male_voom$samples$treatment)

keep_male_voom <- filterByExpr(dge_AB_Male_voom, design_male_voom)
dge_AB_Male_voom <- dge_AB_Male_voom[keep_male_voom, , keep.lib.sizes = FALSE]

cat("Genes after filtering:", nrow(dge_AB_Male_voom), "\n\n")

dge_AB_Male_voom <- calcNormFactors(dge_AB_Male_voom, method = "TMM")

v_male <- voom(dge_AB_Male_voom, design_male_voom, plot = TRUE)

fit_male_voom <- lmFit(v_male, design_male_voom)

contrast_male_voom <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_male_voom
)
cat("Contrast:\n")
print(contrast_male_voom)

fit_male_voom <- contrasts.fit(fit_male_voom, contrast_male_voom)
fit_male_voom <- eBayes(fit_male_voom)

DEG_male_voom <- topTable(fit_male_voom, coef = "Heat_vs_Control", n = Inf, sort.by = "P")
DEG_male_voom$Gene_ID <- rownames(DEG_male_voom)

sig_male_voom <- DEG_male_voom[DEG_male_voom$adj.P.Val < 0.1, ]
sig_male_voom_up <- sig_male_voom[sig_male_voom$logFC > 0, ]
sig_male_voom_down <- sig_male_voom[sig_male_voom$logFC < 0, ]

cat("Male-Only DEG Results 0.1 (limma-voom)\n")
cat("Total significant DEGs:", nrow(sig_male_voom), "\n")
cat("Upregulated in Heat:", nrow(sig_male_voom_up), "\n")
cat("Downregulated in Heat:", nrow(sig_male_voom_down), "\n\n")

stringent_sig_male_voom <- DEG_male_voom[DEG_male_voom$adj.P.Val < 0.05, ]
stringent_sig_male_voom_up <- stringent_sig_male_voom[stringent_sig_male_voom$logFC > 0, ]
stringent_sig_male_voom_down <- stringent_sig_male_voom[stringent_sig_male_voom$logFC < 0, ]

cat("Male-Only DEG Results 0.05 (limma-voom)\n")
cat("Total significant DEGs:", nrow(stringent_sig_male_voom), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_male_voom_up), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_male_voom_down), "\n\n")


# Plot mean variance trend
v_male <- voom(dge_AB_Male_voom, design_male_voom, plot = TRUE)


### Females only

## Remove males samples
AHDB1_AB_Female <- ZF_Liver_sample_list[!grepl("^male$", ZF_Liver_sample_list$sex), ]
head(AHDB1_AB_Female)

## Subset the countdata to only include female Control (A) and Heat (B) samples 
samples_to_keep_female <- rownames(AHDB1_AB_Female)

length(intersect(samples_to_keep_female, colnames(Raw_Counts_AB)))

all(samples_to_keep_female %in% colnames(Raw_Counts_AB))

setdiff(samples_to_keep_female, colnames(Raw_Counts_AB))

setdiff(colnames(Raw_Counts_AB), samples_to_keep_female)

common_samples <- intersect(samples_to_keep_female, colnames(Raw_Counts_AB))

Counts_AB_Female <- Raw_Counts_AB[, common_samples]
AHDB1_AB_Female <- AHDB1_AB_Female[common_samples, ]

AHDB1_AB_Female <- AHDB1_AB_Female[colnames(Counts_AB_Female), ]

identical(rownames(AHDB1_AB_Female), colnames(Counts_AB_Female))

dge_AB_Female_voom <- DGEList(
  counts  = Counts_AB_Female,
  samples = AHDB1_AB_Female
)

# Sanity Check
table(dge_AB_Female_voom$samples$treatment)
cat("Female-Only Subset\n")
cat("Total genes (before filtering):", nrow(dge_AB_Female_voom), "\n\n")

dge_AB_Female_voom$samples$treatment <- droplevels(factor(dge_AB_Female_voom$samples$treatment))

design_female_voom <- model.matrix(~ 0 + treatment, data = dge_AB_Female_voom$samples)
colnames(design_female_voom) <- levels(dge_AB_Female_voom$samples$treatment)

keep_female_voom <- filterByExpr(dge_AB_Female_voom, design_female_voom)
dge_AB_Female_voom <- dge_AB_Female_voom[keep_female_voom, , keep.lib.sizes = FALSE]

cat("Genes after filtering:", nrow(dge_AB_Female_voom), "\n\n")

dge_AB_Female_voom <- calcNormFactors(dge_AB_Female_voom, method = "TMM")

v_female <- voom(dge_AB_Female_voom, design_female_voom, plot = TRUE)

fit_female_voom <- lmFit(v_female, design_female_voom)

contrast_female_voom <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_female_voom
)
cat("Contrast:\n")
print(contrast_female_voom)

fit_female_voom <- contrasts.fit(fit_female_voom, contrast_female_voom)
fit_female_voom <- eBayes(fit_female_voom)

DEG_female_voom <- topTable(fit_female_voom, coef = "Heat_vs_Control", n = Inf, sort.by = "P")
DEG_female_voom$Gene_ID <- rownames(DEG_female_voom)

sig_female_voom <- DEG_female_voom[DEG_female_voom$adj.P.Val < 0.1, ]
sig_female_voom_up <- sig_female_voom[sig_female_voom$logFC > 0, ]
sig_female_voom_down <- sig_female_voom[sig_female_voom$logFC < 0, ]

cat("Female-Only DEG Results 0.1 (limma-voom)\n")
cat("Total significant DEGs:", nrow(sig_female_voom), "\n")
cat("Upregulated in Heat:", nrow(sig_female_voom_up), "\n")
cat("Downregulated in Heat:", nrow(sig_female_voom_down), "\n\n")

stringent_sig_female_voom <- DEG_female_voom[DEG_female_voom$adj.P.Val < 0.05, ]
stringent_sig_female_voom_up <- stringent_sig_female_voom[stringent_sig_female_voom$logFC > 0, ]
stringent_sig_female_voom_down <- stringent_sig_female_voom[stringent_sig_female_voom$logFC < 0, ]

cat("Female-Only DEG Results 0.05 (limma-voom)\n")
cat("Total significant DEGs:", nrow(stringent_sig_female_voom), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_female_voom_up), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_female_voom_down), "\n\n")

# Plot mean variance trend
v_female <- voom(dge_AB_Female_voom, design_female_voom, plot = TRUE)

```

### DGE with DESeq2 ###

```{r}
## Interaction - treatment:Sex

# Build DESeq2 object
dds_interaction <- DESeqDataSetFromMatrix(
  countData = Raw_Counts_AB,
  colData   = ZF_Liver_sample_list,
  design    = ~ 0 + treatment + sex + treatment:sex
)

# Perform a pre-filtering to remove rows (genes) that have less, on average across individuals, 100 reads mapped per individual. 40 individuals x 100 reads = 4000 reads. A gene must have at least 4900 reads total to be included for analysis.

keep_int_deseq <- dds_interaction[ rowSums(counts(dds_interaction)) > 4000, ]
cat("Genes after filtering:", nrow(keep_int_deseq), "\n\n")


dds_interaction <- DESeq(dds_interaction)

res_interaction <- results(
    dds_interaction,
    name = "treatmentHeat.sexmale"
)

summary(res_interaction)

DEG_int_deseq <- as.data.frame(res_interaction)
DEG_int_deseq$Gene_ID <- rownames(DEG_int_deseq)

# Remove NA's
DEG_int_deseq <- DEG_int_deseq[
  !is.na(DEG_int_deseq$padj) & !is.na(DEG_int_deseq$log2FoldChange),
]
cat("Genes after NA removal:", nrow(DEG_int_deseq), "\n\n")

# For GSEA and RRHO2 input
write.csv(DEG_int_deseq, "DEG_int_deseq.csv", row.names = FALSE)

# Subset DEG"s into upregulated and downregulated 
sig_int_deseq      <- DEG_int_deseq[DEG_int_deseq$padj < 0.1, ]
sig_int_deseq_up   <- sig_int_deseq[sig_int_deseq$log2FoldChange > 0, ]
sig_int_deseq_down <- sig_int_deseq[sig_int_deseq$log2FoldChange < 0, ]

write.csv(sig_int_deseq, "sig_int_liver_deseq.csv", row.names = FALSE)
write.csv(sig_int_deseq_up, "sig_int_deseq_liver_up.csv", row.names = FALSE)
write.csv(sig_int_deseq_down, "sig_int_deseq_liver_down.csv", row.names = FALSE)

cat("Interaction DEG Results 0.1 (DESeq2)\n")
cat("Total significant DEGs:", nrow(sig_int_deseq), "\n")
cat("Upregulated in Heat:", nrow(sig_int_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(sig_int_deseq_down), "\n\n")

stringent_sig_int_deseq      <- DEG_int_deseq[DEG_int_deseq$padj < 0.05, ]
stringent_sig_int_deseq_up   <- stringent_sig_int_deseq[stringent_sig_int_deseq$log2FoldChange > 0, ]
stringent_sig_int_deseq_down <- stringent_sig_int_deseq[stringent_sig_int_deseq$log2FoldChange < 0, ]

cat("Interaction DEG Results 0.05 (DESeq2)\n")
cat("Total significant DEGs:", nrow(stringent_sig_int_deseq), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_int_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_int_deseq_down), "\n\n")

## Males only

## Remove female samples
AHDB1_AB_male <- ZF_Liver_sample_list[!grepl("^female$", ZF_Liver_sample_list$sex), ]
head(AHDB1_AB_male)

## Subset the countdata to only include female Control (A) and Heat (B) samples 
samples_to_keep_male <- rownames(AHDB1_AB_male)

length(intersect(samples_to_keep_male, colnames(keep_int_deseq)))

all(samples_to_keep_male %in% colnames(keep_int_deseq))

setdiff(samples_to_keep_male, colnames(keep_int_deseq))

setdiff(colnames(keep_int_deseq), samples_to_keep_male)

common_samples <- intersect(samples_to_keep_male, colnames(keep_int_deseq))

Counts_AB_male <- keep_int_deseq[, common_samples]
AHDB1_AB_male <- AHDB1_AB_male[common_samples, ]

AHDB1_AB_male <- AHDB1_AB_male[colnames(Counts_AB_male), ]

identical(rownames(AHDB1_AB_male), colnames(Counts_AB_male))

# Extract count matrix
counts_male <- counts(Counts_AB_male)
coldata_male <- as.data.frame(colData(Counts_AB_male))

# Build DESeq2 object
dds_male <- DESeqDataSetFromMatrix(
  countData = counts_male,
  colData   = coldata_male,
  design    = ~ 0 + treatment
)

keep_male_deseq <- rowSums(counts(dds_male)) >= 10
dds_male <- dds_male[keep_male_deseq, ]
cat("Genes after filtering:", nrow(dds_male), "\n\n")

dds_male <- DESeq(dds_male)

res_male <- results(
  dds_male,
  contrast = c("treatment", "Heat", "Control"),
  alpha    = 0.05
)

summary(res_male)

DEG_male_deseq <- as.data.frame(res_male)
DEG_male_deseq$Gene_ID <- rownames(DEG_male_deseq)

# Remove NA's
DEG_male_deseq <- DEG_male_deseq[
  !is.na(DEG_male_deseq$padj) & !is.na(DEG_male_deseq$log2FoldChange),
]
cat("Genes after NA removal:", nrow(DEG_male_deseq), "\n\n")

# Subset DEG"s into upregulated and downregulated 
sig_male_deseq      <- DEG_male_deseq[DEG_male_deseq$padj < 0.1, ]
sig_male_deseq_up   <- sig_male_deseq[sig_male_deseq$log2FoldChange > 0, ]
sig_male_deseq_down <- sig_male_deseq[sig_male_deseq$log2FoldChange < 0, ]

cat("Male-Only DEG Results 0.1 (DESeq2)\n")
cat("Total significant DEGs:", nrow(sig_male_deseq), "\n")
cat("Upregulated in Heat:", nrow(sig_male_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(sig_male_deseq_down), "\n\n")

stringent_sig_male_deseq      <- DEG_male_deseq[DEG_male_deseq$padj < 0.05, ]
stringent_sig_male_deseq_up   <- stringent_sig_male_deseq[stringent_sig_male_deseq$log2FoldChange > 0, ]
stringent_sig_male_deseq_down <- stringent_sig_male_deseq[stringent_sig_male_deseq$log2FoldChange < 0, ]

cat("Male-Only DEG Results 0.05 (DESeq2)\n")
cat("Total significant DEGs:", nrow(stringent_sig_male_deseq), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_male_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_male_deseq_down), "\n\n")

## Export DEGs to CSV — Males
write.csv(DEG_male_deseq, "DEG_male_deseq_allgenes.csv", row.names = FALSE)
write.csv(sig_male_deseq, "sig_male_deseq_0.1.csv", row.names = FALSE)
write.csv(sig_male_deseq_up, "sig_male_deseq_up_0.1.csv", row.names = FALSE)
write.csv(sig_male_deseq_down, "sig_male_deseq_down_0.1.csv", row.names = FALSE)

## Females only

## Remove males samples
AHDB1_AB_Female <- ZF_Liver_sample_list[!grepl("^male$", ZF_Liver_sample_list$sex), ]
head(AHDB1_AB_Female)

## Subset the countdata to only include female Control (A) and Heat (B) samples 
samples_to_keep_female <- rownames(AHDB1_AB_Female)

length(intersect(samples_to_keep_female, colnames(keep_int_deseq)))

all(samples_to_keep_female %in% colnames(keep_int_deseq))

setdiff(samples_to_keep_female, colnames(keep_int_deseq))

setdiff(colnames(keep_int_deseq), samples_to_keep_female)

common_samples <- intersect(samples_to_keep_female, colnames(keep_int_deseq))

Counts_AB_Female <- keep_int_deseq[, common_samples]
AHDB1_AB_Female <- AHDB1_AB_Female[common_samples, ]

AHDB1_AB_Female <- AHDB1_AB_Female[colnames(Counts_AB_Female), ]

identical(rownames(AHDB1_AB_Female), colnames(Counts_AB_Female))

# Extract count matrix
counts_female <- counts(Counts_AB_Female)
coldata_female <- as.data.frame(colData(Counts_AB_Female))

# Build DESeq2 object
dds_female <- DESeqDataSetFromMatrix(
  countData = counts_female,
  colData   = coldata_female,
  design    = ~ 0 + treatment
)

keep_female_deseq <- rowSums(counts(dds_female)) >= 10
dds_female <- dds_female[keep_female_deseq, ]
cat("Genes after filtering:", nrow(dds_female), "\n\n")

dds_female <- DESeq(dds_female)

res_female <- results(
  dds_female,
  contrast = c("treatment", "Heat", "Control"),
  alpha    = 0.05
)

summary(res_female)

DEG_female_deseq <- as.data.frame(res_female)
DEG_female_deseq$Gene_ID <- rownames(DEG_female_deseq)

# Remove NA's
DEG_female_deseq <- DEG_female_deseq[
  !is.na(DEG_female_deseq$padj) & !is.na(DEG_female_deseq$log2FoldChange),
]
cat("Genes after NA removal:", nrow(DEG_female_deseq), "\n\n")

# Subset DEG's into upregulated and downregulated
sig_female_deseq      <- DEG_female_deseq[DEG_female_deseq$padj < 0.1, ]
sig_female_deseq_up   <- sig_female_deseq[sig_female_deseq$log2FoldChange > 0, ]
sig_female_deseq_down <- sig_female_deseq[sig_female_deseq$log2FoldChange < 0, ]

cat("Female-Only DEG Results 0.1 (DESeq2)\n")
cat("Total significant DEGs:", nrow(sig_female_deseq), "\n")
cat("Upregulated in Heat:", nrow(sig_female_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(sig_female_deseq_down), "\n\n")

stringent_sig_female_deseq      <- DEG_female_deseq[DEG_female_deseq$padj < 0.05, ]
stringent_sig_female_deseq_up   <- stringent_sig_female_deseq[stringent_sig_female_deseq$log2FoldChange > 0, ]
stringent_sig_female_deseq_down <- stringent_sig_female_deseq[stringent_sig_female_deseq$log2FoldChange < 0, ]

cat("Female-Only DEG Results 0.05 (DESeq2)\n")
cat("Total significant DEGs:", nrow(stringent_sig_female_deseq), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_female_deseq_up), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_female_deseq_down), "\n\n")

## Export DEGs to CSV — Females
write.csv(DEG_female_deseq, "DEG_female_deseq_allgenes.csv", row.names = FALSE)
write.csv(sig_female_deseq, "sig_female_deseq_0.1.csv", row.names = FALSE)
write.csv(sig_female_deseq_up, "sig_female_deseq_up_0.1.csv", row.names = FALSE)
write.csv(sig_female_deseq_down, "sig_female_deseq_down_0.1.csv", row.names = FALSE)

```

### DGE with Edge R ###


```{r}

# Males only
dim(AHDB1_AB_Male)

dge_AB_Male <- DGEList(
  counts  = Counts_AB_Male,
  samples = AHDB1_AB_Male
)

# Set factor and build design matrix
dge_AB_Male$samples$treatment <- droplevels(factor(dge_AB_Male$samples$treatment))

design_male <- model.matrix(~ 0 + treatment, data = dge_AB_Male$samples)
colnames(design_male) <- levels(dge_AB_Male$samples$treatment)
table(dge_AB_Male$samples$treatment)

cat("Total genes (before filtering):", nrow(dge_AB_Male), "\n\n")

# filter using the design matrix
keep_male <- filterByExpr(dge_AB_Male, design_male)
dge_AB_Male <- dge_AB_Male[keep_male, , keep.lib.sizes = FALSE]
cat("Total genes (after filtering):", nrow(dge_AB_Male), "\n\n")

# TMM normalization
dge_AB_Male <- calcNormFactors(dge_AB_Male, method = "TMM")

dge_AB_Male <- estimateDisp(dge_AB_Male, design_male)
fit_male <- glmQLFit(dge_AB_Male, design_male)

contrast_male <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_male
)
cat("Contrast")
print(contrast_male)

qlf_male <- glmQLFTest(fit_male, contrast = contrast_male)

DEG_male <- topTags(qlf_male, n = Inf)$table
DEG_male$Gene_ID <- rownames(DEG_male)

sig_male_EdgeR <- DEG_male[DEG_male$FDR < 0.1, ]
sig_male_up_EdgeR <- sig_male_EdgeR[sig_male_EdgeR$logFC > 0, ]
sig_male_down_EdgeR <- sig_male_EdgeR[sig_male_EdgeR$logFC < 0, ]

cat("Male-Only DEG Results 0.1 - EdgeR")
cat("Total significant DEGs:", nrow(sig_male_EdgeR), "\n")
cat("Upregulated in Heat:", nrow(sig_male_up_EdgeR), "\n")
cat("Downregulated in Heat:", nrow(sig_male_down_EdgeR), "\n\n")


stringent_sig_male_EdgeR <- DEG_male[DEG_male$FDR < 0.05, ]
stringent_sig_male_up_EdgeR <- stringent_sig_male_EdgeR[stringent_sig_male_EdgeR$logFC > 0, ]
stringent_sig_male_down_EdgeR <- stringent_sig_male_EdgeR[stringent_sig_male_EdgeR$logFC < 0, ]

cat("Male-Only DEG Results 0.05 - EdgeR")
cat("Total significant DEGs:", nrow(stringent_sig_male_EdgeR), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_male_up_EdgeR), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_male_down_EdgeR), "\n\n")

#Females only
dim(AHDB1_AB_Female)

dge_AB_Female <- DGEList(
  counts  = Counts_AB_Female,
  samples = AHDB1_AB_Female
)

# Set factor and build design matrix
dge_AB_Female$samples$treatment <- droplevels(factor(dge_AB_Female$samples$treatment))

design_female <- model.matrix(~ 0 + treatment, data = dge_AB_Female$samples)
colnames(design_female) <- levels(dge_AB_Female$samples$treatment)
table(dge_AB_Female$samples$treatment)

cat("Total genes (before filtering):", nrow(dge_AB_Female), "\n\n")

# filter using the design matrix
keep_female <- filterByExpr(dge_AB_Female, design_female)
dge_AB_Female <- dge_AB_Female[keep_female, , keep.lib.sizes = FALSE]
cat("Total genes (after filtering):", nrow(dge_AB_Female), "\n\n")

# TMM normalization
dge_AB_Female <- calcNormFactors(dge_AB_Female, method = "TMM")

dge_AB_Female <- estimateDisp(dge_AB_Female, design_female)
fit_female <- glmQLFit(dge_AB_Female, design_female)

contrast_female <- makeContrasts(
  Heat_vs_Control = Heat - Control,
  levels = design_female
)
cat("Contrast")
print(contrast_female)

qlf_female <- glmQLFTest(fit_female, contrast = contrast_female)

DEG_female <- topTags(qlf_female, n = Inf)$table
DEG_female$Gene_ID <- rownames(DEG_female)

sig_female_EdgeR      <- DEG_female[DEG_female$FDR < 0.1, ]
sig_female_up_EdgeR   <- sig_female_EdgeR[sig_female_EdgeR$logFC > 0, ]
sig_female_down_EdgeR <- sig_female_EdgeR[sig_female_EdgeR$logFC < 0, ]

cat("Female-Only DEG Results 0.1 - EdgeR")
cat("Total significant DEGs:", nrow(sig_female_EdgeR), "\n")
cat("Upregulated in Heat:", nrow(sig_female_up_EdgeR), "\n")
cat("Downregulated in Heat:", nrow(sig_female_down_EdgeR), "\n\n")

stringent_sig_female_EdgeR      <- DEG_female[DEG_female$FDR < 0.05, ]
stringent_sig_female_up_EdgeR   <- stringent_sig_female_EdgeR[stringent_sig_female_EdgeR$logFC > 0, ]
stringent_sig_female_down_EdgeR <- stringent_sig_female_EdgeR[stringent_sig_female_EdgeR$logFC < 0, ]

cat("Female-Only DEG Results 0.05 - EdgeR")
cat("Total significant DEGs:", nrow(stringent_sig_female_EdgeR), "\n")
cat("Upregulated in Heat:", nrow(stringent_sig_female_up_EdgeR), "\n")
cat("Downregulated in Heat:", nrow(stringent_sig_female_down_EdgeR), "\n\n")

```

# Upset Plot for all three methods - Males only

```{r}
# Install & Load packages
library(UpSetR)

# Make characters - Run once & then not again
sig_male_voom_up <- as.character(sig_male_voom_up$Gene_ID)
sig_male_deseq_up <- as.character(sig_male_deseq_up$Gene_ID)
sig_male_up_EdgeR <- as.character(sig_male_up_EdgeR$Gene_ID)

sig_male_voom_down <- as.character(sig_male_voom_down$Gene_ID)
sig_male_deseq_down <- as.character(sig_male_deseq_down$Gene_ID)
sig_male_down_EdgeR <- as.character(sig_male_down_EdgeR$Gene_ID)

# Build lists
up_list <- list(
  LimmaVoom = sig_male_voom_up,
  DESeq2    = sig_male_deseq_up,
  edgeR     = sig_male_up_EdgeR
)

down_list <- list(
  LimmaVoom = sig_male_voom_down,
  DESeq2    = sig_male_deseq_down,
  edgeR     = sig_male_down_EdgeR
)

# Plot upregulated
png("Male_DGEProgramComp_up_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(up_list),
  mainbar.y.label = "Male upregulated intersections",
  sets.x.label = "Gene counts"
)
dev.off()

# Plot downregulated
png("Male_DGEProgramComp_down_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(down_list),
  mainbar.y.label = "Male downregulated intersections",
  sets.x.label = "Gene counts"
)

dev.off()
```

# Upset Plot for all three methods - Females only

```{r}
# Install & Load packages
library(UpSetR)

# Make characters - Run once & then not again
sig_female_voom_up <- as.character(sig_female_voom_up$Gene_ID)
sig_female_deseq_up <- as.character(sig_female_deseq_up$Gene_ID)
sig_female_up_EdgeR <- as.character(sig_female_up_EdgeR$Gene_ID)

sig_female_voom_down <- as.character(sig_female_voom_down$Gene_ID)
sig_female_deseq_down <- as.character(sig_female_deseq_down$Gene_ID)
sig_female_down_EdgeR <- as.character(sig_female_down_EdgeR$Gene_ID)

# Build lists
female_up_list <- list(
  LimmaVoom = sig_female_voom_up,
  DESeq2    = sig_female_deseq_up,
  edgeR     = sig_female_up_EdgeR
)

female_down_list <- list(
  LimmaVoom = sig_female_voom_down,
  DESeq2    = sig_female_deseq_down,
  edgeR     = sig_female_down_EdgeR
)

female_up_list <- lapply(female_up_list, function(x) unique(na.omit(x)))


# Plot upregulated
png("Female_DGEProgramComp_up_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(female_up_list),
  mainbar.y.label = "Female upregulated intersections",
  sets.x.label = "Gene counts"
)

dev.off()

# Plot downregulated
png("Female_DGEProgramComp_down_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(female_down_list),
  mainbar.y.label = "Female downregulated intersections",
  sets.x.label = "Gene counts"
)

dev.off()
```

# Upset Plot for Limma Voom - Sex intersections in heat

```{r}
# Install & Load packages
library(UpSetR)

# Make characters - Run once & then not again
#sig_male_voom_up <- as.character(sig_male_voom_up$Gene_ID)
#sig_female_voom_up <- as.character(sig_female_voom_up$Gene_ID)

#sig_male_voom_down <- as.character(sig_male_voom_down$Gene_ID)
#sig_female_voom_down <- as.character(sig_female_voom_down$Gene_ID)

# Build lists
up_list <- list(
  Male_upregulated_in_heat    = sig_male_voom_up,
  Female_upregulated_in_heat  = sig_female_voom_up
)

down_list <- list(
  Male_downregulated_in_heat    = sig_male_voom_down,
  Female_downregulated_in_heat    = sig_female_voom_down
)

# Plot upregulated
png("LimmaVoom_Sex_up_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(up_list),
  mainbar.y.label = "Upregulated intersections",
  sets.x.label = "Gene counts"
)

dev.off()

# Plot downregulated
png("LimmaVoom_Sex_down_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(down_list),
  mainbar.y.label = "Downregulated intersections",
  sets.x.label = "Gene counts"
)

dev.off()
```

# Upset Plot for DESeq2 - Sex intersections in heat

```{r}
# Install & Load packages
library(UpSetR)

# Make characters - Run once & then not again
sig_male_deseq_up <- as.character(sig_male_deseq_up$Gene_ID)
sig_female_deseq_up <- as.character(sig_female_deseq_up$Gene_ID)

sig_male_deseq_down <- as.character(sig_male_deseq_down$Gene_ID)
sig_female_deseq_down <- as.character(sig_female_deseq_down$Gene_ID)

# Build lists
up_list <- list(
  Male_upregulated_in_heat    = sig_male_deseq_up,
  Female_upregulated_in_heat  = sig_female_deseq_up
)

down_list <- list(
  Male_downregulated_in_heat    = sig_male_deseq_down,
  Female_downregulated_in_heat    = sig_female_deseq_down
)

# Plot upregulated
png("DESeq2_Sex_up_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(up_list),
  mainbar.y.label = "Upregulated intersections",
  sets.x.label = "Gene counts"
)

dev.off()

# Plot downregulated
png("DESeq2_Sex_down_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(down_list),
  mainbar.y.label = "Downregulated intersections",
  sets.x.label = "Gene counts"
)

dev.off()
```

# Upset Plot for EdgeR - Sex intersections in heat

```{r}
# Install & Load packages
library(UpSetR)

# Make characters - Run once & then not again
#sig_male_up_EdgeR <- as.character(sig_male_up_EdgeR$Gene_ID)
#sig_female_up_EdgeR <- as.character(sig_female_up_EdgeR$Gene_ID)

#sig_male_down_EdgeR <- as.character(sig_male_down_EdgeR$Gene_ID)
#sig_female_down_EdgeR <- as.character(sig_female_down_EdgeR$Gene_ID)

# Build lists
up_list <- list(
  Male_upregulated_in_heat    = sig_male_up_EdgeR,
  Female_upregulated_in_heat  = sig_female_up_EdgeR
)

down_list <- list(
  Male_downregulated_in_heat    = sig_male_down_EdgeR,
  Female_downregulated_in_heat    = sig_female_down_EdgeR
)

# Plot upregulated
png("EdgeR_Sex_up_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(up_list),
  mainbar.y.label = "Upregulated intersections",
  sets.x.label = "Gene counts"
)

dev.off()

# Plot downregulated
png("EdgeR_Sex_down_UpSet_Liver.png", width = 2000, height = 1000, res = 150)

par(mfrow = c(1,2))

upset(
  fromList(down_list),
  mainbar.y.label = "Downregulated intersections",
  sets.x.label = "Gene counts"
)

dev.off()
```

# Make a heatmap with DESeq2 adjusted p-values < 0.05
# Ordered Sex -> Treatment
```{r}
library(pheatmap)

vsd_all <- vst(dds_interaction, blind = FALSE)

vsd_mat <- assay(vsd_all)

# Combine gene names
sig_genes <- unique(c(
  rownames(stringent_sig_male_deseq),
  rownames(stringent_sig_female_deseq)
))

# Keep only genes present in your VST matrix
sig_genes <- sig_genes[sig_genes %in% rownames(vsd_mat)]

# Check how many genes are being plotted
length(sig_genes)

# Expression matrix containing BOTH males and females
mat_sig <- vsd_mat[sig_genes, ]


### Make the Heatmap
# Scale each gene (row) to z-scores
mat_scaled_sig <- t(scale(t(mat_sig)))

# Optional sample annotation
annotation_col <- data.frame(
  Treatment = colData(vsd_all)$treatment,
  Sex = colData(vsd_all)$sex
)

rownames(annotation_col) <- colnames(vsd_all)

annotation_col <- annotation_col[colnames(mat_sig), , drop = FALSE]

sample_order <- order(
  annotation_col$Sex,
  annotation_col$Treatment
)

# Reorder expression matrix
mat_scaled_sig <- mat_scaled_sig[, sample_order]

# Reorder annotation
annotation_col <- annotation_col[sample_order, , drop = FALSE]

annotation_colors <- list(
  Sex = c(
    male = "darkgreen",
    female = "deeppink"
  ),
    Treatment = c(
    Control = "gray70",
    Heat = "orange"
  )
)


# Heatmap
pheatmap(
  mat_scaled_sig,
  
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  
  cluster_rows = TRUE,
  
  # Do NOT cluster columns; preserve biological grouping
  cluster_cols = FALSE,
  
  show_rownames = FALSE,
  show_colnames = FALSE,
  
  fontsize_row = 8,
  
  border_color = NA,
  
  color = colorRampPalette(
    c("#2166AC", "white", "#B2182B")
  )(100),
  
  breaks = seq(-2.5, 2.5, length.out = 101)
)
```

# Make a heatmap with DESeq2 adjusted p-values < 0.05
# Ordered Treatment -> Sex
```{r}
library(pheatmap)

vsd_all <- vst(dds_interaction, blind = FALSE)

vsd_mat <- assay(vsd_all)

# Combine gene names
sig_genes <- unique(c(
  rownames(stringent_sig_male_deseq),
  rownames(stringent_sig_female_deseq)
))

# Keep only genes present in your VST matrix
sig_genes <- sig_genes[sig_genes %in% rownames(vsd_mat)]

# Check how many genes are being plotted
length(sig_genes)

# Expression matrix containing BOTH males and females
mat_sig <- vsd_mat[sig_genes, ]


### Make the Heatmap
# Scale each gene (row) to z-scores
mat_scaled_sig <- t(scale(t(mat_sig)))

# Optional sample annotation
annotation_col <- data.frame(
  Treatment = colData(vsd_all)$treatment,
  Sex = colData(vsd_all)$sex
)

rownames(annotation_col) <- colnames(vsd_all)

annotation_col <- annotation_col[colnames(mat_sig), , drop = FALSE]

sample_order <- order(
  annotation_col$Treatment,
  annotation_col$Sex
)

# Reorder expression matrix
mat_scaled_sig <- mat_scaled_sig[, sample_order]

# Reorder annotation
annotation_col <- annotation_col[sample_order, , drop = FALSE]

annotation_colors <- list(
    Treatment = c(
    Control = "gray70",
    Heat = "orange"
  ),
    Sex = c(
    male = "darkgreen",
    female = "deeppink"
  )
)


# Heatmap
png("/Users/brynleighpayne/Downloads/ZebraFinch_DamageFitnessProject/LiverRNASeq/DGE/Liver_DGE_Heatmap.png", width = 6, height = 4, units = "in", res = 300)

pheatmap(
  mat_scaled_sig,
  
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  
  cluster_rows = TRUE,
  
  # Do NOT cluster columns; preserve biological grouping
  cluster_cols = FALSE,
  
  show_rownames = FALSE,
  show_colnames = FALSE,
  
  fontsize_row = 8,
  
  border_color = NA,
  
  color = colorRampPalette(
    c("#2166AC", "white", "#B2182B")
  )(100),
  
  breaks = seq(-2.5, 2.5, length.out = 101)
)

dev.off()
```

# Make a heatmap with DESeq2 adjusted p-values < 0.05 - Interaction Model
# Ordered Treatment -> Sex
```{r}
library(pheatmap)

vsd_all <- vst(dds_interaction, blind = FALSE)

vsd_mat <- assay(vsd_all)

# Combine gene names
sig_genes <- rownames(stringent_sig_int_deseq)

# Keep only genes present in your VST matrix
sig_genes <- sig_genes[sig_genes %in% rownames(vsd_mat)]

# Check how many genes are being plotted
length(sig_genes)

# Expression matrix containing BOTH males and females
mat_sig <- vsd_mat[sig_genes, ]


### Make the Heatmap
# Scale each gene (row) to z-scores
mat_scaled_sig <- t(scale(t(mat_sig)))

# Optional sample annotation
annotation_col <- data.frame(
  Treatment = colData(vsd_all)$treatment,
  Sex = colData(vsd_all)$sex
)

rownames(annotation_col) <- colnames(vsd_all)

annotation_col <- annotation_col[colnames(mat_sig), , drop = FALSE]

sample_order <- order(
  annotation_col$Treatment,
  annotation_col$Sex
)

# Reorder expression matrix
mat_scaled_sig <- mat_scaled_sig[, sample_order]

# Reorder annotation
annotation_col <- annotation_col[sample_order, , drop = FALSE]

annotation_colors <- list(
    Treatment = c(
    Control = "gray70",
    Heat = "orange"
  ),
    Sex = c(
    male = "darkgreen",
    female = "deeppink"
  )
)


# Heatmap
pheatmap(
  mat_scaled_sig,
  
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  
  cluster_rows = TRUE,
  
  # Do NOT cluster columns; preserve biological grouping
  cluster_cols = FALSE,
  
  show_rownames = FALSE,
  show_colnames = FALSE,
  
  fontsize_row = 8,
  
  border_color = NA,
  
  color = colorRampPalette(
    c("#2166AC", "white", "#B2182B")
  )(100),
  
  breaks = seq(-2.5, 2.5, length.out = 101)
)
```
